---
title: "Urban Land-Use Efficiency:<br>A Stable Metric for SDG 11.3.1, and Why Per-Capita Designs Cannot Cleanly Identify Its Drivers"
subtitle: "SDG 11.3.1: The Ratio of Land Consumption Rate to Population Growth Rate<br>193 UN Member Countries x 1985-2020 Panel Data Analysis"
author:
  - name: "Ömer Furkan Çoban"
    affiliation: "Development Economics · SoSe 2026"
  - name: "Prof. Dr. Jürgen Bitzer"
date: "2026-06-13"
date-format: "DD.MM.YYYY"
format:
  revealjs:
    theme: default
    css: custom.css
    embed-resources: true
    html-math-method: katex
    transition: fade
    slide-number: true
    chalkboard: false
    width: 1920
    height: 1080
    footer: "13.06.2026 | Development Economics | Ömer Furkan Çoban<br><span style='font-size:0.85em;opacity:0.6;'>SDG 11.3.1: The Ratio of Land Consumption Rate to Population Growth Rate</span>"
    logo: "uol_logo.svg"
    margin: 0.06
    center: false
    scrollable: true
    highlight-style: github
bibliography: references.bib
link-citations: true
citations-hover: true
resources:
  - population_counter.html
title-slide-attributes:
  data-background-video: "sentinel2-landcover-animation-b5247e.mp4"
  data-background-video-loop: "true"
  data-background-video-muted: "true"
  data-background-color: "#18181b"
  data-background-opacity: "0.55"
  data-visibility: "uncounted"
filters:
  - ofurkancoban/progress-indicator
  - laser-pointer
---

```{r}
#| label: setup
#| include: false
library(tidyverse)
library(fixest)
library(modelsummary)
library(gt)
library(plotly)

panel <- read_csv(
    here::here("03_datasets/processed/reg_panel_urban.csv"),
    show_col_types = FALSE
) |>
    filter(year <= 2020, !is.na(bpcr), !is.na(lcrpgr_log)) |> # observed epochs (drop 2025/2030 projections)
    mutate(
        period_f = factor(year)
    ) |>
    arrange(iso3, year) |>
    group_by(iso3) |>
    mutate(
        bpcr_L = lag(bpcr),
        lcrpgr_L = lag(lcrpgr_log),
        # Density and urban share carry the same urban population count as the
        # per-capita dependent variable, so a common population shock drives
        # either contemporaneous coefficient negative by construction. Both enter
        # lagged (predetermined), exactly as in the paper.
        ln_urban_density_L = lag(ln_urban_density),
        urban_pop_share_L = lag(urban_pop_share),
        urb_rate = (urban_pop_share - lag(urban_pop_share)) / 5 # predetermined urbanisation rate (Δurban share / yr)
    ) |>
    ungroup() |>
    filter(!is.na(bpcr_L), !is.na(lcrpgr_L))

# ── Fixed sample: all covariates present (enables valid cross-model comparison)
panel_fixed <- panel |>
    filter(
        !is.na(ln_gdp_pc),
        !is.na(urban_pop_share),
        !is.na(ln_urban_density),
        !is.na(net_migr_pct),
        !is.na(ln_urban_density_L),
        !is.na(urban_pop_share_L)
    ) |>
    mutate(
        pop_growth = pgr_log, # urban population log growth (the PGR term inside BpCR)
        shrinking = as.integer(!is.na(pgr_log) & pgr_log < 0)
    )

# ── LCRPGR stepwise: variables added by significance order (BpCR benchmark) ──
l1 <- feols(
    lcrpgr_log ~ lcrpgr_L | iso3 + period_f,
    data = panel_fixed,
    vcov = ~iso3
)
l2 <- feols(
    lcrpgr_log ~ lcrpgr_L + ln_urban_density_L | iso3 + period_f,
    data = panel_fixed,
    vcov = ~iso3
)
l3 <- feols(
    lcrpgr_log ~ lcrpgr_L + ln_urban_density_L + net_migr_pct | iso3 + period_f,
    data = panel_fixed,
    vcov = ~iso3
)
l4 <- feols(
    lcrpgr_log ~ lcrpgr_L +
        ln_urban_density_L +
        net_migr_pct +
        ln_gdp_pc |
        iso3 + period_f,
    data = panel_fixed,
    vcov = ~iso3
)
l5 <- feols(
    lcrpgr_log ~ lcrpgr_L +
        ln_urban_density_L +
        net_migr_pct +
        ln_gdp_pc +
        urban_pop_share_L |
        iso3 + period_f,
    data = panel_fixed,
    vcov = ~iso3
)
l5_int <- feols(
    lcrpgr_log ~ lcrpgr_L +
        ln_urban_density_L +
        net_migr_pct +
        ln_gdp_pc +
        urban_pop_share_L |
        iso3 + period_f,
    data = panel_fixed,
    vcov = ~iso3
)

# ── BpCR stepwise: variables added by significance order ─────────────────────
b1 <- feols(bpcr ~ bpcr_L | iso3 + period_f, data = panel_fixed, vcov = ~iso3)
b2 <- feols(
    bpcr ~ bpcr_L + ln_urban_density_L | iso3 + period_f,
    data = panel_fixed,
    vcov = ~iso3
)
b3 <- feols(
    bpcr ~ bpcr_L + ln_urban_density_L + net_migr_pct | iso3 + period_f,
    data = panel_fixed,
    vcov = ~iso3
)
b4 <- feols(
    bpcr ~ bpcr_L +
        ln_urban_density_L +
        net_migr_pct +
        ln_gdp_pc |
        iso3 + period_f,
    data = panel_fixed,
    vcov = ~iso3
)
b5 <- feols(
    bpcr ~ bpcr_L +
        ln_urban_density_L +
        net_migr_pct +
        ln_gdp_pc +
        urban_pop_share_L |
        iso3 + period_f,
    data = panel_fixed,
    vcov = ~iso3
)
b5_int <- feols(
    bpcr ~ bpcr_L +
        ln_urban_density_L +
        net_migr_pct +
        ln_gdp_pc +
        urban_pop_share_L |
        iso3 + period_f,
    data = panel_fixed,
    vcov = ~iso3
)

# ── Static (no lag) stepwise: metric comparison (Models 1 & 2, lag-free) ──────
ls1 <- feols(
    lcrpgr_log ~ ln_urban_density_L | iso3 + period_f,
    data = panel_fixed,
    vcov = ~iso3
)
ls2 <- feols(
    lcrpgr_log ~ ln_urban_density_L + net_migr_pct | iso3 + period_f,
    data = panel_fixed,
    vcov = ~iso3
)
ls3 <- feols(
    lcrpgr_log ~ ln_urban_density_L + net_migr_pct + ln_gdp_pc | iso3 + period_f,
    data = panel_fixed,
    vcov = ~iso3
)
ls4 <- feols(
    lcrpgr_log ~ ln_urban_density_L +
        net_migr_pct +
        ln_gdp_pc +
        urban_pop_share_L |
        iso3 + period_f,
    data = panel_fixed,
    vcov = ~iso3
)
bs1 <- feols(
    bpcr ~ ln_urban_density_L | iso3 + period_f,
    data = panel_fixed,
    vcov = ~iso3
)
bs2 <- feols(
    bpcr ~ ln_urban_density_L + net_migr_pct | iso3 + period_f,
    data = panel_fixed,
    vcov = ~iso3
)
bs3 <- feols(
    bpcr ~ ln_urban_density_L + net_migr_pct + ln_gdp_pc | iso3 + period_f,
    data = panel_fixed,
    vcov = ~iso3
)
bs4 <- feols(
    bpcr ~ ln_urban_density_L +
        net_migr_pct +
        ln_gdp_pc +
        urban_pop_share_L |
        iso3 + period_f,
    data = panel_fixed,
    vcov = ~iso3
)

# ── Main finding = preferred dynamic BpCR TWFE (lagged DV + controls) ─────────
m6 <- feols(
    bpcr ~ bpcr_L +
        ln_urban_density_L +
        net_migr_pct +
        ln_gdp_pc +
        urban_pop_share_L |
        iso3 + period_f,
    data = panel_fixed,
    vcov = ~iso3
)

# ── 2000–2020 subsample robustness (same preferred spec) ─────────────────────
panel_2000 <- panel_fixed |> filter(year >= 2000)
b_2000 <- feols(
    bpcr ~ bpcr_L +
        ln_urban_density_L +
        net_migr_pct +
        ln_gdp_pc +
        urban_pop_share_L |
        iso3 + period_f,
    data = panel_2000,
    vcov = ~iso3
)
l_2000 <- feols(
    lcrpgr_log ~ lcrpgr_L +
        ln_urban_density_L +
        net_migr_pct +
        ln_gdp_pc +
        urban_pop_share_L |
        iso3 + period_f,
    data = panel_2000,
    vcov = ~iso3
)

# ── Shrinking population robustness ──────────────────────────────────────────
b_shrink <- feols(
    bpcr ~ bpcr_L +
        ln_urban_density_L +
        net_migr_pct +
        ln_gdp_pc +
        urban_pop_share_L +
        shrinking |
        iso3 + period_f,
    data = panel_fixed,
    vcov = ~iso3
)

# ── Active/Passive decomposition (same sample as regression: panel_fixed) ─────
decomp <- panel_fixed |>
    filter(!is.na(pop_growth)) |>
    mutate(
        dln_B = bpcr + pop_growth,
        land_type = case_when(
            bpcr > 0 & dln_B > 0 ~ "Active sprawl",
            bpcr > 0 & dln_B <= 0 ~ "Passive sprawl",
            bpcr < 0 & dln_B <= 0 ~ "Active densification",
            bpcr < 0 & dln_B > 0 ~ "Passive densification",
            TRUE ~ NA_character_
        ),
        active_share = dln_B / bpcr,
        passive_share = -pop_growth / bpcr
    )

# ── Within R² helper ─────────────────────────────────────────────────────────
glance_custom.fixest <- function(x, ...) {
    wr2 <- tryCatch(unname(fixest::r2(x)["wr2"]), error = function(e) NA_real_)
    out <- data.frame(within_r2 = wr2, check.names = FALSE)
    lab <- c(
        iso3 = "Country FE",
        period_f = "Period FE",
        period = "Period FE",
        Kennziffer = "District FE",
        il = "Province FE",
        year = "Year FE"
    )
    fes <- tryCatch(x$fixef_vars, error = function(e) character(0))
    for (fe in fes) {
        nm <- if (fe %in% names(lab)) lab[[fe]] else paste0(fe, " FE")
        out[[nm]] <- "Yes"
    }
    out
}

# ── Shared aesthetics ─────────────────────────────────────────────────────────
# significant vars first, then n.s., interaction last
cm_lcrpgr <- c(
    "lcrpgr_L" = "LCRPGR (log)[t−1]",
    "ln_urban_density_L" = "ln(Urban Density)[t-1]",
    "net_migr_pct" = "Net Migration (% Pop)",
    "ln_gdp_pc" = "ln(GDP per capita)",
    "urban_pop_share_L" = "Urban Pop. Share[t-1]"
)
cm_bpcr <- c(
    "bpcr_L" = "BpCR[t−1]",
    "ln_urban_density_L" = "ln(Urban Density)[t-1]",
    "net_migr_pct" = "Net Migration (% Pop)",
    "ln_gdp_pc" = "ln(GDP per capita)",
    "urban_pop_share_L" = "Urban Pop. Share[t-1]"
)

gm <- tribble(
    ~raw, ~clean, ~fmt,
    "nobs", "Observations", 0,
    "within_r2", "Within R²", 3,
    "Country FE", "Country FE", 0,
    "Province FE", "Province FE", 0,
    "District FE", "District FE", 0,
    "Period FE", "Period FE", 0,
    "Year FE", "Year FE", 0
)

# ── Inline-stat helpers (keep prose numbers locked to the actual models) ──────
sig <- function(p) {
    ifelse(p < .01, "***", ifelse(p < .05, "**", ifelse(p < .1, "*", "")))
}
cf <- function(m, v, d = 4) {
    ct <- fixest::coeftable(m)
    sprintf(paste0("%+.", d, "f%s"), ct[v, 1], sig(ct[v, 4]))
}
wr2 <- function(m) sprintf("%.3f", fixest::r2(m)["wr2"])

tbl_style <- function(gt_tbl, dv = NULL, meta = NULL) {
    # Put the dependent variable as a spanner at the very top, over all model
    # columns (i.e. every column except the first "term" column).
    if (!is.null(dv)) {
        model_cols <- gt_tbl[["_boxhead"]][["var"]][-1]
        lab <- if (!is.null(meta)) {
            gt::html(paste0(
                dv,
                "<br><span style='font-weight:400; font-size:0.6em;'>(",
                meta,
                ")</span>"
            ))
        } else {
            dv
        }
        gt_tbl <- gt::tab_spanner(
            gt_tbl,
            label = lab,
            columns = gt::all_of(model_cols)
        )
    }
    gt_tbl |>
        gt::tab_options(
            table.font.size = gt::px(21),
            column_labels.font.weight = "bold",
            data_row.padding = gt::px(3),
            heading.align = "left"
        ) |>
        gt::tab_style(
            style = gt::cell_text(weight = "bold"),
            locations = gt::cells_column_labels()
        ) |>
        gt::tab_style(
            style = gt::cell_text(weight = "bold", size = gt::px(25)),
            locations = gt::cells_column_spanners()
        )
}

# Dynamic-panel GMM results (computed in 02_analysis/24_dynamic_gmm.R; plm is run
# there, not here, so dplyr::lag stays unmasked for the panel build above).
gmm_res <- readRDS(here::here("03_datasets/processed/dynamic_gmm.rds"))

# Auto-fill the Final-Model card chips from the Arellano-Bond column so they never
# go stale. main_combo rows 1-5 are: 1 BpCR[t-1], 2 ln(Urban Density),
# 3 Net Migration, 4 ln(GDP pc), 5 Urban Pop. Share. Cell = "+0.6057***<br>(se)".
ab_coef <- function(row, digits = 4) {
    cell <- gmm_res$main_combo[["Arellano-Bond"]][row]
    cs <- sub("<br>.*$", "", cell) # "+0.6057***"
    num <- as.numeric(sub("^([+-]?[0-9.]+).*$", "\\1", cs))
    star <- sub("^[+-]?[0-9.]+", "", cs) # "***" or ""
    val <- sub("-", "−", sprintf("%+.*f", digits, num)) # proper minus sign
    if (nzchar(star)) {
        paste0(val, gsub("*", "\\*", star, fixed = TRUE))
    } # escape * for Pandoc
    else {
        paste0(val, " (n.s.)")
    }
}
gmm_gt <- function(tb, meta = NULL, fontpx = NULL, note = NULL) {
    n_terms <- 5L # coefficient rows before the diagnostics block
    est_cols <- setdiff(names(tb), "Term") # 3 cols (FE/AB/BB), 4 (+OLS) or 7 (FE stepwise + AB/BB)
    # Same look as every other model table: centred numbers, gof separator,
    # then the shared tbl_style (spanner + meta + fonts + border).
    g <- tb |>
        gt::gt() |>
        gt::cols_label(Term = "") |>
        gt::fmt_markdown(columns = gt::all_of(est_cols)) |> # render any <br>(p=…) line
        gt::cols_align(align = "center", columns = gt::all_of(est_cols)) |>
        gt::tab_style(
            style = gt::cell_borders(
                sides = "top",
                color = "#bbb",
                weight = gt::px(1.5)
            ),
            locations = gt::cells_body(rows = n_terms + 1L)
        )
    # combo table: FE/GMM group spanners added BEFORE the DV spanner so the
    # DV + meta line sits ON TOP and the FE/GMM groups appear directly below it.
    if ("Base" %in% est_cols) {
        g <- g |>
            gt::tab_spanner(
                label = "Fixed Effects",
                id = "fe_grp",
                columns = c(
                    "Base",
                    "+ Density",
                    "+ Net Migr.",
                    "+ GDP",
                    "+ Urban %"
                )
            ) |>
            gt::tab_spanner(
                label = "GMM (Bias-Corrected)",
                id = "gmm_grp",
                columns = c("Arellano-Bond", "Blundell-Bond")
            )
    }
    g <- g |>
        tbl_style(
            dv = paste0(
                "Dependent variable: BpCR",
                "<br><span style='font-weight:400; font-size:0.58em;'>",
                "&Delta;ln(built-up area &divide; population) / z &nbsp;&equiv;&nbsp; log-LCR &minus; log-PGR",
                "</span>"
            ),
            meta = meta
        )
    if (!is.null(fontpx)) {
        g <- g |> gt::tab_options(table.font.size = gt::px(fontpx))
    }
    if (!is.null(note)) {
        g <- g |>
            gt::tab_source_note(source_note = gt::html(note)) |>
            gt::tab_options(source_notes.font.size = gt::px(15))
    }
    if ("Base" %in% est_cols) {
        # FE | GMM vertical divider + notes separator
        g <- g |>
            gt::tab_style(
                style = gt::cell_borders(
                    sides = "right",
                    color = "#D3D3D3",
                    weight = gt::px(2)
                ),
                locations = list(
                    gt::cells_body(columns = "+ Urban %"),
                    gt::cells_column_labels(columns = "+ Urban %")
                )
            ) |>
            gt::tab_style(
                style = gt::cell_borders(
                    sides = "top",
                    color = "#D3D3D3",
                    weight = gt::px(1.5)
                ),
                locations = gt::cells_source_notes()
            ) |>
            gt::tab_style(
                # FE/GMM group spanners smaller than the DV spanner
                style = gt::cell_text(size = gt::px(16), weight = "bold"),
                locations = gt::cells_column_spanners(
                    spanners = c("fe_grp", "gmm_grp")
                )
            )
    }
    g
}
```

# Introduction {.section-slide visibility="uncounted"}

## SDG 11.3.1: Ratio of Land Consumption Rate to Population Growth Rate {.larger}

::: {.columns}

::: {.column width="50%"}
### A planet in motion

<iframe src="population_counter.html" title="Live world population counter" scrolling="no" style="width:100%; height:430px; border:none; border-radius:10px; overflow:hidden;"></iframe>

[Source: Live World Population Counter [@worldometers2024].]{style="font-size:0.60em; opacity:0.55; display:block; margin-top:-35px;"}

- Between 2000 and 2030 the world's urban land is forecast to grow by **1.2 million km²**, an area about the size of **South Africa**, roughly **tripling** in three decades, as housing, roads and industry keep spreading outward [@seto2012global].

:::

::: {.column width="50%"}
### Why It Matters?: Sprawl is a development problem

<div style="display:flex; flex-direction:column; gap:16px; margin-top:12px; font-size:0.95em; line-height:1.35;">

<div style="display:flex; align-items:center; gap:13px; background:#fff7ed; border-left:5px solid #d97706; padding:11px 15px; border-radius:11px; box-shadow:0 1px 5px rgba(0,0,0,0.06);">
<svg width="32" height="32" viewBox="0 0 24 24" fill="none" stroke="#d97706" stroke-width="1.9" stroke-linecap="round" stroke-linejoin="round" style="flex:0 0 auto;"><path d="M3 7a2 2 0 0 1 2-2h12v3"/><path d="M3 7v10a2 2 0 0 0 2 2h14a1 1 0 0 0 1-1v-9a1 1 0 0 0-1-1H5a2 2 0 0 1-2-2z"/><circle cx="16.5" cy="13" r="1.3" fill="#d97706" stroke="none"/></svg>
<div><strong>Public Finance:</strong> <br>More land per person makes roads, water and transit cost far more per resident</div>
</div>

<div style="display:flex; align-items:center; gap:13px; background:#eff6ff; border-left:5px solid #2563eb; padding:11px 15px; border-radius:11px; box-shadow:0 1px 5px rgba(0,0,0,0.06);">
<svg width="32" height="32" viewBox="0 0 24 24" fill="none" stroke="#2563eb" stroke-width="1.9" stroke-linecap="round" stroke-linejoin="round" style="flex:0 0 auto;"><circle cx="9" cy="8" r="3.2"/><path d="M3 20a6 6 0 0 1 12 0"/><path d="M16 5.3a3 3 0 0 1 0 5.4"/><path d="M21 20a6 6 0 0 0-4-5.6"/></svg>
<div><strong>Equity &amp; Inclusion:</strong> <br>Sprawl pushes poorer residents to the periphery and fuels informal settlements</div>
</div>

<div style="display:flex; align-items:center; gap:13px; background:#f0fdf4; border-left:5px solid #16a34a; padding:11px 15px; border-radius:11px; box-shadow:0 1px 5px rgba(0,0,0,0.06);">
<svg width="32" height="32" viewBox="0 0 24 24" fill="none" stroke="#16a34a" stroke-width="1.9" stroke-linecap="round" stroke-linejoin="round" style="flex:0 0 auto;"><path d="M11 20A7 7 0 0 1 4 13C4 7 11 4 20 4c0 9-3 16-9 16z"/><path d="M4 20c4-2 7-5 9-9"/></svg>
<div><strong>Land &amp; Climate:</strong> <br>Sprawl permanently builds over farmland, locks cities into car dependence, and drives high-emission growth</div>
</div>

</div>

:::

:::

### But growth itself is not the problem

→ Cities **must** grow to house more people; that is normal.
<br>→ What matters is **how much land each new resident uses**. 
<br>→ If land grows **at the same pace as** population, the city stays **compact**. 
<br>→ If land grows **much faster** than population, it **sprawls**.
<br>→ So the problem is **sprawl, not growth**.

::: {.notes}
We are in the largest urban expansion in history. Between 2000 and 2030 the world's urban land is forecast to grow by 1.2 million square kilometres, an area about the size of South Africa, which would roughly triple it in three decades.

And as you can see from the live counter on the left, the world population is growing fast.

But growth itself is not the problem. Cities must grow

The question is how much land each new resident uses. 

If the land and population grows at the same pace, the city stays compact
But if the land grows faster than population then cities sprawls

And the Sprawl is a development problem:
Why? Because 

Sprawl causes higher infrastructure cost per person, 
Sprawl pushes poorer people to the edge of the cities, 
and it causes farmland loss and locks cities into car dependence, and drives high emission growth.


To track this United Nation created and indicator – SDG 11.3.1. 

This research asks: is that indicator reliable, and what drives it?"
:::


# Measurement {.section-slide visibility="uncounted"}

## SDG Indicator 11.3.1: How Is It Calculated?

::: {.columns}

::: {.column width="58%"}
### Computing 11.3.1, step by step [@unhabitat2025metadata]

#### Step 0: Define the analysis area: the *urban* area

- Computed **inside cities**, not over whole countries
- Urban area delineated by the **Degree of Urbanisation (DEGURBA)** → **GHS-SMOD** grid
- Built-up & population below measured **only within urban cells**

#### Step 1: Measure built-up area and population at two points in time

**Where:**

::: {style="display:flex; gap:1.6em; font-size:0.82em;"}
::: {style="flex:1;"}
- $V_{past}$: Urban built-up (m²), start of period
- $V_{present}$: Urban built-up (m²), end of period
- $t$: Years in the built-up window (= 5)
:::
::: {style="flex:1;"}
- $Pop_t$: Urban population, start of period
- $Pop_{t+n}$: Urban population, end of period
- $y$: Years in the population window (= 5)
:::
:::

#### Step 2: Land Consumption Rate (LCR)

$$LCR = \frac{V_{present} - V_{past}}{V_{past}} \times \frac{1}{t}$$

How fast is the city's physical footprint expanding per year? → **Arithmetic (linear) rate**

#### Step 3: Population Growth Rate (PGR)

$$PGR = \frac{\ln\!\left(Pop_{t+n} / Pop_t\right)}{y}$$

How fast is the population growing per year? → **Logarithmic (exponential) rate**

#### Step 4: The Ratio

$$\boxed{LCRPGR = \frac{LCR}{PGR}}$$

:::

::: {.column width="42%"}

### How to read the result

:::: {style="font-size:1.7em;"}

::: {.callout-caution title="LCRPGR < 1: Densification"}
Population grows faster than built-up area. More people per km² of city over time.
:::

::: {.callout-tip title="LCRPGR = 1: Balanced [SDG Target]"}
Land consumption and population grow at equal rates. Sustainable expansion.
:::

::: {.callout-warning title="LCRPGR > 1: Sprawl"}
Built-up area expands faster than population. Rising infrastructure costs per resident, farmland loss, transit failure.
:::

::::

:::

:::
::: {.notes}
"So how is 11.3.1 actually computed?
First, it's measured inside cities only – urban cells are bounded by the EU's Degree of Urbanisation method on the GHS-SMOD grid, not over whole countries.

Then we take built-up area and population at two points in time and compute two rates: the Land Consumption Rate measures how fast the built-up footprint grows – and the Population Growth Rate measures how fast the population grows. 

The indicator is simply their ratio.

Reading it is intuitive, as the right side shows: below 1, population outpaces land – the city densifies; 

at 1, they grow together – this is the SDG target; 

above 1, land outpaces people – sprawl.

But notice one detail: the LCR is an arithmetic rate while the PGR is a logarithmic one. That mismatch is exactly the problem I turn to next."
:::
---

## A Flawed Yardstick: Two Structural Problems {.larger}

::: {.columns}

::: {.column width="48%"}
### Problem 1: Mathematical asymmetry

LCR uses an **arithmetic** (linear) rate; PGR uses a **logarithmic** (exponential) rate.

$$\underbrace{\frac{V_{present} - V_{past}}{V_{past}} \times \frac{1}{t}}_{\text{arithmetic LCR}} \;\div\; \underbrace{\frac{\ln\!\left(Pop_{t+n}/Pop_t\right)}{y}}_{\text{logarithmic PGR}}$$

A city with 5% annual growth in **both** built area and population does **not** produce LCRPGR = 1, the two "5%s" are on different mathematical scales. Taking a ratio of two differently-defined rates is dimensionally inconsistent: cross-country comparisons are distorted by the formula, not by real urbanisation.



:::

::: {.column width="48%"}
### Problem 2: Division instability

When $PGR \approx 0$ (shrinking or stable populations), LCRPGR diverges to $\pm\infty$.

::: {style="font-size:1.2em;"}
| Scenario | LCR | PGR | LCRPGR |
|:---------|:---:|:---:|:------:|
| Normal Growth | 2% | 1.5% | 1.33 |
| Shrinking Population | 2% | −0.1% | **−20** |
| Stable Population | 2% | 0.001% | **2,000** |
:::

The same **2% land growth** reads as 1.33, −20, or 2,000 depending only on how close PGR is to zero. Because the denominator can vanish, the blow-up is **structural**: no winsorising or filtering repairs it, it only hides the symptom.

:::

:::

::: {.callout-note title="Worked example: 5%/yr in both, over 5 years"}
Same growth factor $g = 1.05^{5} = 1.276$ for built area **and** population:
$$LCR=\frac{g-1}{5}=\frac{0.276}{5}=5.53\%,\qquad PGR=\frac{\ln g}{5}=\frac{0.244}{5}=4.88\%$$
$$\Rightarrow \mathbf{LCRPGR = 1.13}\ \text{(not 1)}$$
Identical growth, yet the formula alone signals "sprawl" because $(g-1) > \ln g$.
:::
:::{.notes}
 "The official ratio has two structural flaws.

 First, a mathematical asymmetry: the numerator (LCR) is an arithmetic rate, the denominator (PGR) a logarithmic one.
 
 If we look at example below – a city growing 5% over 5 years in both land and population. 
 
 Identical growth, yet the formula returns 1.13, not 1 – it falsely signals 'sprawl' purely because the two rates are defined on different scales.

 Second problem, division instability: when population growth is near zero – shrinking or stable cities – the denominator vanishes and the ratio explodes. The same 2% land growth reads as 1.33, as minus 20, or as 2,000, depending only on how close PGR is to zero.

 This blow-up is structural. So the official yardstick is unreliable, which is exactly why we need a better metric."
:::
---

## Built-up Area per Capita: The Metadata's Secondary Indicator

::: {.columns}

::: {.column width="65%"}
### Recommended in the Official Metadata

The 11.3.1 metadata [@unhabitat2025metadata] itself states LCRPGR is **hard to interpret** and proposes **two secondary indicators**, built from the *same inputs* (§4.f):

#### Secondary indicator 1: Built-up area per capita
$$\text{BUP} = \frac{UrBU_t}{Pop_t}\quad(\text{m}^2/\text{person})$$
$\rightarrow$ Measures **land consumed per person**: the *state* at one point in time.

#### Secondary indicator 2: Total change in built-up area
$$\text{Total change} = \frac{UrBU_{t+n} - UrBU_t}{UrBU_t}\quad(\%)$$
$\rightarrow$ Measures **total urban expansion** over the period (omits population).

We analyse the **growth rate of built-up area per capita**, the dynamic form of secondary indicator 1 (log form following [@Lu2026]):

$$\boxed{BpCR = \frac{\ln\!\left(\dfrac{UrBU_t/Pop_t}{UrBU_{t-z}/Pop_{t-z}}\right)}{z} = LCR_{\log} - PGR_{\log}}$$
$\rightarrow$ Measures the **annual trend in per-capita land use**: densifying ($<0$) or sprawling ($>0$).

**Where:**

::: {style="display:flex; gap:1.6em; font-size:0.82em;"}
::: {style="flex:1;"}
- $UrBU_t$: urban built-up (m²) in year $t$
- $Pop_t$: urban population in year $t$
- $z$: period length in years (= 5)
:::
::: {style="flex:2;"}
- $t{-}z$ / $t{+}n$: earlier / later year
- $LCR_{\log}=\ln(UrBU_t/UrBU_{t-z})/z$: annual log growth of built-up
- $PGR_{\log}=\ln(Pop_t/Pop_{t-z})/z$: annual log growth of population
:::
:::

:::

::: {.column width="35%"}
### From a Level to a Trend

::: {style="background:#FBCA8D; border-radius:12px; padding:10px 16px; margin-bottom:4px;"}
**&#9312;&ensp;Built-up per capita[@unhabitat2025metadata]** 

$$\text{BUP} = UrBU \,/\, Pop$$

<span style="opacity: 0.7;"><span style="color: #9E5905;">The level</span>: land per person, right now.</span>
:::

::: {style="text-align:center; color:#FBCA8D; font-size:1.4em; line-height:1; margin:0;"}
&#8595;
:::

::: {style="background:#FAB561; border-radius:12px; padding:10px 16px; margin-bottom:4px;"}
**&#9313;&ensp;Annual growth of BUP[@Lu2026]** 

$$BpCR = \Delta\ln(\text{BUP})/z = LCR_{\log}-PGR_{\log}$$

<span style="opacity: 0.7;"><span style="color: #9E5905;">The change</span>: is it rising or falling?</span>
::: 

::: {style="text-align:center; color:#F78B08; font-size:1.4em; line-height:1; margin:0;"}
&#8595;
:::

::: {style="background:#f89d2a; border-radius:12px; padding:10px 16px; margin-bottom:4px;"}
**&#9314;&ensp;BpCR** &ensp;<span style="opacity:.6; font-weight:400;">(Our Dependent Variable)</span>

<span style="color:#b91c1c; font-weight:700;">&#9650; BpCR &gt; 0</span> &emsp; Land &gt; People &emsp; **Sprawl**

<span style="color:#4b5563; font-weight:700;">&#9670; BpCR = 0</span> &emsp; Land = People &emsp; **Balanced**

<span style="color:#047857; font-weight:700;">&#9660; BpCR &lt; 0</span> &emsp; People &gt; Land &emsp; **Densification**
:::

:::

:::

::: {.notes}
"The 11.3.1 metadata itself admits the ratio is hard to interpret and offers two secondary indicators, built from the same data – the first being built-up area per capita(BUP): land consumed per person.
That's a level, a snapshot. 

We turn it into a trend – its annual growth rate, in log form, following Lu and Weng.

We call it BpCR(growth rate of built-up area per capita):

The key is the structure: it's a difference of two logarithmic rates – same scale, and a subtraction instead of a division. 

That single change fixes both flaws: no asymmetry, and no blow-up when population is flat.

The interpretation stays simple: positive means sprawl, negative means densification."

:::

# Literature {.section-slide visibility="uncounted"}

## Literature Review: Drivers & Measurement

::: {.columns}

::: {.column width="50%"}
### Drivers of Urban Expansion

::: {style="font-size:0.93em; line-height:1.32;"}

**A Meta-Analysis of Global Urban Land Expansion** [@seto2011meta]

- **Data:** 326 remote-sensing studies, global, 1970&ndash;2000
- **Scope:** rates and drivers of global urban land expansion
- **Finding:** urban land grew **faster than population** almost everywhere; income drives it in most regions, population growth in India and Africa

**Determinants of Urban Sprawl in European Cities** [@oueslati2015sprawl]

- **Data:** panel of 282 European cities (1990, 2000, 2006)
- **Scope:** econometric determinants of sprawl (monocentric model)
- **Finding:** **rising income and population** expand the urban footprint; cities keep spreading even where population is flat

:::
:::

::: {.column width="50%"}
### Measurement & SDG 11.3.1

::: {style="font-size:0.93em; line-height:1.32;"}

**Ratio of Land Consumption Rate to Population Growth Rate: Analysis of Different Formulations Applied to Mainland Portugal** [@nicolau2019lcrpgr]

- **Data:** mainland Portugal, municipality level
- **Scope:** SDG 11.3.1: UN LCRPGR vs an LUE-derived formula
- **Finding:** the UN **LCRPGR formula is structurally unstable**; a log / LUE-based alternative is more robust

**From Data Inconsistency to Reliable SDG 11.3.1 Monitoring: A Global Multi-Dataset Assessment of Land Use Efficiency** [@Lu2026]

- **Data:** 6 built-up &times; 4 population datasets, 176 countries, 2000&ndash;2020
- **Scope:** reliability of LUE (11.3.1) across datasets
- **Finding:** LUE estimates **swing widely with dataset choice**; a **log-based** framework is needed for reliable monitoring

:::
:::

:::

::: {style="background:rgba(248,157,43,0.15); border:2px solid #f89d2b; border-radius:13px; padding:12px 20px; margin-top:12px; font-size:0.95em; box-shadow:0 3px 14px rgba(248,157,43,0.22);"}
[<svg width="27" height="27" viewBox="0 0 24 24" fill="none" stroke="#b1690f" stroke-width="2" stroke-linecap="round" stroke-linejoin="round" style="vertical-align:-0.3em;"><path d="M9 18h6"/><path d="M10 22h4"/><path d="M12 2a7 7 0 0 0-4 12.7c.6.5 1 1.3 1 2.1h6c0-.8.4-1.6 1-2.1A7 7 0 0 0 12 2z"/></svg> **Our Contribution**]{style="color:#b1690f; font-size:1.18em;"}

- **Metric:** The **log-based BpCR** (Lu & Weng; log case also in Nicolau)
- **Method:** Make it **dynamic** with a two-way FE + **GMM(Generalized Method of Moments)** panel of **193 UN Member Countries (1985–2020)** measuring **persistence** (path-dependence)
:::

::: {.notes}
"The literature splits in two. On drivers, Seto and Oueslati show income and population push cities to expand. On measurement, Nicolau and Lu show the official ratio is unstable and recommend a log-based metric.

Our contribution sits at the intersection: we take that log metric and add dynamics – a GMM(Generalized Method of Moments) panel of 193 countries – and find that within a country, it is in-migration, not income, that robustly goes with densification."
:::


# Data {.section-slide visibility="uncounted"}

## Data Sources

::: {.columns}

::: {.column width="33%"}
::: {.source-card}
<div class="source-card-header">
<div><h3>GHSL</h3><p class="source-subtitle"><strong>Global Human Settlement Layer</strong><br><em>European Commission, Joint Research Centre / Copernicus</em></p></div>
<a href="https://ghsl.jrc.ec.europa.eu/" target="_blank"><img src="logos/ghsl_logo.svg" alt="GHSL Logo" class="source-logo-inline"></a>
</div>

- **GHS-BUILT-S** (Built-up Surface), **GHS-POP** (Population), **GHS-SMOD** (Settlement Model)
- Built-up classified from **Landsat & Sentinel-2** by machine learning; GHS-POP disaggregates census onto that surface; **GHS-SMOD applies the Degree of Urbanisation (DEGURBA)** to flag *urban* cells
- **Available 1975–2030** at 5-year epochs (1975–2020 observed; 2025 & 2030 *modelled projections*)

::: {style="font-size:0.80em; opacity:0.7; margin-top:4px;"}
European Union, @pesaresi2024ghsl; @schiavina2023smod
:::
:::
:::

::: {.column width="33%"}
::: {.source-card}
<div class="source-card-header">
<div><h3>GAUL 2024/2025</h3><p class="source-subtitle"><strong>Global Administrative Unit Layers</strong><br><em>FAO, Food and Agriculture Organization</em></p></div>
<a href="https://data.apps.fao.org/catalog/dataset/global-administrative-unit-layers-gaul-2024" target="_blank"><img src="logos/fao_logo.svg" alt="FAO Logo" class="source-logo-inline"></a>
</div>

- UN-validated administrative boundaries: politically **neutral**
- **Level 0** Country Level
- **Level 1** Province/Region Level
- **Global coverage** including all UN member states

::: {style="font-size:0.80em; opacity:0.7; margin-top:4px;"}
Food and Agriculture Organization, @gaul2024
:::
:::
:::

::: {.column width="33%"}
::: {.source-card}
<div class="source-card-header">
<div><h3>UN</h3><p class="source-subtitle"><strong>UN Statistics & Population Division</strong><br><em>United Nations</em></p></div>
<a href="https://data.un.org/" target="_blank"><img src="logos/un_logo.svg" alt="UN Logo" class="source-logo-inline"></a>
</div>

- **Socio-economic controls** (built-up, population, urban share & density come from GHSL/SMOD)

| Control | Source |
|:-------------|:-------|
| GDP per capita Constant 2020 US\$) | **UN SNAAMA** |
| Net migration (% of Population) | **UN DESA WPP 2024** |

::: {style="font-size:0.80em; opacity:0.7; margin-top:4px;"}
@unsd2025snaama; @undesa2024wpp; @unsd2024m49
:::

:::
:::

:::

::: {.full-width-map style="margin-top:6px; width:100%; display:block;"}
<iframe data-mapsrc="../04_outputs/figures/data_sources_map.html" src="about:blank" width="100%" height="510px" style="display:block; border:none; border-radius:12px;"></iframe>
:::

::: {.notes}
"I used three data sources, all are at the global level and official and what UN uses.

GHSL, the European Commission's Global Human Settlement Layer, gives us built-up area, population and the urban area mask – all derived from Landsat and Sentinel satellites, every five years from 1975.

Global Administrative Unit Layers, from the Food and Agriculture Organization of the UN, provides administrative boundaries for every country.

And from the UN come our socio-economic controls: GDP per capita and net migration

:::

# Descriptive Statistics {.section-slide visibility="uncounted"}

## Descriptive Statistics

```{r}
#| label: descriptive
#| echo: false
desc_vars <- c(
    bpcr = "BpCR",
    lcrpgr_log = "LCRPGR (log)",
    ln_gdp_pc = "ln(GDP per capita)",
    urban_pop_share_L = "Urban Pop. Share[t-1]",
    ln_urban_density_L = "ln(Urban Density)[t-1]",
    net_migr_pct = "Net Migration (% Pop)"
)
panel_fixed |>
    select(all_of(names(desc_vars))) |>
    pivot_longer(everything(), names_to = "var", values_to = "v") |>
    group_by(var) |>
    summarise(
        N = sum(!is.na(v)),
        Mean = mean(v, na.rm = TRUE),
        SD = sd(v, na.rm = TRUE),
        Min = min(v, na.rm = TRUE),
        Median = median(v, na.rm = TRUE),
        Max = max(v, na.rm = TRUE),
        .groups = "drop"
    ) |>
    mutate(Variable = desc_vars[var]) |>
    arrange(match(var, names(desc_vars))) |>
    select(Variable, N, Mean, SD, Min, Median, Max) |>
    gt() |>
    gt::fmt_number(columns = c(Mean, SD, Min, Median, Max), decimals = 3) |>
    gt::fmt_number(columns = N, decimals = 0) |>
    tbl_style() |>
    gt::tab_options(table.font.size = gt::px(30))
```

:::: {style="font-size:1.18em;"}

::: {.callout-important title="Read the LCRPGR row"}
**LCRPGR (log)** already screams the problem: **SD ≈ 18** and a range from **−568 to +281**, vs **BpCR**'s SD ≈ **0.017** and range **[−0.11, +0.11]**. A ~1,000× wider spread: driven by LCRPGR's division instability when PGR ≈ 0.
:::

::::
:::{.notes}
The panel covers 193 UN member countries; the estimation sample is 192 of them from 1985 to 2020, giving 1,499 observations.


The official LCRPGR has an Standard Deviation of ~18, ranging from −568 to +281 – the division blow-up, now visible in real data. 

BpCR stays in a tight band, SD ~0.017 – roughly a thousand times narrower.

So even before any regression, the data already shows which variable is usable – and which isn't.
:::

---

## Country Trends: Regional Representatives

<iframe data-mapsrc="../04_outputs/figures/desc_trend_regional.html" src="about:blank" width="100%" height="640px" allowtransparency="true" style="border:none; display:block; background:transparent;" scrolling="no"></iframe>

::: {.callout-note title="Key patterns"}
::: {style="font-size:1.42em;"}
- **China**: strongest, most persistent per-capita sprawl (BpCR ≈ +0.02 to +0.03), peaking around 2000 then easing to 0.016; LCRPGR drifts up to ~4.4.
- **Germany**: trends toward balance (BpCR 0.015 → 0.002), yet its LCRPGR swings wildly (−13.2 to +12.9).
- **Japan**: BpCR negative in the 1980s, then a low, stable ~+0.01, while its LCRPGR lurches from +14.6 to −6.7.
- **Nigeria & Brazil**: oscillate around zero in BpCR (no persistent sprawl), with low LCRPGR (~1).
- **Takeaway**: BpCR gives a stable, interpretable per-capita signal; LCRPGR does not.
:::
:::
:::{.notes}
Seven selected countries, both metrics. On the Left, official indicator LCRPGR is swinging from −13 to +15 – unreadable.

On the Right, BpCR is smooth, interpretable, and consistent.

China has the strongest sprawl, peaking ~2000 then easing;

the rest drift toward balance.


:::


# Empirical Analysis {.section-slide visibility="uncounted"}

## Research Question & Hypotheses

### [Research question]{style="font-size:1.4em"}

::: {style="font-size:1.5em;"}
**What economic and structural factors drive within-country changes in per-capita urban land use over time, and how persistent are these trajectories?**
:::

::: {.callout-note title="Measurement sub-question" style="font-size:1.5em;"}
Does the official SDG 11.3.1 indicator (**LCRPGR**) reliably capture this, or does the **choice of metric** change the conclusions?
:::

### [Hypotheses]{style="font-size:1.4em"}

::: {style="font-size:2.7em;"}
| | Hypothesis | Expected Sign |
|:-:|:------------------------------------------------|:----:|
| **H1** | BpCR measures efficiency better than LCRPGR | Higher Fit |
| **H4** | Net in-migration → less land/capita | − |
| **H5** | Per-capita land use is path-dependent | + |
| **H2** | Higher urban density → less land/capita | − |
| **H3** | Higher income → less land/capita | − |
:::

:::{.notes}
My research question is
what drives within-country changes in per-capita urban land use, how persistent are they – and does the metric choice change the answer?

Five hypotheses, ordered the way the results come out:

H1, BpCR measures efficiency better than LCRPGR;

H4, in-migration densifies;

H5, land use is path-dependent;

H2, denser cities use less land per person;

and H3, richer countries densify rather than sprawl.


:::

## Empirical Strategy

### One specification, two dependent variables

::: {style="background:#f89d2b; border:3px solid #b1690f; border-radius:10px; padding:4px 14px; margin:10px auto; max-width:99%;"}
$$\small \text{Metric}_{i,t} = \rho\,\text{Metric}_{i,t-1} + \beta_1\ln(UrbDens)_{i,t} + \beta_2\,UrbShare_{i,t} + \beta_3\ln(GDPpc)_{i,t} + \beta_4\,NetMigr_{i,t} + \mu_i + \delta_t + \varepsilon_{i,t}$$
:::

::: {style="text-align:left; font-size:1.1em; background:rgba(248,157,43,0.14); border:2px solid #f89d2b; border-radius:10px; padding:10px 18px; margin:10px auto; max-width:99%;"}
**Two-stage plan**

**①** Compare the metric **statically** (drop $\rho\,\text{Metric}_{i,t-1}$) to isolate **LCRPGR vs BpCR**

**②** **Add the lagged DV** for dynamics and address its Nickell bias with **GMM** *(Final Model)*
:::

::: {.columns}

::: {.column style="width:47%; margin-right:5%;"}
**The same model is estimated twice; only $Y$ changes:**

::: {style="font-size:1.3em;"}
| | Model 1 | Model 2 |
|:--|:-------:|:-------:|
| **DV** | **LCRPGR** (UN) | **BpCR** (Ours) |
| FE | Country + Period | Country + Period |
| SE | Clustered (country) | Clustered (country) |
| N | `r nobs(l5_int)` | `r nobs(b5_int)` |
:::


- **Dynamic two-way FE panel**: country + period fixed effects ($\mu_i,\ \delta_t$)
- **Same RHS, two DVs** (LCRPGR vs BpCR): any divergence isolates the *metric*, not the data
- **Lagged DV** ($\rho\,Y_{t-1}$): captures inertia; the FE $\rho$ is Nickell-biased downward, GMM-corrected (see Final Model)
- **Controls** (contemporaneous, at $t$): urban density, urban share, GDP pc, net migration
- **192 countries $\times$ 8 five-year periods** (1985–2020), 1,499 obs; SE clustered by country

:::

::: {.column style="width:48%;"}
### Why Each Variable Enters

::: {style="font-size:1.3em;"}
| Variables | Rationale |
|:-----|:-------------|
| **Lagged DV** ($\rho$)| Captures inertia: once a country sprawls, it tends to keep sprawling |
| **ln(Urban Density)** | Urban form: denser cities consume less new land per resident |
| **Urban Pop. Share** | How urbanised the country already is; less rural land left to convert |
| **ln(GDP Per Capita)** | Tests whether richer countries use more or less built-up land per person |
| **Net Migration** | Demographic pressure: do newcomers densify cities or push outward expansion? |
| **Country & Period FE** | Absorb fixed geography and institutions, plus shocks shared in each period |
:::

:::

:::
:::{.notes}
The empirical strategy is straightforward. I run one model and estimate it twice – once with the official LCRPGR, once with BpCR. Everything else stays identical: same controls, country and period fixed effects, errors clustered by country. So we're really comparing the metrics, not the data.

And we do it in two steps. First, statically: does switching the metric change the story? Then we add the lagged dependent variable for dynamics, and correct its downward Nickell bias with GMM (Generalized Method of Moments).
:::


---

## Model 1: UN Approach: LCRPGR Regression

::: {style="border:3px solid #f89d2b; border-radius:10px; padding:2px 16px; margin:8px auto; max-width:83%;"}
$$\small LCRPGR_{i,t} = \beta_1\ln(UrbDens)_{i,t} + \beta_2\,NetMigr_{i,t} + \beta_3\ln(GDPpc)_{i,t} + \beta_4\,UrbShare_{i,t} + \mu_i + \delta_t + \varepsilon_{i,t}$$
:::

::: {.eqmatch}
```{r}
#| label: lcrpgr-stepwise
#| echo: false
modelsummary(
    list(
        "Density" = ls1,
        "+ Net Migr." = ls2,
        "+ GDP pc" = ls3,
        "+ Urban %" = ls4
    ),
    coef_map = cm_lcrpgr,
    gof_map = gm,
    stars = c("*" = .10, "**" = .05, "***" = .01),
    fmt = fmt_sprintf("%.3f"),
    output = "gt",
    notes = "SE clustered at country level. DV: LCRPGR (log) = Land Consumption Rate ÷ Population Growth Rate (both log rates). Static two-way FE (no lagged DV); GHSL/SMOD + UN (GDP, net migration)."
) |>
    tbl_style(
        dv = "Dependent variable: LCRPGR (log)",
        meta = sprintf(
            "%d countries × %d periods, 1985–2020",
            n_distinct(panel_fixed$iso3),
            n_distinct(panel_fixed$period_f)
        )
    )
```
:::
:::{.notes}
Here's Model 1, using the official LCRPGR as the dependent variable, controls added one by one.

Nothing is significant, and the within-country R² is about 0.003 – so low. The metric is so noisy that the coefficients come out huge and unstable.

:::

---

## Model 2: Our Approach: BpCR Regression

::: {style="border:3px solid #f89d2b; border-radius:10px; padding:2px 16px; margin:8px auto; max-width:83%;"}
$$\small BpCR_{i,t} = \beta_1\ln(UrbDens)_{i,t} + \beta_2\,NetMigr_{i,t} + \beta_3\ln(GDPpc)_{i,t} + \beta_4\,UrbShare_{i,t} + \mu_i + \delta_t + \varepsilon_{i,t}$$
:::

::: {.eqmatch}
```{r}
#| label: bpcr-stepwise
#| echo: false
modelsummary(
    list(
        "Density" = bs1,
        "+ Net Migr." = bs2,
        "+ GDP pc" = bs3,
        "+ Urban %" = bs4
    ),
    coef_map = cm_bpcr,
    gof_map = gm,
    stars = c("*" = .10, "**" = .05, "***" = .01),
    fmt = fmt_sprintf("%.4f"),
    output = "gt",
    notes = "SE clustered at country level. DV: BpCR = Δln(built-up per capita)/z ≡ LCR − PGR. Static two-way FE (no lagged DV); GHSL/SMOD + UN (GDP, net migration)."
) |>
    tbl_style(
        dv = "Dependent variable: BpCR",
        meta = sprintf(
            "%d countries × %d periods, 1985–2020",
            n_distinct(panel_fixed$iso3),
            n_distinct(panel_fixed$period_f)
        )
    )
```
:::
:::{.notes}
Now Model 2 – exact same controls, same fixed effects. The only change is the dependent variable: BpCR instead of LCRPGR.

And everything changes. Under LCRPGR not a single control is distinguishable from zero – the largest t-statistic is 1.66. Swap in BpCR, changing nothing else, and the very same migration variable goes from a t of minus 1.1 to minus 5. I am deliberately not quoting a ratio of the two R-squareds here: they are shares of the variance of two different dependent variables, so their quotient means nothing. The point is detectability, not explained variance – the official ratio buries the signal in outlier noise, the log difference recovers it.

Same data, same model – just a better-built metric. This is the clearest evidence for H1.
:::
---

## Final Model: BpCR under Dynamic-Panel GMM

::: {style="border:3px solid #f89d2b; border-radius:10px; padding:2px 16px; margin:6px auto; max-width:100%;"}
$$\small BpCR_{i,t} = \rho\,BpCR_{i,t-1} + \beta_1\ln(UrbDens)_{i,t} + \beta_2\,NetMigr_{i,t} + \beta_3\ln(GDPpc)_{i,t} + \beta_4\,UrbShare_{i,t} + \mu_i + \delta_t + \varepsilon_{i,t}$$
:::

The dynamic FE model is built up control-by-control (first five columns); the last two columns target the lagged-DV **Nickell bias** with **Arellano-Bond** and **Blundell-Bond** GMM (instrumenting the lagged BpCR with its deeper lags).

::: {.columns}

::: {.column width="64%"}

::: {.leftalign}

```{r}
#| label: gmm-fullmodel
#| echo: false
gmm_gt(
    gmm_res$main_combo,
    meta = sprintf(
        "%d countries × %d periods, 1985–2020",
        n_distinct(panel_fixed$iso3),
        n_distinct(panel_fixed$period_f)
    ),
    fontpx = 14,
    note = paste0(
        "<b>* p&lt;0.10, ** p&lt;0.05, *** p&lt;0.01.</b><br>",
        "Standard errors clustered by country.<br>",
        "<b>Notes:</b><br>",
        "<b>1)</b> Cols 1&ndash;5: dynamic two-way FE (country + period FE), controls added stepwise. <br>",
        "<b>2)</b> <b>AB</b> = Arellano-Bond difference GMM; <b>BB</b> = Blundell-Bond system GMM, instrumenting the lagged BpCR with lags &ge; 2 (two-step, Windmeijer-robust SE).<br>",
        "<b>3)</b> <b>Hansen J</b> = over-identification test (p &gt; 0.10 &rarr; valid instruments); <b>AR(1)/AR(2)</b> = first/second-order serial correlation (AR(1) significant by construction; AR(2) p &gt; 0.10 &rarr; valid).<br>"
    )
)
```

:::

:::

::: {.column width="36%"}

:::: {style="font-size:0.8em; background:rgba(248,157,43,0.13); border-left:4px solid #b1690f; border-radius:8px; padding:5px 11px; margin:0 0 7px 0;"}
[<svg width="1.3em" height="1.3em" viewBox="0 0 24 24" fill="none" stroke="#b1690f" stroke-width="2.3" stroke-linecap="round" stroke-linejoin="round" style="vertical-align:-0.12em;"><path d="M3 12a9 9 0 0 1 15-6.7L21 8"/><polyline points="21 3 21 8 16 8"/><path d="M21 12a9 9 0 0 1-15 6.7L3 16"/><polyline points="3 21 3 16 8 16"/></svg> &nbsp;**BpCR[t−1]**]{style="font-size:1.14em; white-space:nowrap;"} &nbsp;[INERTIA]{style="background:#b1690f; color:#fff; border-radius:9px; padding:1px 8px; font-size:0.82em; font-weight:700;"}<br>
[ρ = `r ab_coef(1L, 2L)`]{style="background:rgba(0,0,0,0.07); border-radius:6px; padding:0 6px; font-weight:700;"} Here ρ is imprecise (SE 0.10) and its moments are marginal, so we do not read it: it sits *just* below the Bond bracket, within sampling error. Where the moments are clean: **ρ ≈ 0.46 (SE 0.07)**, close to the Nickell-implied 0.45; Blundell-Bond lands nearby too (0.45), but its own Hansen test also fails there, so we read that as coincidence, not corroboration.<br>[**H5** ✓ Supported: ~half of last period's BpCR persists]{style="color:#0e7c66; font-weight:700; font-size:0.85em;"}
::::

:::: {style="font-size:0.8em; background:rgba(192,57,43,0.10); border-left:4px solid #c0392b; border-radius:8px; padding:5px 11px; margin:7px 0;"}
[<svg width="1.3em" height="1.3em" viewBox="0 0 24 24" fill="none" stroke="#c0392b" stroke-width="2" stroke-linecap="round" stroke-linejoin="round" style="vertical-align:-0.12em;"><rect x="3" y="9" width="7" height="12" rx="1"/><rect x="13" y="3" width="8" height="18" rx="1"/><path d="M6 13h1M6 17h1M16 7h2M16 11h2M16 15h2"/></svg> &nbsp;**ln(Urban Density)[t−1]**]{style="font-size:1.14em; white-space:nowrap;"} &nbsp;[SIGN FLIPS]{style="background:#c0392b; color:#fff; border-radius:9px; padding:1px 8px; font-size:0.82em; font-weight:700;"}<br>
[β₁ = `r ab_coef(2L, 3L)`]{style="background:rgba(0,0,0,0.07); border-radius:6px; padding:0 6px; font-weight:700;"} BpCR contains **−ln P₍ₜ₎ + ln P₍ₜ₋₁₎**, so density inherits the sign of its own date: **−0.055\*\*\*** at *t*, **positive** at *t−1*. Clean *t−2* timing: **+0.011\*\***.<br>[**H2** ✗ Rejected: the compact-city sign was arithmetic]{style="color:#c0392b; font-weight:700; font-size:0.85em;"}
::::

:::: {style="font-size:0.8em; background:rgba(14,124,102,0.10); border-left:4px solid #0e7c66; border-radius:8px; padding:5px 11px; margin:7px 0;"}
[<svg width="1.3em" height="1.3em" viewBox="0 0 24 24" fill="none" stroke="#0e7c66" stroke-width="2" stroke-linecap="round" stroke-linejoin="round" style="vertical-align:-0.12em;"><circle cx="9" cy="8" r="3.2"/><path d="M3 20a6 6 0 0 1 12 0"/><path d="M16 5.3a3 3 0 0 1 0 5.4"/><path d="M21 20a6 6 0 0 0-4-5.6"/></svg> &nbsp;**Net Migration**]{style="font-size:1.14em; white-space:nowrap;"} &nbsp;[DENSIFICATION]{style="background:#0e7c66; color:#fff; border-radius:9px; padding:1px 8px; font-size:0.82em; font-weight:700;"}<br>
[β₂ = `r ab_coef(3L, 4L)`]{style="background:rgba(0,0,0,0.07); border-radius:6px; padding:0 6px; font-weight:700;"} **Our strongest result.** Migration moves with urban population **+0.88 pp/yr** but built-up only **+0.24** → the gap is densification. Holds in **all 3** development groups.<br>[**H4** ✓ Supported as an *association*: passes a strict-exogeneity test, but migrants may select into fast-building economies]{style="color:#0e7c66; font-weight:700; font-size:0.85em;"}
::::

:::: {style="font-size:0.8em; background:rgba(154,160,166,0.12); border-left:4px solid #9aa0a6; border-radius:8px; padding:5px 11px; margin:7px 0;"}
[<svg width="1.3em" height="1.3em" viewBox="0 0 24 24" fill="none" stroke="#9aa0a6" stroke-width="2" stroke-linecap="round" stroke-linejoin="round" style="vertical-align:-0.12em;"><path d="M3 3v18h18"/><polyline points="7 14 11 10 14 13 19 7"/></svg> &nbsp;**ln(GDP per capita)**]{style="font-size:1.14em; white-space:nowrap;"} &nbsp;[NOT SIGNIFICANT]{style="background:#9aa0a6; color:#fff; border-radius:9px; padding:1px 8px; font-size:0.82em; font-weight:700;"}<br>
[β₃ = `r ab_coef(4L, 3L)`]{style="background:rgba(0,0,0,0.07); border-radius:6px; padding:0 6px; font-weight:700;"} Sign is negative in **every** spec and group, but never significant once migration is measured over the right window.<br>[**H3** ✗ Not supported: consistent sign, insufficient precision]{style="color:#9aa0a6; font-weight:700; font-size:0.85em;"}
::::

:::: {style="font-size:0.8em; background:rgba(154,160,166,0.12); border-left:4px solid #9aa0a6; border-radius:8px; padding:5px 11px; margin:7px 0;"}
[<svg width="1.3em" height="1.3em" viewBox="0 0 24 24" fill="none" stroke="#9aa0a6" stroke-width="2" stroke-linecap="round" stroke-linejoin="round" style="vertical-align:-0.12em;"><circle cx="12" cy="12" r="9"/><path d="M8 12h8"/></svg> &nbsp;**Urban Pop. Share[t−1]**]{style="font-size:1.14em; white-space:nowrap;"} &nbsp;[NOT SIGNIFICANT]{style="background:#9aa0a6; color:#fff; border-radius:9px; padding:1px 8px; font-size:0.82em; font-weight:700;"}<br>
[β₄ = `r ab_coef(5L, 3L)`]{style="background:rgba(0,0,0,0.07); border-radius:6px; padding:0 6px; font-weight:700;"} The *level* of urbanisation adds nothing once the rest is controlled → urban **form** matters.<br>[Control · No dedicated hypothesis]{style="color:#9aa0a6; font-weight:700; font-size:0.85em;"}
::::

:::

:::

::: {.notes}
"Now the final model. I add the lagged dependent variable to capture the inertia effect.

That creates the Nickell bias: once a lagged dependent variable enters a fixed-effects model it is correlated with the demeaned error, which pushes persistence downward. The intended fix is dynamic-panel GMM. Arellano-Bond differences the equation and instruments the lagged term with its own deeper lags; Blundell-Bond adds the levels equation.

Now, a fair objection: difference GMM is exactly the estimator that is known to have weak instruments on persistent series – that is why Blundell and Bond built the system version. And in the full model my rho is 0.21, which sits just below the within estimate of 0.26, below the Bond bracket. Now, that crossing on its own means nothing – the gap is 0.04 against a standard error of 0.10, so it is well inside sampling error, and I will not hang an argument on it. But it is the classic weak-instrument symptom, so I checked it rather than assumed it away.

It is not weak instruments. Strip the model back to a pure AR(1) and difference GMM gives 0.46, with a standard error of 0.07, against a within estimate of 0.25 – it moves away from within by 0.20, about three times its own standard error, and precisely the correction the Nickell formula predicts for eight periods. Weak instruments could not do that; they would pull it toward within. And the first-stage F is 14, above the usual threshold. Two numbers agree there: Nickell-implied 0.45 and Arellano-Bond 0.46. Blundell-Bond lands nearby too, at 0.45, but its own Hansen test fails here as well, so I treat that as coincidence, not corroboration.

So what breaks it? The controls built on that same population count. GDP contributes through its per-capita denominator – it takes about a third of the fall – but the big movers are density and urban share, the same two that killed H2. Differencing ties them to the lagged error, and across the four specifications rho slides from 0.46 to 0.21 while its standard error grows from 0.07 to 0.10 and the Hansen test slides from 0.42 to 0.05, in step. That joint pattern is the evidence – not the single bracket crossing. So I read persistence where the moment conditions hold: about 0.46, roughly half of last period's growth carries over. The same defect explains both H2 and the persistence anomaly, which is why I keep coming back to that one piece of arithmetic.

Now the important part, and the part that changed my conclusions. Write BpCR out and you see it contains minus log population at t, plus log population at t minus one. Two of my controls – density and urban share – are built from that same population count. So each one inherits the sign of the date it is measured at: negative if contemporaneous, positive if lagged. I verify this directly: regressing BpCR on the two population terms gives minus 0.10 and plus 0.11, equal and opposite. So lagging by one period does not fix the problem, it just flips the bias. Estimating all four timing combinations confirms it: the negative sign follows whichever control is contemporaneous. The only clean timing is t minus two, which shares no term with BpCR – and there density is plus 0.011, still not negative. So H2, the compact-city hypothesis, is rejected: that famous negative sign is arithmetic, not economics.

What survives is migration. It is negative, highly significant, and essentially identical in every estimator, every timing, and all three development groups. And it is not definitional: decomposing BpCR shows urban population moves 0.88 points a year with migration while built-up area moves only 0.24 – a pattern consistent with cities absorbing migrants more than three and a half times faster than they build for them, and that gap is the densification. One honest caveat: I enter migration as strictly exogenous. It does pass a Wooldridge lead test, and the estimate barely moves if I treat it as predetermined – but I cannot rule out that migrants simply choose economies that are already building fast. So I report it as a conditional association, not a causal effect.

Income is the honest null. The sign is negative everywhere, but it is never significant once migration is measured over the same five-year window as the dependent variable. Part of what looked like an income effect was mismeasured migration.

On diagnostics: AR(2) is comfortably satisfied, which is the condition the lag structure needs. The Hansen test is borderline at 0.05, driven by the deepest lags; a shallower instrument set clears it and leaves every conclusion unchanged. I do not rest any claim on GMM alone – the fixed-effects columns, which need no instruments, agree.
:::





# Conclusion {.section-slide visibility="uncounted"}

## Conclusion


::: {style="font-size:1.8em; line-height:1.4;"}

- **Sprawl is real and measurable.** In most world regions urban land grows faster than population: a policy-relevant inefficiency.

    - BpCR > 0 (sprawl) in **63%** of country-period observations; **67% of countries (129 of 193)** sprawl on average over time [(Source: own calculations (GHSL/SMOD + UN data, 1985–2020))]{style="opacity:0.6; font-size:0.85em;"}
    - Median BpCR = **+0.0041** (built-up per capita rising); LCRPGR > 1 in **56%** of cases [(Source: own calculations (GHSL/SMOD + UN data, 1985–2020))]{style="opacity:0.6; font-size:0.85em;"}

- **The official LCRPGR is unreliable.** A raw ratio that explodes when population growth is near zero (division instability), producing sign-flipping outliers exactly where efficiency matters most.

- **The log-based BpCR is the better metric** (Lu & Weng 2026, not ours; the log-alternative case is also made by Nicolau et al. 2019). Symmetric and numerically stable, it is finite for every country-period. **Our part:** showing on **identical data** that it recovers an economic signal the official ratio leaves statistically undetectable.

- **What survives the diagnostics.** Sprawl is **path-dependent** (ρ ≈ 0.46 where the moments are clean); **net migration** is the one robust correlate. Density is *not* a compact-city effect (**H2 rejected**: the sign is arithmetic), and income cannot be distinguished from zero (**H3 not supported**).

- **We complement the literature, not contradict it.** Between countries, richer ones are more sprawled (the usual finding); within a single country the income sign turns negative, but it is not precisely estimated. What our country fixed effects *do* recover robustly is the **migration** channel: cities absorb in-migrants **more than 3.5× faster than they build for them**, so in-migration means densification, in developed, developing and least-developed economies alike.

- **Methodological contribution.** A dynamic two-way FE + **GMM** panel of **193 UN member countries (1985–2020)** that targets the Nickell bias and brackets persistence.

:::

::: {.notes}
"To conclude. Sprawl is real: in about two-thirds of countries, urban land grows faster than population over time.

But our central message is that the metric matters. The official LCRPGR is mathematically unreliable: on identical data it detects no driver at the 5% level, while BpCR recovers a precise one.

And with a reliable metric, one driver survives the diagnostics: net in-migration. Countries receiving migrants consume markedly less land per resident, by a similar amount in developed, developing and least-developed economies alike. Sprawl is also path-dependent: about half of one period's growth carries into the next.

I want to be equally clear about what did not survive. The compact-city result is arithmetic, not economics: BpCR contains urban population negatively at t and positively at t minus one, so any control built from that same population carries the sign of its own date. Dated cleanly, density is not negative at all. And income, which I expected to be the headline, carries a consistently negative sign but cannot be distinguished from zero once migration is measured over the window the dependent variable actually spans.

Even the migration result is an association, not an identified effect: I cannot rule out migrants selecting into economies that are already building fast.

The bottom line: a better yardstick gives better answers, and it also tells you which answers you never had. Thank you.

:::


## References {visibility="uncounted"}

::: {style="font-size:1.5em; line-height:1.35;"}

- **Lu, W. & Weng, Q. (2026).** From data inconsistency to reliable SDG 11.3.1 monitoring: A global multi-dataset assessment of land use efficiency. *Land Use Policy*, 168, 108088.
- **Seto, K. C., et al. (2011).** A meta-analysis of global urban land expansion. *PLOS ONE*, 6(8).
- **Oueslati, W., Alvanides, S. & Garrod, G. (2015).** Determinants of urban sprawl in European cities. *Urban Studies*, 52(9), 1594–1614.
- **Nicolau, R., David, J., Caetano, M. & Pereira, J. M. C. (2019).** Ratio of Land Consumption Rate to Population Growth Rate: analysis of different formulations applied to mainland Portugal. *ISPRS Int. J. Geo-Information*, 8(1), 10.
- **UN-Habitat. (2025).** *SDG Indicator 11.3.1 Metadata: Ratio of Land Consumption Rate to Population Growth Rate*. SDG indicator metadata, harmonized template v1.1.
- **Pesaresi, M., Schiavina, M., Politis, P., Freire, S., Krasnodębska, K., Uhl, J. H., … Kemper, T. (2024).** Advances on the Global Human Settlement Layer by joint assessment of Earth Observation and population survey data. *International Journal of Digital Earth*, 17(1). https://doi.org/10.1080/17538947.2024.2390454
- **GHS-BUILT-S (R2023A).** European Commission, Joint Research Centre (JRC). Product: GHS-BUILT-S; epoch 2020; 100 m; Mollweide. © European Union, 1995–2025: CC BY 4.0.
- **Schiavina, M., Melchiorri, M. & Pesaresi, M. (2023).** GHS-SMOD R2023A: GHS settlement layers, application of the Degree of Urbanisation methodology (stage I) to GHS-POP R2023A and GHS-BUILT-S R2023A, multitemporal (1975–2030). European Commission, JRC. doi:10.2905/A0DF7A6F-49DE-46EA-9BDE-563437A6E2BA
- **FAO (2024).** The Global Administrative Unit Layers (GAUL) 2024. Food and Agriculture Organization of the United Nations: CC BY 4.0.
- **UN Statistics Division (2025).** National Accounts Main Aggregates Database (SNAAMA): per-capita GDP, constant 2020 US$.
- **UN DESA, Population Division (2024).** World Population Prospects 2024: crude net-migration rate.
- **Worldometer (2024).** *World Population Clock: Current World Population*. https://www.worldometers.info/world-population/ (real-time projection based on UN DESA WPP 2024).


:::

::: {#refs style="display:none;"}
:::

# Thank you for your attention! {.section-slide visibility="uncounted"}

# Questions? {.section-slide visibility="uncounted"}

```{=html}
<script>
// Lazy-load the large interactive map iframes (kept out of the self-contained
// bundle via data-mapsrc, so they are not inlined as oversized data: URIs).
// Reveal loads each map when its slide becomes active, so Leaflet initialises at
// the correct visible size.
(function () {
  function nudge(f) {
    // make the embedded Leaflet/Plotly recompute its size once the slide is
    // actually visible (loading during a slide transition can size it to 0)
    try { f.contentWindow.dispatchEvent(new Event('resize')); } catch (e) {}
  }
  function loadMaps(scope) {
    (scope || document).querySelectorAll('iframe[data-mapsrc]').forEach(function (f) {
      if (!f.dataset.loaded) {
        f.dataset.loaded = '1';
        f.addEventListener('load', function () { nudge(f); setTimeout(function () { nudge(f); }, 300); });
        f.src = f.getAttribute('data-mapsrc');
      } else {
        nudge(f);
      }
    });
  }
  function wire() {
    if (window.Reveal && Reveal.on) {
      Reveal.on('ready', function (e) { loadMaps(e.currentSlide); });
      Reveal.on('slidechanged', function (e) {
        loadMaps(e.currentSlide);
        setTimeout(function () { loadMaps(e.currentSlide); }, 450);  // after the transition settles
      });
      // if Reveal already finished initialising before this ran, load the
      // current slide's map now (deep-links / fast loads miss the 'ready' event)
      if (Reveal.isReady && Reveal.isReady()) { loadMaps(Reveal.getCurrentSlide()); }
    } else { setTimeout(wire, 150); }
  }
  wire();
})();
</script>
```
