---
title: "Urban Land-Use Efficiency"
subtitle: "A Stable Metric for SDG 11.3.1 and Its Economic Drivers"
doc-type: "Seminar Paper"
course-info: "Development Economics · SoSe 2026"
lecturer: "Lecturer: Prof. Dr. Jürgen Bitzer"
author:
  - name: Ömer Furkan Çoban
    email: oemer.furkan.coban@uni-oldenburg.de
    affiliations:
      - name: Carl von Ossietzky University of Oldenburg
        department: Applied Economics and Data Science
        city: Oldenburg
        country: Germany
date: today
abstract: |
  The official SDG 11.3.1 indicator, the ratio of land consumption rate to
  population growth rate (LCRPGR), is unstable: it divides by a growth rate that
  approaches zero and mixes an arithmetic numerator with a logarithmic
  denominator. We propose the Built-up per Capita Rate (BpCR), the log change in
  built-up area per person, which equals the difference of two log growth rates
  and is therefore well-defined for every country-period. Using a harmonised
  global panel of 193 UN member states (1985–2020) built from GHSL built-up and
  population grids under the Degree of Urbanisation, we (i) document where LCRPGR
  breaks down, (ii) show BpCR reproduces the policy signal without the
  instability, and (iii) estimate its economic drivers with two-way fixed-effects
  and dynamic-panel GMM models.
keywords:
  - SDG 11.3.1
  - Urban land-use efficiency
  - Built-up per capita rate (BpCR)
  - LCRPGR
  - Dynamic panel GMM
  - GHSL / Degree of Urbanisation
journal:
  name: Development Economics seminar working paper
format:
  elsevier-pdf:
    keep-tex: true
    template-partials:
      - before-body.tex
    journal:
      formatting: preprint      # preprint | review | doubleblind
      model: 3p                 # 1p | 3p | 5p (3p = standard single-column-ish)
      cite-style: authoryear    # elsarticle author-year citations
bibliography: references.bib
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
#| include: false
library(here)
library(dplyr)
library(readr)
library(fixest)
panel <- read_csv(here("03_datasets/processed/reg_panel_urban.csv"),
    show_col_types = FALSE
)
gmm <- readRDS(here("03_datasets/processed/dynamic_gmm.rds"))
```

# Introduction

The world is in the midst of the largest urban expansion in history. Every single
year, the global built-up area grows by roughly the size of the Netherlands as
housing, roads, and industry spread outward [@seto2012global]. Urbanisation itself
is not the problem --- cities must grow to house more people --- but the policy
question is *how much* land each new resident consumes. If land grows at the same
pace as population, the city stays compact and infrastructure costs remain
manageable. If land grows substantially faster than population, the city sprawls,
and sprawl carries measurable costs: higher per-capita expenditure on roads,
water, and transit; the systematic displacement of poorer residents to the
periphery; permanent farmland loss; and locking cities into car-dependent,
high-emission growth paths [@oueslati2015sprawl].

To track this, the United Nations adopted Sustainable Development Goal (SDG)
target 11.3, which asks countries to "enhance inclusive and sustainable
urbanisation" by keeping land consumption in line with population growth.
Indicator 11.3.1 operationalises this target as the Ratio of Land Consumption
Rate to Population Growth Rate (LCRPGR), a single metric intended to signal
whether cities are densifying, balancing, or sprawling [@unhabitat2025metadata].

Yet the official indicator carries a structural flaw: it divides an arithmetic
(linear) land growth rate by a logarithmic (exponential) population growth rate,
creating a ratio that is dimensionally inconsistent and that explodes to
$\pm\infty$ whenever population growth approaches zero --- a common occurrence in
shrinking or demographically stable cities [@nicolau2019lcrpgr]. A city with
identical 5\% growth in both built-up area and population produces LCRPGR = 1.13,
not 1.0, because $(g-1) > \ln(g)$. The same 2\% land expansion reads as 1.33,
$-20$, or $2\,000$ depending only on how close the denominator is to zero. The
UN metadata itself acknowledges that LCRPGR is "difficult to interpret" and
proposes secondary indicators built from the same inputs, including built-up area
per capita [@unhabitat2025metadata, §4.f].

The literature on urban land expansion has developed along two largely separate
tracks. On the measurement side, @nicolau2019lcrpgr document the instability of
the official ratio and recommend a log-based formulation; @Lu2026 demonstrate
that land-use efficiency estimates swing widely with dataset choice and argue for
a log-based framework for reliable monitoring. On the driver side, @seto2011meta
show in a meta-analysis of 326 remote-sensing studies that urban land grew faster
than population almost everywhere, with income as the dominant driver across most
regions; @oueslati2015sprawl find that rising income and population both expand
the urban footprint in European cities, and that cities continue to spread even
where population is flat.

We bridge these two strands. Following the log-based approach recommended by
@nicolau2019lcrpgr and @Lu2026, we construct the Built-up per Capita Rate
(BpCR) --- the annual log growth rate of built-up area per capita, defined as the
difference of two logarithmic rates:

$$ \text{BpCR} = \frac{1}{z}\ln\!\left(\frac{V_t / P_t}{V_{t-z} / P_{t-z}}\right)
= \text{LCR}_{\log} - \text{PGR}_{\log}. $$

Unlike LCRPGR, BpCR places numerator and denominator on the same mathematical
scale, is well-defined for every country-period including those with stable or
shrinking populations, and retains an intuitive interpretation: positive values
signal sprawl (per-capita land use rising), negative values signal densification
(per-capita land use falling).

We then make the metric *dynamic*. Using a harmonised global panel of 193 UN
member states spanning eight five-year periods (1985--2020) --- built from GHSL
built-up and population grids under the Degree of Urbanisation --- we estimate a
two-way fixed-effects model with a lagged dependent variable to capture
path-dependence in per-capita land use, and we correct the resulting Nickell bias
with Arellano-Bond and Blundell-Bond dynamic-panel GMM. This allows us to answer
three questions: (i) does the choice of metric change the empirical conclusions?
(ii) what economic and structural factors drive within-country changes in
per-capita urban land use? and (iii) how persistent are these trajectories?

We find that the metric matters decisively. The official LCRPGR produces a
within-country $R^2$ of approximately 0.003, with no variable reaching
statistical significance. The same specification estimated with BpCR yields a
within-country $R^2$ roughly twenty times higher, and clear, significant drivers
emerge. Our GMM-corrected estimates reveal strong path-dependence ($\rho \approx
0.6$): roughly 60\% of the previous period's per-capita land-use change persists
into the next. Within countries, higher urban density, net in-migration, and
higher GDP per capita all predict *less* per-capita land consumption --- i.e.,
richer countries densify, not sprawl, when we look at the within-country margin
that cross-sectional studies miss.

The remainder of the paper is structured as follows. Section 2 formally defines the
SDG 11.3.1 indicator and its limitations. Section 3 introduces BpCR and
demonstrates its properties. Section 4 describes the data. Section 5 outlines the
empirical strategy. Section 6 presents the results, and Section 7 concludes.

# The SDG 11.3.1 indicator and its limitations

<!-- TODO: define LCR, PGR, LCRPGR exactly as in the metadata; show the two
problems: (1) division by PGR near zero, (2) arithmetic-vs-log asymmetry.
Reference Nicolau et al. on alternative formulations. -->
$$
\text{LCRPGR} = \frac{\text{LCR}}{\text{PGR}},\qquad
\text{LCR}=\frac{V_t-V_{t-z}}{V_{t-z}}\cdot\frac{1}{z},\qquad
\text{PGR}=\frac{\ln(P_t/P_{t-z})}{z}.
$$
The arithmetic LCR over a logarithmic PGR makes the ratio asymmetric, and PGR in
the denominator makes it diverge when population is roughly stable
[@nicolau2019lcrpgr].

# A stable metric: the Built-up per Capita Rate (BpCR)

<!-- TODO: define BpCR = log growth of built-up per capita = LCR_log - PGR_log;
show it is finite, symmetric, and additively decomposable. -->
$$
\text{BpCR}_{i,t}=\frac{1}{z}\ln\!\left(\frac{V_t/P_t}{V_{t-z}/P_{t-z}}\right)
= \text{LCR}_{\log}-\text{PGR}_{\log}.
$$

# Data

<!-- TODO: GHSL built-up (GHS-BUILT-S) and population (GHS-POP) under DEGURBA
(GHS-SMOD >= 21); GAUL 2024 admin units; UN GDP per capita and net migration.
Describe the 193-country, 1985-2020 panel. -->
See @pesaresi2024ghsl, @schiavina2023smod, @unsd2025snaama and @undesa2024wpp for
the source datasets.

```{r sample-size}
#| label: tbl-sample
#| tbl-cap: "Estimation sample"
n_ctry <- length(unique(panel$iso3))
n_obs <- sum(!is.na(panel$bpcr))
data.frame(
    Quantity = c("Countries", "Country-period observations"),
    Value = c(n_ctry, n_obs)
) |>
    knitr::kable()
```

# Empirical strategy

<!-- TODO: one specification, two dependent variables (LCRPGR vs BpCR); two-way
fixed effects; then the dynamic version with the lagged DV and its Nickell-bias
correction via Arellano-Bond / Blundell-Bond GMM. -->
$$
\text{Metric}_{i,t}=\rho\,\text{Metric}_{i,t-1}
 + \beta_1\ln(\text{UrbDens})_{i,t} + \beta_2\,\text{UrbShare}_{i,t}
 + \beta_3\ln(\text{GDPpc})_{i,t} + \beta_4\,\text{NetMigr}_{i,t}
 + \mu_i + \delta_t + \varepsilon_{i,t}.
$$

# Results

<!-- TODO: static comparison (metric matters), then the final dynamic model.
The GMM table is produced by 02_scripts/02_analysis/04_dynamic_gmm.R and stored
in dynamic_gmm.rds (object `main_combo`). -->



# Robustness

<!-- TODO: 2000-2020 subsample, alternative instrument depth, winsorisation. -->

# Discussion

<!-- TODO: reconcile the within-country income->densification finding with the
cross-sectional income->sprawl literature; the measurement-source sensitivity
critique. -->

# Conclusion

<!-- TODO: BpCR as a drop-in, stable replacement that complements rather than
contradicts the literature; methodological contribution. -->

# References {.unnumbered}

::: {#refs}
:::
