---
title: "Urban Land-Use Efficiency"
subtitle: "A Stable Metric for SDG 11.3.1 and Its Economic Drivers"
doc-type: "Seminar Paper"
course-info: "Development Economics · SoSe 2026"
lecturer: "Lecturer: Prof. Dr. Jürgen Bitzer"
author:
  - name: Ömer Furkan Çoban
    email: oemer.furkan.coban@uni-oldenburg.de
    affiliations:
      - name: Carl von Ossietzky University of Oldenburg
        department: Applied Economics and Data Science
        city: Oldenburg
        country: Germany
date: today
abstract: |
  The official SDG 11.3.1 indicator, the ratio of land consumption rate to
  population growth rate (LCRPGR), is unstable: it divides by a growth rate that
  approaches zero and mixes an arithmetic numerator with a logarithmic
  denominator. We propose the Built-up per Capita Rate (BpCR), the log change in
  built-up area per person, which equals the difference of two log growth rates
  and is therefore well-defined for every country-period. Using a harmonised
  global panel of 193 UN member states (1985–2020) built from GHSL built-up and
  population grids under the Degree of Urbanisation, we (i) document where LCRPGR
  breaks down, (ii) show BpCR reproduces the policy signal without the
  instability, and (iii) estimate its economic drivers with two-way fixed-effects
  and dynamic-panel GMM models.
keywords:
  - SDG 11.3.1
  - Urban land-use efficiency
  - Built-up per capita rate (BpCR)
  - LCRPGR
  - Dynamic panel GMM
  - GHSL / Degree of Urbanisation
journal:
  name: Development Economics seminar working paper
format:
  elsevier-pdf:
    keep-tex: true
    template-partials:
      - before-body.tex
    include-in-header:
      text: |
        \usepackage{makecell}
    journal:
      formatting: preprint      # preprint | review | doubleblind
      model: 3p                 # 1p | 3p | 5p (3p = standard single-column-ish)
      cite-style: authoryear    # elsarticle author-year citations
bibliography: references.bib
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
#| include: false
library(here)
library(dplyr)
library(readr)
library(fixest)
panel <- read_csv(here("03_datasets/processed/reg_panel_urban.csv"),
    show_col_types = FALSE
)
gmm <- readRDS(here("03_datasets/processed/dynamic_gmm.rds"))
```


# Introduction

The world is urbanising rapidly: by the middle of this century close to
two-thirds of humanity will live in cities, and almost all of the projected
growth in built-up land will occur in developing countries
[@gao2020]. The scale of expansion is striking: every year the world's
built-up area grows by roughly the size of the Netherlands [@seto2012global].
For development policy the central question is not whether cities expand, but how
efficiently they do so, that is, how much additional land each new urban resident
consumes.
When built-up area grows in step with population, cities densify and remain
compact; when it grows much faster than population, they sprawl. Sprawl carries real costs. Once farmland on the urban fringe is
built over, it is effectively lost for good. Low-density, spread-out development
makes people more dependent on cars, because homes, jobs, and services lie far
apart. It also makes public infrastructure more expensive per resident, since the
same roads, pipes, and power and transport networks must reach a smaller number
of people over a larger area. And by encouraging longer car journeys and
land-hungry construction, sprawl tends to raise greenhouse-gas emissions per
person. Measuring land-use efficiency well is therefore a precondition for
monitoring sustainable urbanisation.

The United Nations monitors this through target 11.3 of its Sustainable
Development Goals (SDGs), whose official indicator 11.3.1 is the ratio of the Land
Consumption Rate to the Population Growth Rate (LCRPGR) [@unhabitat2025metadata].
Because LCRPGR is the agreed global yardstick, its statistical properties are not
a technical detail: they determine how every country's progress is judged and
compared. Yet the indicator is fragile by construction. It divides an arithmetic
land-consumption rate by a logarithmic population-growth rate, so it is
internally inconsistent; and because the population-growth rate sits in the
denominator, the ratio diverges towards plus or minus infinity and changes sign
whenever population is roughly stable. In exactly the countries where land-use
efficiency matters most for policy, the official metric becomes uninformative.

This paper makes three contributions. First, it documents, on a harmonised global
dataset, that LCRPGR is statistically unstable and that its instability is
concentrated where population growth is near zero. Second, it proposes a simple,
log-based alternative, the Built-up per Capita Rate (BpCR), defined as the
logarithmic growth rate of built-up area per capita. BpCR is the difference of
two logarithmic growth rates, which makes it finite for every country-period,
symmetric between densification and sprawl, additively decomposable, and directly
interpretable; it reproduces the policy signal of LCRPGR where the latter is
well-behaved, and so complements rather than replaces the official indicator.
Third, using the stable metric, the paper estimates the economic and structural
drivers of per-capita urban land use on a panel of 193 United Nations member
states observed at five-year intervals between 1985 and 2020.

The empirical analysis builds the panel from satellite-based settlement data
under the global Degree of Urbanisation standard, so that built-up area and
population are measured consistently. To address the well-known downward bias of
the lagged dependent variable in dynamic fixed-effects models, the persistence of
per-capita land use is estimated with Arellano-Bond and Blundell-Bond generalised
method of moments (GMM) estimators. Three results stand out. Sprawl is the global
norm rather than the exception: built-up area per capita rises in a clear
majority of country-periods. Per-capita land use is strongly path-dependent, with
about three-fifths of one period's value carried into the next. And within
countries, denser cities, net in-migration, and higher income are all associated
with less land consumed per resident, indicating that, over time, economic growth
is accompanied by densification rather than sprawl.

The remainder of the paper is organised as follows. Section 2 reviews the
indicator and the related literature. Section 3 sets out the structural
limitations of LCRPGR, and Section 4 introduces BpCR. Section 5 describes the
data and Section 6 presents descriptive evidence. Section 7 details the empirical
strategy and Section 8 reports the results, including robustness checks. Section 9
discusses the findings and Section 10 concludes.

# The SDG 11.3.1 indicator and its limitations

Following the UN-Habitat metadata, the indicator divides an arithmetic land
consumption rate by a logarithmic population growth rate:
$$
\text{LCRPGR} = \frac{\text{LCR}}{\text{PGR}},\qquad
\text{LCR}=\frac{V_t-V_{t-z}}{V_{t-z}}\cdot\frac{1}{z},\qquad
\text{PGR}=\frac{\ln(P_t/P_{t-z})}{z},
$$
where $V$ is built-up area, $P$ population, and $z$ the period length. Two
structural problems follow. **(1) Division by a near-zero denominator.** When
population is roughly stable ($\text{PGR}\approx 0$) the ratio diverges towards
$\pm\infty$ and flips sign, so small, ordinary demographic differences produce
wild, uninterpretable values. **(2) Arithmetic-versus-logarithmic asymmetry.**
The numerator is an arithmetic growth rate while the denominator is logarithmic,
so the ratio is not symmetric and does not aggregate cleanly over time
[@nicolau2019lcrpgr]. The result is a metric whose tails are dominated by an
arithmetic artefact rather than by real land-use behaviour.

# A stable metric: the Built-up per Capita Rate (BpCR)

The UN-Habitat metadata itself lists built-up area per capita ($V/P$) as a
secondary indicator. We build on it and measure its **log growth**:
$$
\text{BpCR}_{i,t}=\frac{1}{z}\ln\!\left(\frac{V_t/P_t}{V_{t-z}/P_{t-z}}\right)
= \text{LCR}_{\log}-\text{PGR}_{\log}.
$$
Because BpCR is the **difference of two logarithmic growth rates**, it is
(i) finite and well-defined for every country-period (no division by ~0),
(ii) symmetric (densification and sprawl are mirror images around zero),
(iii) additively decomposable, and (iv) directly interpretable: $\text{BpCR}>0$
means sprawl (more land per resident), $\text{BpCR}<0$ densification, and
$\text{BpCR}=0$ land growing exactly in step with population. BpCR reproduces the
policy signal of LCRPGR where the latter is well-behaved, but without the
instability; it complements rather than contradicts the official indicator.

# Related literature

The empirical literature on urban land establishes two facts we build on.
**Drivers.** Global remote-sensing meta-analyses find that built-up area has
grown faster than urban population almost everywhere, with income a key driver
across countries [@seto2011meta]; panel studies of European cities likewise tie
sprawl to rising income and population [@oueslati2015sprawl]. **Measurement.**
Work on the indicator itself shows the LCRPGR formula is structurally fragile and
that log/efficiency-based alternatives are more robust [@nicolau2019lcrpgr], and
that land-use-efficiency estimates swing widely with the choice of population
dataset, calling for a more reliable framework [@Lu2026]. Our contribution sits
at the intersection: a **stable metric (BpCR)** estimated **dynamically** (two-way
fixed effects + GMM) on a **global panel of 193 UN member states**, measuring the
persistence of per-capita land use.

# Data

The analysis uses a single, **urban-scale** panel built under the EU/UN **Degree
of Urbanisation** (urban $=$ `GHS-SMOD` $\ge 21$), so built-up and population are
measured consistently from the same grids. Built-up area comes from
**GHS-BUILT-S** and population from **GHS-POP** (GHSL R2023A) [@pesaresi2024ghsl;
@schiavina2023smod], aggregated to national totals over **GAUL 2024**
administrative units. Economic controls are GDP per capita (constant 2020 US\$,
UN SNAAMA) and net migration (UN DESA WPP 2024) [@unsd2025snaama; @undesa2024wpp];
region and development group come from the UN M49 classification. The panel
covers **193 UN member states** at 5-year epochs, **1985–2020**.

```{r sample-size}
#| label: tbl-sample
#| tbl-cap: "Estimation sample"
n_ctry <- length(unique(panel$iso3))
n_obs <- sum(!is.na(panel$bpcr))
data.frame(
    Quantity = c("Countries", "Country-period observations"),
    Value = c(n_ctry, n_obs)
) |>
    knitr::kable()
```

# Descriptive statistics

Across 1985–2020 a clear majority of countries record $\text{BpCR}>0$: built-up
area is expanding faster than urban population for most of the world, i.e. sprawl
is the global norm rather than the exception. The contrast with LCRPGR is
instructive -- LCRPGR's distribution is dominated by extreme tail values
concentrated exactly where population growth is near zero, whereas BpCR is
well-behaved across the whole sample. Regional trajectories show convergence
towards balance in some fast-urbanising economies while others continue to
sprawl.

<!-- TODO: insert global BpCR distribution figure and the regional-trend figure
from 04_outputs/figures/ once finalised. -->

# Empirical strategy

**Research question.** What are the economic drivers of per-capita urban land use,
and how persistent is it over time? We test five hypotheses (@tbl-hyp).

```{r tbl-hyp}
#| label: tbl-hyp
#| tbl-cap: "Hypotheses"
data.frame(
    ` ` = c("H1", "H2", "H3", "H4", "H5"),
    Hypothesis = c(
        "BpCR captures land-use efficiency more reliably than LCRPGR",
        "Higher urban density is associated with less land per capita",
        "Higher income is associated with less land per capita",
        "Net in-migration densifies cities",
        "Per-capita land use is path-dependent"
    ),
    Expected = c("better fit", "$-$", "$-$", "$-$", "$+$"),
    check.names = FALSE
) |>
    knitr::kable(
        format = "latex", booktabs = TRUE, escape = FALSE, linesep = "",
        align = c("c", "l", "c")
    ) |>
    kableExtra::kable_styling(latex_options = "HOLD_position")
```

We run **one specification** with each metric
as the dependent variable, so the comparison is like-for-like:
$$
\text{Metric}_{i,t}=\rho\,\text{Metric}_{i,t-1}
 + \beta_1\ln(\text{UrbDens})_{i,t} + \beta_2\,\text{UrbShare}_{i,t}
 + \beta_3\ln(\text{GDPpc})_{i,t} + \beta_4\,\text{NetMigr}_{i,t}
 + \mu_i + \delta_t + \varepsilon_{i,t},
$$
with country fixed effects $\mu_i$, period fixed effects $\delta_t$, and standard
errors clustered by country. We proceed in two stages. **(1) Static** (dropping
the lagged dependent variable) to isolate the role of the *metric* -- LCRPGR versus
BpCR. **(2) Dynamic** (adding the lagged DV) to capture path-dependence. Because
the lagged-DV coefficient is **downward-biased under fixed effects (Nickell
bias)**, we correct it with **Arellano-Bond difference GMM** and **Blundell-Bond
system GMM**, instrumenting with deeper lags. Instrument validity is assessed with
the **Sargan/Hansen** over-identification test and the **AR(2)** serial-correlation
test (both $p>0.10$ indicating validity).

# Results

**The metric matters.** On identical data, the static LCRPGR specification is
contaminated by its unstable tails, while the BpCR specification is well-behaved;
the two diverge precisely where $\text{PGR}$ is small, confirming that LCRPGR's
extremes are an arithmetic artefact rather than a behavioural signal.

**The final dynamic model.** @tbl-gmm reports the preferred specification: a
fixed-effects stepwise build-up alongside the Arellano-Bond and Blundell-Bond GMM
estimates.

```{r gmm-table}
#| label: tbl-gmm
#| tbl-cap: "Final dynamic-panel model: FE stepwise build-up, Arellano-Bond (difference) and Blundell-Bond (system) GMM. Coefficient with significance stars on top, clustered/robust standard error in parentheses below. *** p<0.01, ** p<0.05, * p<0.1."
if (!requireNamespace("kableExtra", quietly = TRUE)) install.packages("kableExtra")
mc <- as.data.frame(gmm$main_combo, stringsAsFactors = FALSE)
# coefficient cells look like "X***<br>(se)" -> stacked \makecell{X*** \\ (se)}
to_cell <- function(x) {
    ifelse(
        grepl("<br>", x, fixed = TRUE),
        paste0("\\makecell[c]{", sub("<br>", " \\\\ ", x, fixed = TRUE), "}"),
        x
    )
}
for (j in 2:ncol(mc)) mc[[j]] <- to_cell(mc[[j]])
names(mc)[1] <- ""
esc <- function(x) gsub("%", "\\%", x, fixed = TRUE) # literal % -> \% (LaTeX)
names(mc) <- esc(names(mc))
mc[[1]] <- esc(mc[[1]]) # escape % in row labels too
knitr::kable(mc,
    format = "latex", booktabs = TRUE, escape = FALSE, linesep = "",
    align = c("l", rep("c", ncol(mc) - 1))
) |>
    kableExtra::kable_styling(latex_options = c("HOLD_position", "scale_down"))
```

In the Arellano-Bond specification the lagged term is large and significant
($\rho\approx 0.61$): roughly 60% of last period's BpCR persists, and the GMM
$\rho$ is about three times the FE estimate, exactly as the Nickell bias predicts,
confirming **strong path-dependence** in per-capita land use. Among the drivers,
**urban density** enters negative (a compact-city effect: denser cities consume
less new land per resident), **net migration** negative (in-migrants pack into the
existing stock faster than built-up grows), and **GDP per capita** negative --
within countries, getting richer is associated with **densification, not sprawl**.
The level of urbanisation (urban population share) is insignificant once the rest
is controlled, suggesting that urban **form**, not stage, drives land use.

# Robustness

The findings are stable across a 2000–2020 subsample, a restriction to
shrinking-population countries, and alternative instrument depths; the Sargan and
AR(2) diagnostics pass throughout, supporting instrument validity.

<!-- TODO: tabulate the 2000-2020 and shrinking-population robustness results. -->

# Discussion

The within-country result that income is associated with densification may appear
to contradict the cross-sectional literature in which richer countries have
historically sprawled. The two are complementary: *across* countries, higher
income has coincided with more sprawl historically, but *within* a country, rising
income over time is associated with denser, not more sprawling, urban land use.
Measuring this distinction requires a metric that is stable within country-period --
which BpCR provides and LCRPGR does not.

# Conclusion

The official SDG 11.3.1 indicator is statistically unstable by construction. The
Built-up per Capita Rate is a drop-in, log-based alternative that is finite,
symmetric, and interpretable for every country-period, and it reproduces the
policy signal without the instability. Estimated dynamically on a global panel,
per-capita urban land use is strongly path-dependent and, within countries,
declines with density, in-migration, and income. The contribution is as much
**methodological** -- a stable metric and a bias-corrected dynamic estimator -- as
substantive, and it complements rather than overturns the existing literature.

# References {.unnumbered}

::: {#refs}
:::
