---
title: "Urban Land-Use Efficiency: A Stable Metric for SDG 11.3.1 and Its Economic Drivers"
author: "Ömer Furkan Çoban"
date: today
abstract: |
  The official SDG 11.3.1 indicator, the ratio of land consumption rate to
  population growth rate (LCRPGR), is unstable: it divides by a growth rate that
  approaches zero and mixes an arithmetic numerator with a logarithmic
  denominator. We propose the Built-up per Capita Rate (BpCR), the log change in
  built-up area per person, which equals the difference of two log growth rates
  and is therefore well-defined for every country-period. Using a harmonised
  global panel of 193 UN member states (1985–2020) built from GHSL built-up and
  population grids under the Degree of Urbanisation, we (i) document where LCRPGR
  breaks down, (ii) show BpCR reproduces the policy signal without the
  instability, and (iii) estimate its economic drivers with two-way fixed-effects
  and dynamic-panel GMM models.
format:
  pdf:
    documentclass: article
    geometry: margin=1in
    fontsize: 11pt
    number-sections: true
    toc: false
    keep-tex: false
bibliography: references.bib
csl: https://www.zotero.org/styles/apa
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
#| include: false
library(here)
library(dplyr)
library(readr)
library(fixest)
panel <- read_csv(here("03_datasets/processed/reg_panel_urban.csv"),
                  show_col_types = FALSE)
gmm   <- readRDS(here("03_datasets/processed/dynamic_gmm.rds"))
```

# Introduction

<!-- TODO: motivate global urban land monitoring; state the gap (LCRPGR
instability) and the contribution (BpCR + drivers). ~1 page. -->
Cities must grow to house more people; the policy question is how much land each
new resident consumes [@seto2011meta; @oueslati2015sprawl]. SDG target 11.3 asks
countries to keep land consumption in line with population growth, and indicator
11.3.1 operationalises this as the LCRPGR ratio [@unhabitat2025metadata].

# The SDG 11.3.1 indicator and its limitations

<!-- TODO: define LCR, PGR, LCRPGR exactly as in the metadata; show the two
problems: (1) division by PGR near zero, (2) arithmetic-vs-log asymmetry.
Reference Nicolau et al. on alternative formulations. -->
$$
\text{LCRPGR} = \frac{\text{LCR}}{\text{PGR}},\qquad
\text{LCR}=\frac{V_t-V_{t-z}}{V_{t-z}}\cdot\frac{1}{z},\qquad
\text{PGR}=\frac{\ln(P_t/P_{t-z})}{z}.
$$
The arithmetic LCR over a logarithmic PGR makes the ratio asymmetric, and PGR in
the denominator makes it diverge when population is roughly stable
[@nicolau2019lcrpgr].

# A stable metric: the Built-up per Capita Rate (BpCR)

<!-- TODO: define BpCR = log growth of built-up per capita = LCR_log - PGR_log;
show it is finite, symmetric, and additively decomposable. -->
$$
\text{BpCR}_{i,t}=\frac{1}{z}\ln\!\left(\frac{V_t/P_t}{V_{t-z}/P_{t-z}}\right)
= \text{LCR}_{\log}-\text{PGR}_{\log}.
$$

# Data

<!-- TODO: GHSL built-up (GHS-BUILT-S) and population (GHS-POP) under DEGURBA
(GHS-SMOD >= 21); GAUL 2024 admin units; UN GDP per capita and net migration;
World Bank income groups; WGI. Describe the 193-country, 1985-2020 panel. -->
See @pesaresi2024ghsl, @schiavina2023smod, @unsd2025snaama and @undesa2024wpp for
the source datasets.

```{r sample-size}
#| label: tbl-sample
#| tbl-cap: "Estimation sample"
n_ctry <- length(unique(panel$iso3))
n_obs  <- sum(!is.na(panel$bpcr))
data.frame(Quantity = c("Countries", "Country-period observations"),
           Value    = c(n_ctry, n_obs)) |>
  knitr::kable()
```

# Empirical strategy

<!-- TODO: one specification, two dependent variables (LCRPGR vs BpCR); two-way
fixed effects; then the dynamic version with the lagged DV and its Nickell-bias
correction via Arellano-Bond / Blundell-Bond GMM. -->
$$
\text{Metric}_{i,t}=\rho\,\text{Metric}_{i,t-1}
 + \beta_1\ln(\text{UrbDens})_{i,t} + \beta_2\,\text{UrbShare}_{i,t}
 + \beta_3\ln(\text{GDPpc})_{i,t} + \beta_4\,\text{NetMigr}_{i,t}
 + \mu_i + \delta_t + \varepsilon_{i,t}.
$$

# Results

<!-- TODO: static comparison (metric matters), then the final dynamic model.
The GMM table is produced by 02_scripts/02_analysis/04_dynamic_gmm.R and stored
in dynamic_gmm.rds (object `main_combo`). -->



# Robustness

<!-- TODO: 2000-2020 subsample, alternative instrument depth, winsorisation;
results from 02_scripts/02_analysis/09_robustness_checks.R. -->

# Discussion

<!-- TODO: reconcile the within-country income->densification finding with the
cross-sectional income->sprawl literature; the measurement-source sensitivity
critique. -->

# Conclusion

<!-- TODO: BpCR as a drop-in, stable replacement that complements rather than
contradicts the literature; methodological contribution. -->

# References {.unnumbered}

::: {#refs}
:::
