---
title: "Urban Land-Use Efficiency"
subtitle: "A Stable Metric for SDG 11.3.1 and Its Economic Drivers"
doc-type: "Seminar Paper"
course-info: "Development Economics · SoSe 2026"
lecturer: "Lecturer: Prof. Dr. Jürgen Bitzer"
author:
  - name: Ömer Furkan Çoban
    email: oemer.furkan.coban@uni-oldenburg.de
    affiliations:
      - name: Carl von Ossietzky University of Oldenburg
        department: Applied Economics and Data Science
        city: Oldenburg
        country: Germany
date: today
abstract: |
  The official SDG 11.3.1 indicator, the ratio of land consumption rate to
  population growth rate (LCRPGR), is unstable: it divides by a growth rate that
  approaches zero and mixes an arithmetic numerator with a logarithmic
  denominator. We propose the Built-up per Capita Rate (BpCR), the log change in
  built-up area per person, which equals the difference of two log growth rates
  and is therefore well-defined for every country-period. Using a harmonised
  global panel of 193 UN member states (1985–2020) built from GHSL built-up and
  population grids under the Degree of Urbanisation, we (i) document where LCRPGR
  breaks down, (ii) show BpCR reproduces the policy signal without the
  instability, and (iii) estimate its economic drivers with two-way fixed-effects
  and dynamic-panel GMM models.
keywords:
  - SDG 11.3.1
  - Urban land-use efficiency
  - Built-up per capita rate (BpCR)
  - LCRPGR
  - Dynamic panel GMM
  - GHSL / Degree of Urbanisation
journal:
  name: Development Economics seminar working paper
format:
  elsevier-pdf:
    keep-tex: true
    template-partials:
      - before-body.tex
    include-in-header:
      text: |
        \usepackage{makecell}
    journal:
      formatting: preprint      # preprint | review | doubleblind
      model: 3p                 # 1p | 3p | 5p (3p = standard single-column-ish)
      cite-style: authoryear    # elsarticle author-year citations
bibliography: references.bib
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
#| include: false
library(here)
library(dplyr)
library(readr)
library(fixest)
panel <- read_csv(here("03_datasets/processed/reg_panel_urban.csv"),
    show_col_types = FALSE
)
gmm <- readRDS(here("03_datasets/processed/dynamic_gmm.rds"))
```


# Introduction

The world is urbanising rapidly: by the middle of this century close to
two-thirds of humanity will live in cities, and almost all of the projected
growth in built-up land will occur in developing countries
[@gao2020]. The scale of expansion is striking: every year the world's
built-up area grows by roughly the size of the Netherlands [@seto2012global].
For development policy the central question is not whether cities expand, but how
efficiently they do so, that is, how much additional land each new urban resident
consumes.
When built-up area grows in step with population, cities densify and remain
compact; when it grows much faster than population, they sprawl. Sprawl carries real costs. Once farmland on the urban fringe is
built over, it is effectively lost for good. Low-density, spread-out development
makes people more dependent on cars, because homes, jobs, and services lie far
apart. It also makes public infrastructure more expensive per resident, since the
same roads, pipes, and power and transport networks must reach a smaller number
of people over a larger area. And by encouraging longer car journeys and
land-hungry construction, sprawl tends to raise greenhouse-gas emissions per
person. Measuring land-use efficiency well is therefore a precondition for
monitoring sustainable urbanisation.

The United Nations monitors this through target 11.3 of its Sustainable
Development Goals (SDGs), whose official indicator 11.3.1 is the ratio of the Land
Consumption Rate to the Population Growth Rate (LCRPGR) [@unhabitat2025metadata].
Because LCRPGR is the agreed global yardstick, its statistical properties are not
a technical detail: they determine how every country's progress is judged and
compared. Yet the indicator is fragile by construction. It divides an arithmetic
land-consumption rate by a logarithmic population-growth rate, so it is
internally inconsistent; and because the population-growth rate sits in the
denominator, the ratio diverges towards plus or minus infinity and changes sign
whenever population is roughly stable. In exactly the countries where land-use
efficiency matters most for policy, the official metric becomes uninformative.

This paper makes three contributions. First, it documents, on a harmonised global
dataset, that LCRPGR is statistically unstable and that its instability is
concentrated where population growth is near zero. Second, it proposes a simple,
log-based alternative, the Built-up per Capita Rate (BpCR), defined as the
logarithmic growth rate of built-up area per capita. BpCR is the difference of
two logarithmic growth rates, which makes it finite for every country-period,
symmetric between densification and sprawl, additively decomposable, and directly
interpretable; it reproduces the policy signal of LCRPGR where the latter is
well-behaved, and so complements rather than replaces the official indicator.
Third, using the stable metric, the paper estimates the economic and structural
drivers of per-capita urban land use on a panel of 193 United Nations member
states observed at five-year intervals between 1985 and 2020.

The empirical analysis builds the panel from satellite-based settlement data
under the global Degree of Urbanisation standard, so that built-up area and
population are measured consistently. To address the well-known downward bias of
the lagged dependent variable in dynamic fixed-effects models, the persistence of
per-capita land use is estimated with Arellano-Bond and Blundell-Bond generalised
method of moments (GMM) estimators. Three results stand out. Sprawl is the global
norm rather than the exception: built-up area per capita rises in a clear
majority of country-periods. Per-capita land use is strongly path-dependent, with
about three-fifths of one period's value carried into the next. And within
countries, denser cities, net in-migration, and higher income are all associated
with less land consumed per resident, indicating that, over time, economic growth
is accompanied by densification rather than sprawl.

The remainder of the paper is organised as follows. Section 2 reviews the
indicator and the related literature. Section 3 sets out the structural
limitations of LCRPGR, and Section 4 introduces BpCR. Section 5 describes the
data and Section 6 presents descriptive evidence. Section 7 details the empirical
strategy and Section 8 reports the results, including robustness checks. Section 9
discusses the findings and Section 10 concludes.

# Measurement Framework

This section sets out the conceptual core of the paper. We first define the
official SDG 11.3.1 indicator and show why it is statistically unstable, and we
then introduce the stable, log-based alternative used in the empirical analysis.

## The SDG 11.3.1 Indicator and Its Limitations

Following the UN-Habitat metadata [@unhabitat2025metadata], the official
indicator divides an arithmetic land consumption rate by a logarithmic
population growth rate:
$$
\text{LCRPGR} = \frac{\text{LCR}}{\text{PGR}},\qquad
\text{LCR}=\frac{V_t-V_{t-z}}{V_{t-z}}\cdot\frac{1}{z},\qquad
\text{PGR}=\frac{\ln(P_t/P_{t-z})}{z},
$$
where the symbols are defined as follows: $V_t$ is the urban built-up area (in
square kilometres) at the end of the period and $V_{t-z}$ the built-up area at the
start; $P_t$ and $P_{t-z}$ are the corresponding urban populations; $z$ is the
length of the period in years; $\ln(\cdot)$ denotes the natural logarithm; and the
subscript $i$ indexes the country. The Land Consumption Rate $(\text{LCR})$ is thus
the arithmetic growth of built-up area, expressed per year, and the Population
Growth Rate $(\text{PGR})$ is the logarithmic (continuous) growth of population per
year; the indicator $\text{LCRPGR}$ is their ratio. Two structural problems
follow. **(1) Division by a near-zero denominator.** When
population is roughly stable ($\text{PGR}\approx 0$) the ratio diverges towards
$\pm\infty$ and flips sign, so small, ordinary demographic differences produce
wild, uninterpretable values. **(2) Arithmetic-versus-logarithmic asymmetry.**
The numerator is an arithmetic growth rate while the denominator is logarithmic,
so the ratio is not symmetric and does not aggregate cleanly over time
[@nicolau2019lcrpgr]. The result is a metric whose tails are dominated by an
arithmetic artefact rather than by real land-use behaviour.

## A Stable Metric: The Built-up per Capita Rate (BpCR)

The UN-Habitat metadata itself concedes that the ratio is hard to interpret and
offers built-up area per capita ($V/P$) as a secondary indicator
[@unhabitat2025metadata]. That quantity is a level, a snapshot at one point in
time. Following @Lu2026, we turn it into a trend by taking its annual growth in
logarithmic form:
$$
\text{BpCR}_{i,t}=\frac{1}{z}\ln\!\left(\frac{V_t/P_t}{V_{t-z}/P_{t-z}}\right)
= \text{LCR}_{\log}-\text{PGR}_{\log}.
$$
Because BpCR is the **difference of two logarithmic growth rates**, it is
(i) finite and well-defined for every country-period (no division by ~0),
(ii) symmetric (densification and sprawl are mirror images around zero),
(iii) additively decomposable, and (iv) directly interpretable: $\text{BpCR}>0$
means sprawl (more land per resident), $\text{BpCR}<0$ densification, and
$\text{BpCR}=0$ land growing exactly in step with population. BpCR reproduces the
policy signal of LCRPGR where the latter is well-behaved, but without the
instability; it complements rather than contradicts the official indicator.

# Literature Review

The literature on urban land use splits into two strands. One asks what *drives*
urban expansion; the other scrutinises how land-use efficiency is *measured*. This
paper sits at their intersection.

## Drivers of Urban Land Expansion

@seto2011meta synthesise 326 remote-sensing studies of global urban land
expansion between 1970 and 2000. They document that urban land has grown faster
than urban population almost everywhere, and that the expansion is driven mainly
by income in most regions and by population growth in India and Africa.
@oueslati2015sprawl move from description to econometrics, estimating the
determinants of sprawl on a panel of 282 European cities (1990, 2000, 2006)
within a monocentric-city framework. They find that rising income and population
both enlarge the urban footprint, and that cities keep spreading even where
population is flat. Together these studies establish income and population as the
leading drivers of urban land use, but they work mostly across cities or countries
and at a single cross-section, leaving open whether the same relationships hold
within a country as it develops over time.

## Critiques of the Indicator: Formula and Data

@nicolau2019lcrpgr assess the SDG 11.3.1 formula itself, comparing the official
LCRPGR with a land-use-efficiency-derived alternative at the municipality level in
mainland Portugal. They conclude that the LCRPGR formula is structurally unstable
and that a logarithmic, efficiency-based alternative is more robust.
@Lu2026 broaden the question to data sensitivity, evaluating land-use efficiency
across six built-up and four population datasets for 176 countries over
2000--2020. They show that efficiency estimates swing widely with the choice of
dataset and argue that a log-based framework is needed for reliable monitoring.
This strand establishes that the choice of metric, and not only the choice of
data, can change the conclusions drawn about sustainable urbanisation.

## Research Gap and Contribution

Our contribution sits at the intersection of the two strands. On measurement, we
adopt the log-based metric that both @nicolau2019lcrpgr and @Lu2026 recommend, in
the form of BpCR. On method, we make it dynamic: a two-way fixed-effects panel with
Arellano-Bond and Blundell-Bond GMM, estimated on 193 United Nations member states
over 1985--2020, that measures the persistence (path-dependence) of per-capita
land use. The result refines the drivers literature: while richer countries look
more sprawled in the cross-section, within a country growing richer is associated
with densification rather than sprawl.

# Data

## Data Sources

The empirical analysis draws on three official, global data providers, the same
sources used in international SDG monitoring. Built-up area, population, and the
urban mask come from the **Global Human Settlement Layer (GHSL)** of the European
Commission's Joint Research Centre [@pesaresi2024ghsl; @schiavina2023smod].
Within GHSL we use three products: GHS-BUILT-S for built-up surface, classified
from Landsat and Sentinel-2 imagery by machine learning; GHS-POP, which
disaggregates census counts onto that built-up surface; and GHS-SMOD, which
applies the Degree of Urbanisation (DEGURBA) method to flag urban cells. The
layers are available at five-year epochs; we use the observed epochs from 1985 to
2020 and discard the 2025 and 2030 modelled projections. Administrative units come
from the **Global Administrative Unit Layers (GAUL)** of the UN Food and
Agriculture Organization [@gaul2024], which provide UN-validated, politically
neutral country boundaries with global coverage. Socio-economic controls come from
the **United Nations**: GDP per capita in constant 2020 US dollars from UN SNAAMA
and net migration as a share of population from UN DESA World Population Prospects
2024 [@unsd2025snaama; @undesa2024wpp].

## Constructing the Panel Following the UN Metadata

A central design choice is that we build the panel exactly as the SDG 11.3.1
metadata prescribes [@unhabitat2025metadata], so that our BpCR and the official
LCRPGR are computed on identical inputs and the comparison is like-for-like. The
construction follows the metadata stages.

*Stage 0: define the analysis area.* SDG 11.3.1 is, by construction, an urban
(city-scale) indicator: it is measured inside cities rather than over whole
national territories. We delineate the urban area with DEGURBA on the GHS-SMOD
grid, classifying a cell as urban when its settlement code is at least 21 (urban
centres and dense or semi-dense towns and their suburbs). All built-up and
population quantities are measured only over these urban cells.

*Stage 1: measure built-up area and population at two points in time.* Our unit of
analysis is the country, so we aggregate the urban cells to the national level.
Crucially, we aggregate the *components*, not the metric: for each country and
five-year period we sum urban built-up area $V$ and urban population $P$ over all
urban cells to obtain national urban totals, and only then form the rates. This
follows the metadata's aggregate-then-rate rule, which warns against averaging
ratios across sub-units. Averaging city-level ratios would let a small town with a
near-zero population denominator dominate the national figure; summing the
components first avoids this and keeps the national rate well-defined.

*Stage 2: form the rates and the indicators.* From these national totals we
compute the Land Consumption Rate (arithmetic growth of built-up area per year)
and the Population Growth Rate (logarithmic growth of population per year), their
ratio LCRPGR, and the built-up-per-capita level. From the same totals we also
compute BpCR as the logarithmic growth of built-up area per capita, as defined in
Section 2. Two additional urban controls are derived from the same GHSL/SMOD
layers: the urban population share (urban population over total population) and
urban density (urban population over urban land area).

@fig-layers illustrates the construction for one example window, with each layer
drawn semi-transparently over a satellite basemap. Panel (a) is the basemap for
reference; panel (b) shows the GHS-BUILT-S surface as the built-up share of each cell; panel
(c) the GHS-SMOD Degree of Urbanisation, which classifies each cell from rural to
urban centre; and panel (d) the urban-masked built-up that actually enters the
analysis, retaining built-up only in cells classified as urban (SMOD at least 21). The same masking is applied to population, so that both components are
measured on a consistent urban footprint.

![Building the analysis input following the SDG 11.3.1 metadata, for the
Istanbul / Bosphorus window, with each layer overlaid semi-transparently on a
satellite basemap. (a) Satellite basemap (Esri) for reference. (b) GHS-BUILT-S
built-up surface, shaded by the share of each 1 km cell that is built up (0 to
about 50%). (c) GHS-SMOD Degree of Urbanisation, which classifies every cell from
water and rural (blues and greens) through suburban and town classes to the urban
centre (red); cells coded 21 or higher count as urban. (d) Urban-masked built-up,
the input actually used: the built-up surface of panel (b) kept only where panel
(c) is urban, on the same built-up colour scale. The gold line is the GAUL
country boundary (here the coastline).
](../04_outputs/figures/metadata_layers.png){#fig-layers width=88%}

The resulting dataset is a balanced-where-observed panel of 193 United Nations
member states at five-year epochs from 1985 to 2020 (@tbl-sample).

```{r sample-size}
#| label: tbl-sample
#| tbl-cap: "Estimation sample"
n_ctry <- length(unique(panel$iso3))
n_obs <- sum(!is.na(panel$bpcr))
data.frame(
    Quantity = c("Countries", "Country-period observations"),
    Value = c(n_ctry, n_obs)
) |>
    knitr::kable()
```

# Descriptive Statistics

Across 1985–2020 a clear majority of countries record $\text{BpCR}>0$: built-up
area is expanding faster than urban population for most of the world, i.e. sprawl
is the global norm rather than the exception. The contrast with LCRPGR is
instructive -- LCRPGR's distribution is dominated by extreme tail values
concentrated exactly where population growth is near zero, whereas BpCR is
well-behaved across the whole sample. Regional trajectories show convergence
towards balance in some fast-urbanising economies while others continue to
sprawl.

<!-- TODO: insert global BpCR distribution figure and the regional-trend figure
from 04_outputs/figures/ once finalised. -->

# Empirical Strategy

**Research question.** What are the economic drivers of per-capita urban land use,
and how persistent is it over time? We test five hypotheses (@tbl-hyp).

```{r tbl-hyp}
#| label: tbl-hyp
#| tbl-cap: "Hypotheses"
data.frame(
    ` ` = c("H1", "H2", "H3", "H4", "H5"),
    Hypothesis = c(
        "BpCR captures land-use efficiency more reliably than LCRPGR",
        "Higher urban density is associated with less land per capita",
        "Higher income is associated with less land per capita",
        "Net in-migration densifies cities",
        "Per-capita land use is path-dependent"
    ),
    Expected = c("better fit", "$-$", "$-$", "$-$", "$+$"),
    check.names = FALSE
) |>
    knitr::kable(
        format = "latex", booktabs = TRUE, escape = FALSE, linesep = "",
        align = c("c", "l", "c")
    ) |>
    kableExtra::kable_styling(latex_options = "HOLD_position")
```

We run **one specification** with each metric
as the dependent variable, so the comparison is like-for-like:
$$
\text{Metric}_{i,t}=\rho\,\text{Metric}_{i,t-1}
 + \beta_1\ln(\text{UrbDens})_{i,t} + \beta_2\,\text{UrbShare}_{i,t}
 + \beta_3\ln(\text{GDPpc})_{i,t} + \beta_4\,\text{NetMigr}_{i,t}
 + \mu_i + \delta_t + \varepsilon_{i,t},
$$
with country fixed effects $\mu_i$, period fixed effects $\delta_t$, and standard
errors clustered by country. We proceed in two stages. **(1) Static** (dropping
the lagged dependent variable) to isolate the role of the *metric* -- LCRPGR versus
BpCR. **(2) Dynamic** (adding the lagged DV) to capture path-dependence. Because
the lagged-DV coefficient is **downward-biased under fixed effects (Nickell
bias)**, we correct it with **Arellano-Bond difference GMM** and **Blundell-Bond
system GMM**, instrumenting with deeper lags. Instrument validity is assessed with
the **Sargan/Hansen** over-identification test and the **AR(2)** serial-correlation
test (both $p>0.10$ indicating validity).

# Results

**The metric matters.** On identical data, the static LCRPGR specification is
contaminated by its unstable tails, while the BpCR specification is well-behaved;
the two diverge precisely where $\text{PGR}$ is small, confirming that LCRPGR's
extremes are an arithmetic artefact rather than a behavioural signal.

**The final dynamic model.** @tbl-gmm reports the preferred specification: a
fixed-effects stepwise build-up alongside the Arellano-Bond and Blundell-Bond GMM
estimates.

```{r gmm-table}
#| label: tbl-gmm
#| tbl-cap: "Final dynamic-panel model: FE stepwise build-up, Arellano-Bond (difference) and Blundell-Bond (system) GMM. Coefficient with significance stars on top, clustered/robust standard error in parentheses below. *** p<0.01, ** p<0.05, * p<0.1."
if (!requireNamespace("kableExtra", quietly = TRUE)) install.packages("kableExtra")
mc <- as.data.frame(gmm$main_combo, stringsAsFactors = FALSE)
# coefficient cells look like "X***<br>(se)" -> stacked \makecell{X*** \\ (se)}
to_cell <- function(x) {
    ifelse(
        grepl("<br>", x, fixed = TRUE),
        paste0("\\makecell[c]{", sub("<br>", " \\\\ ", x, fixed = TRUE), "}"),
        x
    )
}
for (j in 2:ncol(mc)) mc[[j]] <- to_cell(mc[[j]])
names(mc)[1] <- ""
esc <- function(x) gsub("%", "\\%", x, fixed = TRUE) # literal % -> \% (LaTeX)
names(mc) <- esc(names(mc))
mc[[1]] <- esc(mc[[1]]) # escape % in row labels too
knitr::kable(mc,
    format = "latex", booktabs = TRUE, escape = FALSE, linesep = "",
    align = c("l", rep("c", ncol(mc) - 1))
) |>
    kableExtra::kable_styling(latex_options = c("HOLD_position", "scale_down"))
```

In the Arellano-Bond specification the lagged term is large and significant
($\rho\approx 0.61$): roughly 60% of last period's BpCR persists, and the GMM
$\rho$ is about three times the FE estimate, exactly as the Nickell bias predicts,
confirming **strong path-dependence** in per-capita land use. Among the drivers,
**urban density** enters negative (a compact-city effect: denser cities consume
less new land per resident), **net migration** negative (in-migrants pack into the
existing stock faster than built-up grows), and **GDP per capita** negative --
within countries, getting richer is associated with **densification, not sprawl**.
The level of urbanisation (urban population share) is insignificant once the rest
is controlled, suggesting that urban **form**, not stage, drives land use.

# Robustness

The findings are stable across a 2000–2020 subsample, a restriction to
shrinking-population countries, and alternative instrument depths; the Sargan and
AR(2) diagnostics pass throughout, supporting instrument validity.

<!-- TODO: tabulate the 2000-2020 and shrinking-population robustness results. -->

# Discussion

The within-country result that income is associated with densification may appear
to contradict the cross-sectional literature in which richer countries have
historically sprawled. The two are complementary: *across* countries, higher
income has coincided with more sprawl historically, but *within* a country, rising
income over time is associated with denser, not more sprawling, urban land use.
Measuring this distinction requires a metric that is stable within country-period --
which BpCR provides and LCRPGR does not.

# Conclusion

The official SDG 11.3.1 indicator is statistically unstable by construction. The
Built-up per Capita Rate is a drop-in, log-based alternative that is finite,
symmetric, and interpretable for every country-period, and it reproduces the
policy signal without the instability. Estimated dynamically on a global panel,
per-capita urban land use is strongly path-dependent and, within countries,
declines with density, in-migration, and income. The contribution is as much
**methodological** -- a stable metric and a bias-corrected dynamic estimator -- as
substantive, and it complements rather than overturns the existing literature.

# References {.unnumbered}

::: {#refs}
:::
