---
title: "Urban Land-Use Efficiency"
subtitle: "A Stable Metric for SDG 11.3.1 and Its Economic Drivers"
doc-type: "Seminar Paper"
course-info: "Development Economics · SoSe 2026"
lecturer: "Lecturer: Prof. Dr. Jürgen Bitzer"
author:
  - name: Ömer Furkan Çoban
    email: oemer.furkan.coban@uni-oldenburg.de
    affiliations:
      - name: Carl von Ossietzky University of Oldenburg
        department: Applied Economics and Data Science
        city: Oldenburg
        country: Germany
date: today
abstract: |
  The official SDG 11.3.1 indicator, the ratio of land consumption rate to
  population growth rate (LCRPGR), is unstable: it divides by a growth rate that
  approaches zero and mixes an arithmetic numerator with a logarithmic
  denominator. We propose the Built-up per Capita Rate (BpCR), the log change in
  built-up area per person, which equals the difference of two log growth rates
  and is therefore well-defined for every country-period. Using a harmonised
  global panel of 193 UN member states (1985–2020) built from GHSL built-up and
  population grids under the Degree of Urbanisation, we (i) document where LCRPGR
  breaks down, (ii) show BpCR reproduces the policy signal without the
  instability, and (iii) estimate its economic drivers with two-way fixed-effects
  and dynamic-panel GMM models, documenting heterogeneity across development groups.
keywords:
  - SDG 11.3.1
  - Urban land-use efficiency
  - Built-up per capita rate (BpCR)
  - LCRPGR
  - Dynamic panel GMM
  - GHSL / Degree of Urbanisation
journal:
  name: Development Economics seminar working paper
format:
  elsevier-pdf:
    keep-tex: true
    template-partials:
      - before-body.tex
    include-in-header:
      text: |
        \usepackage{makecell}
        \usepackage{float}
        \geometry{top=3cm,bottom=3cm}
        \floatplacement{figure}{H}
        \floatplacement{table}{H}
        \setlength{\floatsep}{6pt plus 2pt minus 2pt}
        \setlength{\textfloatsep}{7pt plus 2pt minus 2pt}
        \setlength{\intextsep}{6pt plus 2pt minus 2pt}
    journal:
      formatting: preprint      # preprint | review | doubleblind
      model: 3p                 # 1p | 3p | 5p (3p = standard single-column-ish)
      cite-style: authoryear    # elsarticle author-year citations
bibliography: references.bib
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
#| include: false
library(here)
library(dplyr)
library(readr)
library(fixest)
panel <- read_csv(here("03_datasets/processed/reg_panel_urban.csv"),
    show_col_types = FALSE
)
gmm <- readRDS(here("03_datasets/processed/dynamic_gmm.rds"))
```

# Introduction

The world is urbanising rapidly: by the middle of this century close to
two-thirds of humanity will live in cities, and almost all of the projected
growth in built-up land will occur in developing countries
[@gao2020]. The scale of expansion is striking: every year the world's
built-up area grows by roughly the size of the Netherlands [@seto2012global].
For development policy the central question is not whether cities expand, but how
efficiently they do so, that is, how much additional land each new urban resident
consumes.
When built-up area grows in step with population, cities stay compact; when it
grows much faster, they sprawl. Sprawl is costly: it paves over fringe farmland for
good, locks residents into car dependence as homes, jobs, and services spread
apart, raises the per-resident cost of infrastructure that must reach fewer people
over more ground, and lifts per-capita emissions. Measuring land-use efficiency
well is therefore a precondition for monitoring sustainable urbanisation.

The United Nations monitors this through target 11.3 of its Sustainable
Development Goals (SDGs), whose official indicator 11.3.1 is the ratio of the Land
Consumption Rate to the Population Growth Rate (LCRPGR) [@unhabitat2025metadata].
Because LCRPGR is the agreed global yardstick, its statistical properties are not
a technical detail: they determine how every country's progress is judged and
compared. Yet the indicator is fragile by construction: because the population-growth rate
sits in the denominator, the ratio diverges and flips sign whenever population is
roughly stable, so in exactly the countries where land-use efficiency matters most
for policy it becomes uninformative.

This paper makes three contributions. First, it documents, on a harmonised global
dataset, that LCRPGR is statistically unstable and that its instability is
concentrated where population growth is near zero. Second, it proposes a simple,
log-based alternative, the Built-up per Capita Rate (BpCR), the logarithmic growth
rate of built-up area per capita. As the difference of two logarithmic growth
rates it is finite for every country-period and directly interpretable, and it
reproduces the policy signal of LCRPGR where the latter is well-behaved.
Third, using the stable metric, the paper estimates the economic and structural
drivers of per-capita urban land use on a panel of 193 United Nations member
states observed at five-year intervals between 1985 and 2020.

The empirical analysis builds the panel from satellite-based settlement data
under the global Degree of Urbanisation standard, so that built-up area and
population are measured consistently. To address the well-known downward bias of
the lagged dependent variable in dynamic fixed-effects models, the persistence of
per-capita land use is estimated with Arellano-Bond and Blundell-Bond generalised
method of moments (GMM) estimators. Three results stand out. Sprawl is the global
norm rather than the exception: built-up area per capita rises in a clear
majority of country-periods. Per-capita land use is strongly path-dependent, with
about three-fifths of one period's value carried into the next. And within
countries, denser cities, net in-migration, and higher income are all associated
with less land consumed per resident, indicating that, over time, economic growth
is accompanied by densification rather than sprawl. These patterns hold across
development groups, with the compact-city effect strongest in the least-developed
economies.

The remainder of the paper is organised as follows. @sec-measurement sets out the
SDG 11.3.1 indicator, its structural limitations, and the BpCR alternative.
@sec-literature reviews the related literature. @sec-data describes the data and
@sec-descriptive presents descriptive evidence. @sec-strategy details the
empirical strategy and @sec-results reports the results. @sec-discussion discusses
the findings and @sec-conclusion concludes.

# Measurement Framework {#sec-measurement}

This section sets out the conceptual core of the paper. We first define the
official SDG 11.3.1 indicator and show why it is statistically unstable, and we
then introduce the stable, log-based alternative used in the empirical analysis.

## The SDG 11.3.1 Indicator and Its Limitations

Following the UN-Habitat metadata [@unhabitat2025metadata], the official
indicator divides an arithmetic land consumption rate by a logarithmic
population growth rate:
$$
\text{LCRPGR}_i = \frac{\text{LCR}_i}{\text{PGR}_i},\qquad
\text{LCR}_i=\frac{V_{i,t}-V_{i,t-z}}{V_{i,t-z}}\cdot\frac{1}{z},\qquad
\text{PGR}_i=\frac{\ln(P_{i,t}/P_{i,t-z})}{z},
$$
where, for country $i$, $V_{i,t}$ is the urban built-up area at the
end of the period and $V_{i,t-z}$ the built-up area at the
start; $P_{i,t}$ and $P_{i,t-z}$ are the corresponding urban populations; $z$ is the
length of the period in years; $\ln(\cdot)$ denotes the natural logarithm. The Land Consumption Rate $(\text{LCR})$ is thus
the arithmetic growth of built-up area, expressed per year, and the Population
Growth Rate $(\text{PGR})$ is the logarithmic (continuous) growth of population per
year; the indicator $\text{LCRPGR}$ is their ratio. Two structural problems
follow. **(1) Division by a near-zero denominator.** When
population is roughly stable ($\text{PGR}\approx 0$) the ratio diverges towards
$\pm\infty$ and flips sign, so small, ordinary demographic differences produce
wild, uninterpretable values. **(2) Arithmetic-versus-logarithmic asymmetry.**
The numerator is an arithmetic growth rate while the denominator is logarithmic,
so the ratio is not symmetric and does not aggregate cleanly over time
[@nicolau2019lcrpgr]. The result is a metric whose tails are dominated by an
arithmetic artefact rather than by real land-use behaviour.

## A Stable Metric: The Built-up per Capita Rate (BpCR)

The UN-Habitat metadata itself concedes that the ratio is hard to interpret and
offers built-up area per capita ($V/P$) as a secondary indicator
[@unhabitat2025metadata]. That quantity is a level, a snapshot at one point in
time. Following @Lu2026, we turn it into a trend by taking its annual growth in
logarithmic form:
$$
\text{BpCR}_{i,t}=\frac{1}{z}\ln\!\left(\frac{V_{i,t}/P_{i,t}}{V_{i,t-z}/P_{i,t-z}}\right)
= \text{LCR}_{i,\log}-\text{PGR}_{i,\log},
$$
where $\text{LCR}_{i,\log}=\frac{1}{z}\ln(V_{i,t}/V_{i,t-z})$ and
$\text{PGR}_{i,\log}=\frac{1}{z}\ln(P_{i,t}/P_{i,t-z})$ are the logarithmic growth
rates of built-up area and population. Here $\text{PGR}_{i,\log}$ is identical to
the PGR of the official indicator above, which is already logarithmic, whereas
$\text{LCR}_{i,\log}$ is the logarithmic counterpart of the arithmetic
$\text{LCR}_i$ in that indicator. Replacing the arithmetic LCR with its
logarithmic form is precisely what removes the arithmetic-versus-logarithmic
asymmetry that destabilises LCRPGR.

Because BpCR is the **difference of two logarithmic growth rates**, it is
(i) finite and well-defined for every country-period (no division by ~0),
(ii) symmetric (densification and sprawl are mirror images around zero),
(iii) additively decomposable, and (iv) directly interpretable: $\text{BpCR}>0$
means sprawl (more land per resident), $\text{BpCR}<0$ densification, and
$\text{BpCR}=0$ land growing exactly in step with population. BpCR reproduces the
policy signal of LCRPGR where the latter is well-behaved, but without the
instability; it complements rather than contradicts the official indicator.

In the empirical sections we benchmark BpCR against the logarithmic form of the
ratio, $\text{LCRPGR}_{\log}=\text{LCR}_{i,\log}/\text{PGR}_{i,\log}$, which we
label *LCRPGR (log)*. This is a deliberately conservative benchmark: it already
replaces the arithmetic LCR with its logarithmic counterpart, so it removes the
arithmetic-versus-logarithmic asymmetry and differs from BpCR only in being a
ratio rather than a difference. Any instability it shows therefore comes purely
from dividing by a near-zero growth rate; the official arithmetic ratio would be
more unstable still.

# Literature Review {#sec-literature}

The literature on urban land use splits into two strands. One asks what *drives*
urban expansion; the other scrutinises how land-use efficiency is *measured*. This
paper sits at their intersection.

## Drivers of Urban Land Expansion

@seto2011meta synthesise 326 remote-sensing studies of global urban land
expansion between 1970 and 2000. They document that urban land has grown faster
than urban population almost everywhere, and that the expansion is driven mainly
by income in most regions and by population growth in India and Africa.
@oueslati2015sprawl move from description to econometrics, estimating the
determinants of sprawl on a panel of 282 European cities (1990, 2000, 2006)
within a monocentric-city framework. They find that rising income and population
both enlarge the urban footprint, and that cities keep spreading even where
population is flat. Together these studies establish income and population as the
leading drivers of urban land use, but they work mostly across cities or countries
and at a single cross-section, leaving open whether the same relationships hold
within a country as it develops over time.

## Critiques of the Indicator: Formula and Data

@nicolau2019lcrpgr assess the SDG 11.3.1 formula itself, comparing the official
LCRPGR with a land-use-efficiency-derived alternative at the municipality level in
mainland Portugal. They conclude that the LCRPGR formula is structurally unstable
and that a logarithmic, efficiency-based alternative is more robust.
@Lu2026 broaden the question to data sensitivity, evaluating land-use efficiency
across six built-up and four population datasets for 176 countries over
2000--2020. They show that efficiency estimates swing widely with the choice of
dataset and argue that a log-based framework is needed for reliable monitoring.
This strand establishes that the choice of metric, and not only the choice of
data, can change the conclusions drawn about sustainable urbanisation.

## Research Gap and Contribution

Our contribution sits at the intersection of the two strands. On measurement, we
adopt the log-based metric that both @nicolau2019lcrpgr and @Lu2026 recommend, in
the form of BpCR. On method, we make it dynamic: a two-way fixed-effects panel with
Arellano-Bond and Blundell-Bond GMM, estimated on 193 United Nations member states
over 1985--2020, that measures the persistence of per-capita
land use and its economic drivers.

# Data {#sec-data}

## Data Sources

The empirical analysis draws on three official, global data providers, the same
sources used in international SDG monitoring. Built-up area, population, and the
urban mask come from the **Global Human Settlement Layer (GHSL)** of the European
Commission's Joint Research Centre [@pesaresi2024ghsl; @schiavina2023smod].
Within GHSL we use three products: GHS-BUILT-S for built-up surface, classified
from Landsat and Sentinel-2 imagery by machine learning; GHS-POP, which
disaggregates census counts onto that built-up surface; and GHS-SMOD, which
applies the Degree of Urbanisation (DEGURBA) method to flag urban cells. The
layers are available at five-year epochs; we use the observed epochs from 1985 to
2020 and discard the 2025 and 2030 modelled projections. Administrative units come
from the **Global Administrative Unit Layers (GAUL)** of the UN Food and
Agriculture Organization [@gaul2024], which provide UN-validated, politically
neutral country boundaries with global coverage. Socio-economic controls come from
the **United Nations**: GDP per capita in constant 2020 US dollars from UN SNAAMA
and net migration as a share of population from UN DESA World Population Prospects
2024 [@unsd2025snaama; @undesa2024wpp]. For the heterogeneity analysis we classify countries into United Nations development groups, developed economies, developing economies, and least-developed countries (LDCs), using the UN M49 regional classification together with the UN Committee for Development Policy LDC list (obtained through the `countrycode` R package).

@tbl-data summarises the technical characteristics of each dataset.

```{r tbl-data}
#| label: tbl-data
#| tbl-cap: "Dataset specifications. GHSL grids use the World Mollweide projection (ESRI:54009); built-up and population at 100 m, the SMOD urban mask at 1 km. Vector and tabular sources show their native form."
data.frame(
    Dataset = c(
        "GHS-BUILT-S R2023A", "GHS-POP R2023A", "GHS-SMOD R2023A",
        "GAUL 2024/2025", "UN SNAAMA", "UN DESA WPP 2024", "UN M49 / CDP"
    ),
    Provider = c("EC JRC", "EC JRC", "EC JRC", "FAO", "UN Stats", "UN DESA", "UN"),
    Type = c("Raster", "Raster", "Raster", "Vector", "Tabular", "Tabular", "Classification"),
    Resolution = c("100 m", "100 m", "1 km", "Levels 0-2", "Country-year", "Country-year", "Country"),
    CRS = c(
        "Mollweide (54009)", "Mollweide (54009)", "Mollweide (54009)",
        "WGS84 (EPSG:4326)", "--", "--", "--"
    ),
    Years = c(
        "1975-2030, 5-yr", "1975-2030, 5-yr", "1975-2030, 5-yr",
        "2024/2025", "1970-2024", "1950-2023", "2024"
    ),
    Variable = c(
        "Built-up area (m2/cell)", "Population (persons/cell)",
        "Degree of Urbanisation (class 10-30)", "Admin boundaries",
        "GDP p.c. (const. 2020 US$)", "Net migration (CNMR /1000)", "Development group"
    ),
    check.names = FALSE
) |>
    knitr::kable(
        format = "latex", booktabs = TRUE, escape = TRUE, linesep = "",
        align = c("l", "l", "l", "l", "l", "l", "l")
    ) |>
    kableExtra::kable_styling(latex_options = c("HOLD_position", "scale_down"))
```

## Constructing the Panel Following the UN Metadata

A central design choice is that we build the panel exactly as the SDG 11.3.1
metadata prescribes [@unhabitat2025metadata], so that our BpCR and the official
LCRPGR are computed on identical inputs and the comparison is like-for-like. The
construction follows the metadata stages.

![Building the analysis input following the SDG 11.3.1 metadata (Istanbul /
Bosphorus window), each layer drawn over an Esri satellite basemap. (a) Basemap.
(b) GHS-BUILT-S built-up surface (share of each 1 km cell built up). (c) GHS-SMOD
Degree of Urbanisation; cells coded 21 or higher count as urban. (d) Urban-masked
built-up, the input actually used: panel (b) kept only where panel (c) is urban.
The gold line is the GAUL country boundary.
](../04_outputs/figures/metadata_layers.png){#fig-layers width=74%}

@fig-layers illustrates the construction for one example window, with each layer
drawn over a satellite basemap. Panel (a) is the basemap for
reference; panel (b) shows the GHS-BUILT-S surface as the built-up share of each cell; panel
(c) the GHS-SMOD Degree of Urbanisation, which classifies each cell from rural to
urban centre; and panel (d) the urban-masked built-up that actually enters the
analysis, retaining built-up only in cells classified as urban (SMOD at least 21). The same masking is applied to population, so that both components are
measured on a consistent urban footprint.
*Analysis area.* SDG 11.3.1 is an urban (city-scale) indicator, measured inside
cities rather than over whole national territories. We delineate the urban area
with DEGURBA on the GHS-SMOD grid (a cell is urban when its settlement code is at
least 21, that is, urban centres and dense or semi-dense towns and suburbs) and
measure all built-up and population quantities only over these urban cells.

*Components and aggregation.* Our unit of analysis is the country, so we aggregate
the urban cells to the national level, aggregating the *components* rather than the
metric: for each country-period we sum urban built-up area $V$ and urban population
$P$ before forming any rate. This follows the metadata's aggregate-then-rate rule;
averaging city-level ratios would let a small town with a near-zero population
denominator dominate the national figure, whereas summing components first keeps
the national rate well-defined.

*Rates and indicators.* From these national totals we form the Land Consumption
Rate (arithmetic growth of built-up area), the Population Growth Rate (logarithmic
growth of population), their ratio LCRPGR, the built-up-per-capita level, and BpCR
as the logarithmic growth of built-up area per capita (@sec-measurement).

*Control variables.* The two urban controls are computed from the same GHSL/SMOD
layers. **Urban density** is national urban population divided by urban land area,
the area of urban cells (distinct from, and about ten times larger than, the
built-up surface in the dependent variable), used in logs. **Urban population
share** is urban population divided by national total population. The remaining
controls come from international sources: **GDP per capita** (constant 2020 US
dollars, logs) from UN SNAAMA and **net migration** (share of population) from UN
DESA WPP 2024. All variables align to the same country-period grid.

The resulting dataset is a balanced-where-observed panel of 193 United Nations
member states across eight five-year epochs from 1985 to 2020, giving 1,538
country-periods with an observed BpCR. The dynamic regressions in @sec-results use
the subset with the lagged dependent variable and all four controls present
(complete-case estimation), which reduces the sample to roughly 1,500
country-period observations.

# Descriptive Statistics {#sec-descriptive}

```{r desc-stats}
#| include: false
d <- subset(panel, !is.na(bpcr) & year >= 1985 & year <= 2020)
ds <- list(
  n      = nrow(d),
  lcr_sd = sd(d$lcrpgr_log, na.rm = TRUE),
  lcr_lo = min(d$lcrpgr_log, na.rm = TRUE),
  lcr_hi = max(d$lcrpgr_log, na.rm = TRUE),
  bp_sd  = sd(d$bpcr, na.rm = TRUE),
  ratio  = round(sd(d$lcrpgr_log, na.rm = TRUE) / sd(d$bpcr, na.rm = TRUE), -1),
  shr_cp = 100 * mean(d$bpcr > 0, na.rm = TRUE),
  med    = median(d$bpcr, na.rm = TRUE),
  shr_l1 = 100 * mean(d$lcrpgr > 1, na.rm = TRUE)
)
cm <- aggregate(bpcr ~ iso3, d, mean)
ds$ctry_pos <- sum(cm$bpcr > 0); ds$ctry_tot <- nrow(cm)
ds$ctry_pct <- round(100 * ds$ctry_pos / ds$ctry_tot)
```

```{r tbl-desc}
#| label: tbl-desc
#| tbl-cap: "Summary statistics, 1985-2020 (country-periods with an observed BpCR). LCRPGR is the official log ratio; BpCR is the log growth of built-up area per capita."
dd <- subset(panel, !is.na(bpcr) & year >= 1985 & year <= 2020)
vars <- c(bpcr="BpCR", lcrpgr_log="LCRPGR (log)", ln_urban_density="ln(Urban Density)",
          urban_pop_share="Urban Pop. Share", ln_gdp_pc="ln(GDP per capita)",
          net_migr_pct="Net Migration (\\% Pop)")
rows <- lapply(names(vars), function(v){
  x <- dd[[v]]
  data.frame(Variable=vars[[v]], N=sum(!is.na(x)),
             Mean=mean(x,na.rm=TRUE), SD=sd(x,na.rm=TRUE),
             Min=min(x,na.rm=TRUE), Median=median(x,na.rm=TRUE), Max=max(x,na.rm=TRUE),
             check.names=FALSE)
})
tab <- do.call(rbind, rows)
num <- sapply(tab[,3:7], function(c) ifelse(abs(c)>=100, sprintf("%.0f",c), sprintf("%.3f",c)))
tab[,3:7] <- num
knitr::kable(tab, format="latex", booktabs=TRUE, escape=FALSE, linesep="",
             align=c("l","r","r","r","r","r","r"), row.names=FALSE) |>
  kableExtra::kable_styling(latex_options=c("HOLD_position"), font_size=9)
```

@tbl-desc reports summary statistics for the dependent variables and controls over
the full 1985--2020 sample.

Even before any regression, the raw data already reveal why the choice of metric
matters. Over the full 1985--2020 sample (`r ds$n` country-periods), LCRPGR (log)
has a standard deviation of about `r round(ds$lcr_sd)` and
ranges from roughly `r round(ds$lcr_lo)` to `r round(ds$lcr_hi)`: its division by a
near-zero population-growth rate produces enormous, sign-flipping outliers. BpCR,
by contrast, stays in a tight band with a standard deviation of about
`r sprintf("%.3f", ds$bp_sd)`, roughly `r format(ds$ratio, big.mark=",")` times
narrower. The two metrics describe the same underlying built-up and population
data, yet one is dominated by an arithmetic artefact and the other is not.

Sprawl is the global norm rather than the exception. BpCR is positive in
`r sprintf("%.0f", ds$shr_cp)`% of country-periods, and
`r ds$ctry_pos` of `r ds$ctry_tot` countries (`r ds$ctry_pct`%) have a positive
average BpCR over the period, meaning built-up area per capita is rising: cities
are consuming more land per resident. The median country-period BpCR is
`r sprintf("%+.4f", ds$med)`, and the official LCRPGR exceeds one (its sprawl
threshold) in `r sprintf("%.0f", ds$shr_l1)`% of cases. The qualitative message
is the same under either metric; the difference is that BpCR delivers it without
the numerical instability.

Country trajectories make the contrast concrete (@fig-trends). For a set of
regional representatives, LCRPGR (log) swings erratically from one period
to the next, crossing zero and spiking to large positive and negative values that
are hard to read as land-use behaviour. The BpCR series for the same countries are
smooth and interpretable: most drift gradually towards balance, while the
fastest-urbanising economies show the strongest sprawl earlier in the period
before easing. The same data, read through a stable metric, become legible.

![Country trajectories for seven regional representatives, 1985--2020. (a) LCRPGR
(log) swings erratically across its zero and unity reference lines; (b) BpCR is
smooth on the same data, with the fastest-urbanising economies showing the
strongest early sprawl.](../04_outputs/figures/country_trends.pdf){#fig-trends width=88%}

# Empirical Strategy {#sec-strategy}

**Research question.** We ask two linked questions. First, a substantive one: what
economic and structural factors drive *within-country* changes in per-capita urban
land use over time, and how persistent are those changes? Second, a measurement
one: does the official SDG 11.3.1 indicator capture this reliably, or does the
choice of metric change the conclusions? We formalise these in five hypotheses
(@tbl-hyp).

```{r tbl-hyp}
#| label: tbl-hyp
#| tbl-cap: "Hypotheses"
data.frame(
    ` ` = c("H1", "H2", "H3", "H4", "H5"),
    Hypothesis = c(
        "BpCR captures land-use efficiency more reliably than LCRPGR",
        "Higher urban density is associated with less land per capita",
        "Higher income is associated with less land per capita",
        "Net in-migration densifies cities",
        "Per-capita land use is path-dependent"
    ),
    Expected = c("Better Fit", "$-$", "$-$", "$-$", "$+$"),
    check.names = FALSE
) |>
    knitr::kable(
        format = "latex", booktabs = TRUE, escape = FALSE, linesep = "",
        align = c("c", "l", "c")
    ) |>
    kableExtra::kable_styling(latex_options = "HOLD_position")
```

## Specification

We proceed in two stages, each with its own specification. In the **static** stage
we compare the two metrics on an equal footing. For each metric
$\text{Metric} \in \{\text{LCRPGR}, \text{BpCR}\}$ we estimate a two-way fixed-effects model
without dynamics:
$$
\small \text{Metric}_{i,t} = \beta_1\ln(\text{UrbDens})_{i,t} + \beta_2\,\text{UrbShare}_{i,t} + \beta_3\ln(\text{GDPpc})_{i,t} + \beta_4\,\text{NetMigr}_{i,t} + \mu_i + \delta_t + \varepsilon_{i,t}.
$$ {#eq-static}
Because the data, controls, and fixed effects are held identical, any difference
between the LCRPGR and BpCR estimates of @eq-static is attributable to the metric
alone (H1).

In the **dynamic** stage, having established BpCR as the usable metric, we add its
own lag to measure path-dependence, so the dependent variable is now specifically
BpCR:
$$
\small \text{BpCR}_{i,t} = \rho\,\text{BpCR}_{i,t-1} + \beta_1\ln(\text{UrbDens})_{i,t} + \beta_2\,\text{UrbShare}_{i,t} + \beta_3\ln(\text{GDPpc})_{i,t} + \beta_4\,\text{NetMigr}_{i,t} + \mu_i + \delta_t + \varepsilon_{i,t}.
$$ {#eq-dynamic}
The coefficient $\rho$ measures how strongly a country's per-capita land use
persists from one period to the next (H5).

In both specifications the country fixed effects $\mu_i$ absorb all time-invariant
national characteristics (geography, institutions, the initial level of
urbanisation), so identification comes from changes *within* a country over time
rather than from comparisons *between* countries. The period fixed effects
$\delta_t$ absorb global shocks common to all countries in a given epoch
(technology, the global business cycle, and any epoch-specific features of the
satellite record). Standard errors are clustered by country to allow for serial
correlation within a country's series.

The four regressors are chosen because each has a clear theoretical link to how
much land a city consumes per resident. **Urban density** captures the
compact-city mechanism: where settlement is already dense there is both less room
and less need to expand outward, so additional residents are accommodated within
the existing footprint, implying less new land per capita (H2).^[Although both the dependent variable and this control involve urban population, they are not mechanically linked: the dependent variable is the growth rate of built-up surface per capita, whereas the control is the level of population per unit of urban land area, and built-up surface is only about a tenth of urban land area. The two are therefore far from collinear, and the two-way fixed effects further absorb persistent cross-country scale differences. As a further check, the income and migration results are unchanged when urban density enters with a one-period lag; the density coefficient itself is sensitive to this timing, consistent with the contemporaneous compact-city association partly reflecting simultaneity.] **GDP per capita**
has an ambiguous sign a priori. Across countries, higher income has historically
gone with more sprawl, through larger homes and car-based mobility; but within a
country, rising income can also fund vertical and infill construction,
redevelopment, and stronger land-use regulation, which densify rather than expand
the footprint. Our within-country design tests which force dominates over time
(H3). **Net migration** proxies population pressure: in-migrants typically settle
in the existing urban stock, so housing demand is absorbed by densification faster
than the built-up area grows (H4). Finally, the **lagged dependent variable**
captures path-dependence: the building stock, road network, and zoning that shape
land use are long-lived, so a country's per-capita land use tends to persist from
one period to the next (H5). H1 concerns the metric itself and is addressed by
comparing the LCRPGR and BpCR runs.

## Dynamic-Panel Estimation and the Nickell Bias

Adding a lagged dependent variable to a fixed-effects model introduces a
well-known problem. The within (demeaning) transformation that removes $\mu_i$
makes the transformed lagged regressor mechanically correlated with the
transformed error, biasing the estimate of $\rho$ downward; the bias is of order
$1/T$ and is therefore substantial in a panel with only eight periods such as
ours. To correct it we use generalised method of moments (GMM) estimators
designed for dynamic panels.

The **Arellano-Bond** (difference) estimator first-differences the equation to
eliminate the country fixed effects and instruments the differenced lagged term
with deeper lags in levels (lags two to five of BpCR). The **Blundell-Bond** (system) estimator augments this
with a second, levels equation in which the lagged term is instrumented by its own
past differences; this added moment information makes it more efficient when the
series is persistent, as BpCR is. We treat **Arellano-Bond as the preferred
specification** and report Blundell-Bond alongside it as a robustness comparison.
Both are estimated in two steps with Windmeijer-corrected standard errors [@windmeijer2005], and the
instrument matrix is collapsed (one column per lag distance) to keep the instrument count from proliferating with the time dimension.

The two estimators draw on the sample differently: Arellano-Bond uses the
differenced equation alone, so differencing costs the first usable period per
country, while Blundell-Bond adds a levels equation that recovers it and, more
importantly, contributes extra moment conditions rather than a larger sample. The
exact counts are reported with the results in @sec-results.

Instrument validity is assessed with the **Sargan/Hansen** test of
over-identifying restrictions and the **AR(2)** test for second-order serial
correlation in the differenced residuals; in both, a $p$-value above 0.10
indicates that the instruments are valid. The estimation uses complete cases
(observations with the lagged dependent variable and all four controls present),
about 1,500 country-periods before the GMM transformations described above.

# Results {#sec-results}

## The Metric Matters: LCRPGR vs. BpCR

We first compare the two metrics in the static (lag-free) two-way fixed-effects
model of @eq-static, holding the data, controls, and fixed effects identical so that any
difference is due to the dependent variable alone. To keep every specification
comparable, all models in this section, static and dynamic, are estimated on the
common sample for which the lagged dependent variable and all four controls are
available; the static models therefore report the same number of observations as
the dynamic ones (1,499), so that differences across columns reflect the
specification and not a change in the sample. The contrast between the metrics is
stark (@tbl-static).

```{r tbl-static}
#| label: tbl-static
#| tbl-cap: "Static two-way fixed-effects model on the estimation sample (1985-2020), same sample and controls for both metrics."
suppressMessages({library(fixest); library(dplyr)})
# same estimation sample as the dynamic model: complete cases with the lagged DV
sp <- panel |>
  filter(year <= 2020, !is.na(bpcr), !is.na(lcrpgr_log)) |>
  arrange(iso3, year) |> group_by(iso3) |> mutate(bpcr_L = lag(bpcr)) |> ungroup() |>
  filter(!is.na(ln_urban_density), !is.na(urban_pop_share),
         !is.na(ln_gdp_pc), !is.na(net_migr_pct), !is.na(bpcr_L))
ctrl <- "ln_urban_density + urban_pop_share + ln_gdp_pc + net_migr_pct"
mL <- feols(as.formula(paste("lcrpgr_log ~", ctrl, "| iso3 + year")), sp, vcov = ~iso3)
mB <- feols(as.formula(paste("bpcr ~",       ctrl, "| iso3 + year")), sp, vcov = ~iso3)
star <- function(p) ifelse(p<.01,"***",ifelse(p<.05,"**",ifelse(p<.1,"*","")))
est <- function(m,v){ ct<-coeftable(m); sprintf("%+.4f%s", ct[v,1], star(ct[v,4])) }
sev <- function(m,v){ ct<-coeftable(m); sprintf("(%.4f)", ct[v,2]) }
vars <- c(ln_urban_density="ln(Urban Density)", net_migr_pct="Net Migration (\\% Pop)",
          ln_gdp_pc="ln(GDP per capita)", urban_pop_share="Urban Pop. Share")
# two physical rows per variable (estimate, then SE) so zebra shading fills cleanly
body <- do.call(rbind, lapply(names(vars), function(v) data.frame(
  Variables=c(vars[[v]], ""),
  `LCRPGR (log)`=c(est(mL,v), sev(mL,v)),
  `BpCR`        =c(est(mB,v), sev(mB,v)),
  check.names=FALSE, row.names=NULL)))
gof <- data.frame(Variables=c("Observations","Within R$^2$","Country FE","Period FE"),
                  `LCRPGR (log)`=c(format(nobs(mL),big.mark=","), sprintf("%.4f",r2(mL)["wr2"]),"Yes","Yes"),
                  `BpCR`        =c(format(nobs(mB),big.mark=","), sprintf("%.4f",r2(mB)["wr2"]),"Yes","Yes"),
                  check.names=FALSE)
nb <- nrow(body); nv <- length(vars)
coef_grey <- as.vector(sapply(which(seq_len(nv) %% 2 == 1), function(k) c(2*k-1, 2*k)))
gof_grey  <- nb + which(seq_len(nrow(gof)) %% 2 == 1)
knitr::kable(rbind(body,gof), format="latex", booktabs=TRUE, escape=FALSE, linesep="",
             align=c("l","c","c")) |>
  kableExtra::kable_styling(latex_options="HOLD_position", font_size=9) |>
  kableExtra::row_spec(c(coef_grey, gof_grey), background="#efefef") |>
  kableExtra::row_spec(nb, hline_after=TRUE) |>
  kableExtra::footnote(general=c("*** p<0.01, ** p<0.05, * p<0.1.",
                                 "Standard errors clustered by country, in parentheses."),
                       general_title="", footnote_as_chunk=FALSE, escape=FALSE, threeparttable=TRUE)
```

```{r r2-inline}
#| include: false
r2L <- round(r2(mL)["wr2"], 4); r2B <- round(r2(mB)["wr2"], 4)
r2ratio <- round(r2(mB)["wr2"] / r2(mL)["wr2"])
```

With the official LCRPGR as the dependent variable, the model is essentially
uninformative: the within $R^2$ is only `r sprintf("%.3f", r2L)` and not a
single control is statistically significant, because the metric's arithmetic
tail outliers swamp any systematic variation. Replacing the dependent variable
with BpCR, and changing nothing else, raises the within $R^2$ to
`r sprintf("%.3f", r2B)`, roughly `r r2ratio` times higher, and the coefficients
become significant and correctly signed: denser cities, richer countries, and
higher in-migration are all associated with less land per capita. This is the
clearest evidence for H1: the choice of metric, not the data, determines whether
the relationships can be detected at all.

## The Final Dynamic Model

```{r gmm-table}
#| label: tbl-gmm
#| results: asis
#| tbl-cap: "Final dynamic-panel model: FE stepwise build-up, Arellano-Bond (difference) and Blundell-Bond (system) GMM."
if (!requireNamespace("kableExtra", quietly = TRUE)) install.packages("kableExtra")
mc <- as.data.frame(gmm$main_combo, stringsAsFactors = FALSE)
names(mc)[1] <- "Variables"
ncoef <- 5L                                   # first 5 rows are coefficients
# split each coefficient row "coef<br>(se)" into two physical rows (est, se)
split_two <- function(row){
  est <- se <- row
  for (j in 2:length(row)){
    x <- row[[j]]
    if (grepl("<br>", x, fixed=TRUE)){
      est[[j]] <- sub("<br>.*$","",x); se[[j]] <- sub("^.*<br>","",x)
    } else { est[[j]] <- x; se[[j]] <- "" }
  }
  est[[1]] <- row[[1]]; se[[1]] <- ""
  rbind(est, se)
}
coef_long <- do.call(rbind, lapply(seq_len(ncoef), function(i) split_two(mc[i,])))
gof <- mc[(ncoef+1):nrow(mc), ]
tab <- rbind(coef_long, gof)
esc <- function(x) gsub("%", "\\%", x, fixed = TRUE)
names(tab) <- esc(names(tab)); tab[[1]] <- esc(tab[[1]])
nb <- nrow(coef_long)
coef_grey <- as.vector(sapply(which(seq_len(ncoef) %% 2 == 1), function(k) c(2*k-1, 2*k)))
gof_grey  <- nb + which(seq_len(nrow(gof)) %% 2 == 1)
kbl <- knitr::kable(tab, format="latex", booktabs=TRUE, escape=FALSE, linesep="",
             align=c("l", rep("c", ncol(tab)-1)), row.names=FALSE) |>
  kableExtra::kable_styling(latex_options=c("HOLD_position","scale_down")) |>
  kableExtra::row_spec(c(coef_grey, gof_grey), background="#efefef") |>
  kableExtra::row_spec(nb, hline_after=TRUE) |>
  kableExtra::column_spec(6, border_right=TRUE) |>
  kableExtra::add_header_above(c(" "=1, "Fixed Effects"=5, "GMM (Bias-Corrected)"=2)) |>
  kableExtra::add_header_above(c(" "=1, "Dependent variable: BpCR"=7))
x <- as.character(kbl)
x <- sub("(Period FE[^\n]*\\\\\\\\)", "\\1\n\\\\midrule", x)
cat(x)
```

\vspace{-1.0em}

\footnotesize
*Note:* \*\*\* p<0.01, \*\* p<0.05, \* p<0.1. Standard errors in parentheses: clustered by country in the fixed-effects columns; two-step GMM with the Windmeijer finite-sample correction [@windmeijer2005] in the Arellano-Bond and Blundell-Bond columns. The lagged dependent variable is instrumented by lags 2 to 5 of BpCR, with the instrument matrix collapsed.
\normalsize

Throughout, recall that the dependent variable BpCR is an annual logarithmic
growth rate of built-up area per capita, so a coefficient measures the change in
that growth rate, and the coefficients on the logged regressors are
semi-elasticities.

*Path-dependence (H5).* In the preferred Arellano-Bond estimate of @eq-dynamic, the lagged
term is large and highly significant, $\rho = 0.61$: about three-fifths of a
country's per-capita land-use growth in one period carries into the next, and more
than a third still remains two periods later ($0.61^2 \approx 0.37$).
The GMM estimate is nearly three times the fixed-effects estimate ($\rho = 0.23$),
exactly the direction and magnitude the Nickell bias predicts, which is why the
naive fixed-effects estimate understates persistence. Sprawl, once underway, is
sticky: the long-lived building stock, road network, and zoning entrench a
trajectory rather than resetting each period.

*Urban density (H2).* The coefficient on log urban density is $-0.052$ and highly
significant: holding income and migration fixed, a 10% higher urban density is
associated with roughly $0.005$ lower annual BpCR, which is sizeable against a
median BpCR of about $0.004$. This is the compact-city mechanism, where
already-dense settlement accommodates new residents by building up rather than
out.

*Income (H3).* Log GDP per capita enters negative and significant at $-0.014$:
within a country, a doubling of income is associated with about $0.010$ lower
annual BpCR. This is the paper's central substantive result: over time, as a country grows
richer, its per-capita land use falls, consistent with rising income funding
infill, vertical development, and tighter land regulation. We return to the
contrast with the cross-section in @sec-discussion.

*Net migration (H4).* Net migration is negative and marginally significant
($-0.0005$): each additional percentage point of net in-migration is associated
with slightly lower per-capita land use, consistent with in-migrants being
absorbed into the existing urban stock faster than the built-up footprint expands.

*Urban population share.* The level of urbanisation is insignificant once the
other drivers are controlled, indicating that what matters is the urban *form*
(how densely a country builds) rather than the *stage* of urbanisation (how urban
it already is).

The two estimators use 1,351 (Arellano-Bond) and 1,544 (Blundell-Bond)
observations and 15 versus 21 instruments, the one-period and moment-condition
differences anticipated in @sec-strategy. Both pass the diagnostic tests: the
Sargan/Hansen test does not reject the over-identifying restrictions and the AR(2)
test finds no second-order serial correlation (both $p>0.10$). The Blundell-Bond
estimates are close to Arellano-Bond in sign and significance, so the findings are
not an artefact of the chosen estimator.

## Heterogeneity by Development Group

The pooled model imposes a common slope on a very heterogeneous set of countries:
the response of urban land use to income need not be the same in Germany as in
Nigeria. @tbl-hetero relaxes this by estimating the static model of @eq-static
separately for the three United Nations development groups, on the same
estimation sample as @tbl-static. We use the static fixed-effects estimator rather
than GMM here because the subgroups (44 to 101 countries) are too small for the
deep-lag instruments to be reliable.

```{r tbl-hetero}
#| label: tbl-hetero
#| tbl-cap: "Heterogeneity: static two-way fixed-effects BpCR model by United Nations development group (same estimation sample as Table 4)."
suppressMessages({library(fixest); library(dplyr)})
sph <- panel |>
  filter(year <= 2020, !is.na(bpcr), !is.na(lcrpgr_log)) |>
  arrange(iso3, year) |> group_by(iso3) |> mutate(bpcr_L = lag(bpcr)) |> ungroup() |>
  filter(!is.na(ln_urban_density), !is.na(urban_pop_share), !is.na(ln_gdp_pc),
         !is.na(net_migr_pct), !is.na(bpcr_L), !is.na(income_group))
ctrl <- "ln_urban_density + urban_pop_share + ln_gdp_pc + net_migr_pct"
fit <- function(g) feols(as.formula(paste("bpcr ~", ctrl, "| iso3 + year")),
                         filter(sph, income_group == g), vcov = ~iso3)
m1 <- fit("Developed"); m2 <- fit("Developing"); m3 <- fit("LDC")
star <- function(p) ifelse(p<.01,"***",ifelse(p<.05,"**",ifelse(p<.1,"*","")))
est <- function(m,v){ ct<-coeftable(m); sprintf("%+.4f%s", ct[v,1], star(ct[v,4])) }
sev <- function(m,v){ ct<-coeftable(m); sprintf("(%.4f)", ct[v,2]) }
nC  <- function(g) as.character(n_distinct(filter(sph, income_group==g)$iso3))
vars <- c(ln_urban_density="ln(Urban Density)", net_migr_pct="Net Migration (\\% Pop)",
          ln_gdp_pc="ln(GDP per capita)", urban_pop_share="Urban Pop. Share")
body <- do.call(rbind, lapply(names(vars), function(v) data.frame(
  Variables=c(vars[[v]], ""),
  `Developed`  = c(est(m1,v), sev(m1,v)),
  `Developing` = c(est(m2,v), sev(m2,v)),
  `LDC`        = c(est(m3,v), sev(m3,v)),
  check.names=FALSE, row.names=NULL)))
gof <- data.frame(Variables=c("Observations","Within R$^2$","Countries","Country FE","Period FE"),
  `Developed`  = c(format(nobs(m1),big.mark=","), sprintf("%.4f",r2(m1)["wr2"]), nC("Developed"),"Yes","Yes"),
  `Developing` = c(format(nobs(m2),big.mark=","), sprintf("%.4f",r2(m2)["wr2"]), nC("Developing"),"Yes","Yes"),
  `LDC`        = c(format(nobs(m3),big.mark=","), sprintf("%.4f",r2(m3)["wr2"]), nC("LDC"),"Yes","Yes"),
  check.names=FALSE)
nb <- nrow(body); nv <- length(vars)
coef_grey <- as.vector(sapply(which(seq_len(nv) %% 2 == 1), function(k) c(2*k-1, 2*k)))
gof_grey  <- nb + which(seq_len(nrow(gof)) %% 2 == 1)
knitr::kable(rbind(body,gof), format="latex", booktabs=TRUE, escape=FALSE, linesep="",
             align=c("l","c","c","c")) |>
  kableExtra::kable_styling(latex_options="HOLD_position", font_size=9) |>
  kableExtra::row_spec(c(coef_grey, gof_grey), background="#efefef") |>
  kableExtra::row_spec(nb, hline_after=TRUE) |>
  kableExtra::add_header_above(c(" "=1, "Dependent variable: BpCR"=3)) |>
  kableExtra::footnote(general=c("*** p<0.01, ** p<0.05, * p<0.1.",
                                 "Standard errors are clustered at the country level, in parentheses, to robustify against potential serial correlation within subgroups."),
                       general_title="", footnote_as_chunk=FALSE, escape=FALSE, threeparttable=TRUE)
```

Three patterns stand out. First, the compact-city effect of density is present in
every group but strengthens markedly from the developed (about $-0.015$) through
the developing ($-0.018$) to the least-developed countries ($-0.031$): where
cities are still forming, denser settlement absorbs new residents most strongly.
Second, the income-densification result, the paper's central finding, carries its
expected negative sign in all three groups, so it is not a rich-country artefact;
it is estimated most precisely in the developed and developing groups (significant
at the 5\% level) and is larger but noisier in the small least-developed subsample
(significant only at the 10\% level). Third, net migration densifies in the
developed and developing groups but is insignificant, and wrongly signed, for the
least-developed countries, consistent with international migration mattering little
there relative to the internal rural-to-urban flows that our national measure
cannot capture.


## Hypotheses

Taken together, the results support all five hypotheses. The metric comparison
supports **H1** (BpCR is the more reliable, better-fitting metric); the dynamic
model supports **H2** (density reduces land per capita), **H3** (income reduces
it, within countries), **H4** (in-migration densifies, in the developed and developing groups), and
**H5** (per-capita land use is strongly path-dependent).

# Discussion {#sec-discussion}

## Interpretation and Relation to the Literature

The headline result is that, *within* a country, higher income is associated with
*less* land consumed per urban resident. At first glance this seems to contradict
the established literature, in which richer countries are typically the more
sprawled ones. The two findings are complementary rather than conflicting because
they operate on different margins. *Across* countries, and at a single point in
time, higher income has historically coincided with larger urban footprints,
through bigger dwellings and car-based mobility; this is the cross-sectional
pattern documented by @seto2011meta and @oueslati2015sprawl. *Within* a country
over time, however, growing richer is associated with densification, as rising
income funds infill and vertical construction, redevelopment, and tighter
land-use regulation. Our country fixed effects hold national characteristics
constant and so isolate this second, within-country margin, which cross-sectional
designs cannot recover. We therefore extend, rather than overturn, the drivers
literature: income and density still shape urban land use, but the sign of the
income relationship depends on whether one compares countries or follows a single
country over time. The development-group analysis in @tbl-hetero sharpens this
reading: the income relationship is negative in every group, but the compact-city
density channel is markedly stronger in the developing and least-developed
economies, where urban form is still being set, than in the already-dense
developed group.

The results also bear directly on the measurement debate. @nicolau2019lcrpgr argue
on formal grounds that the official LCRPGR ratio is unstable, and @Lu2026 show that
land-use-efficiency estimates are highly sensitive to the underlying data. Our
analysis supplies a complementary, empirical demonstration on a single harmonised
dataset: holding the data and specification fixed and changing only the dependent
variable, the within $R^2$ rises roughly twenty-fold when LCRPGR is replaced with
BpCR, and coefficients that are uninformative under LCRPGR become significant and
interpretable. The choice of metric, and not only the choice of data, can
determine the conclusions one draws about sustainable urbanisation.

## Limitations and Future Research

Several caveats apply. First, the estimates are associations, not causal effects.
The dynamic-panel GMM corrects the bias of the lagged dependent variable and the
country and period fixed effects absorb a large class of confounders, but the
design falls short of a research design that would identify causal effects, and
the controls may still be correlated with omitted determinants of land use.
Second, the panel is observed at five-year epochs, so it captures medium-run
trajectories rather than short-run adjustment. Third, the analysis area is defined
by the Degree of Urbanisation at a fixed threshold (GHS-SMOD of at least 21);
plausible alternative thresholds would shift the urban boundary and could affect
the levels, if not the within-country patterns. Fourth, the built-up and
population layers are satellite-based estimates and the economic controls are
national aggregates, both of which carry measurement error. In particular, net migration is measured at the national level and so reflects international migration only: it does not capture internal rural-to-urban migration, which is often the dominant demographic driver of urban expansion in developing countries, so the migration channel we estimate is likely understated. Finally, the pooled
specification imposes a common slope; the development-group analysis in
@tbl-hetero shows that the income and density relationships, although consistent in
sign, vary in strength across groups, and a finer treatment of cross-country
heterogeneity remains for future work. These limitations point to natural extensions:
a sub-national analysis exploiting the ADM1 detail already in the data, which would also allow internal rural-to-urban migration to be measured directly; designs
aimed at causal identification, and an assessment of BpCR as a practical,
stable complement to LCRPGR in official SDG 11.3.1 monitoring.

# Conclusion {#sec-conclusion}

Urban sprawl is real and measurable: across 1985--2020, built-up area per capita
rose in almost two-thirds of country-periods and in about two-thirds of
countries on average. The difficulty is not the phenomenon but the yardstick. The
official SDG 11.3.1 indicator, LCRPGR, is unstable by construction, dividing an
arithmetic numerator by a near-zero denominator and producing sign-flipping
outliers exactly where land-use efficiency matters most. We proposed a simple,
log-based alternative, the Built-up per Capita Rate, which is finite, symmetric,
and interpretable for every country-period and reproduces the policy signal
without the instability; on identical data, replacing LCRPGR with BpCR raises the
within-country explanatory power roughly twenty-fold.

Using the stable metric on a dynamic panel of 193 United Nations member states, we
find that per-capita urban land use is strongly path-dependent and that, within
countries, it declines with urban density, net in-migration, and income, so that
economic growth over time is accompanied by densification rather than sprawl. The
contribution is as much methodological as substantive: a stable metric paired with
a bias-corrected dynamic estimator. For monitoring sustainable urbanisation, the
choice of metric is not a technicality but a precondition for drawing the right
conclusions.

# References {.unnumbered}

::: {#refs}
:::
