
---
title: "Urban Land-Use Efficiency"
subtitle: "A Stable Metric for SDG 11.3.1, and Why Per-Capita Designs Cannot Cleanly Identify Its Drivers: 193 UN Member States x 1985-2020 Panel Analysis"
doc-type: "Seminar Paper"
course-info: "Development Economics · SoSe 2026"
lecturer: "Lecturer: Prof. Dr. Jürgen Bitzer"
author:
  - name: Ömer Furkan Çoban
    email: oemer.furkan.coban@uni-oldenburg.de
    affiliations:
      - name: Carl von Ossietzky University of Oldenburg
        department: Applied Economics and Data Science
        city: Oldenburg
        country: Germany
    note: |
      The author used the Claude Opus 4.8 and Gemini 3.5 Flash large language models for
      improving the writing and for coding assistance; all analysis, results, and conclusions
      are the author's own.
date: "June 26, 2026"
abstract: |
  The official SDG 11.3.1 indicator, the ratio of land consumption rate to
  population growth rate (LCRPGR), is unstable: it divides by a near-zero growth rate
  and mixes an arithmetic numerator with a logarithmic denominator. We adopt the
  Built-up per Capita Rate (BpCR), the log change in built-up area per person, a
  difference of two log growth rates that is finite wherever urban built-up and
  population are positive. On a
  harmonised panel of 193 UN member states (1985–2020) built from GHSL grids under
  the Degree of Urbanisation, we first show that, on identical data, no driver is
  detectable at the 5% level even under the ratio's conservative log-form benchmark,
  while the same design recovers precise associations once BpCR replaces it.

  Estimating the drivers of BpCR with
  two-way fixed-effects and dynamic-panel GMM, we find that built-up area per capita
  rose in 63% of country-periods (67% of countries on average) and that per-capita
  land use is clearly path-dependent, though its persistence is only bracketed, not
  point-identified, once the dynamic model's failing moment conditions are respected.
  The driver results are largely cautionary: in per-capita metrics of the BpCR family
  the compact-city density coefficient is a measurement artefact (urban population
  enters the metric negatively at $t$ and positively at $t-1$, so density inherits the
  sign of the date it is measured at); income is indistinguishable from zero; and the
  one association that survives every check, net in-migration, has a negative *sign*
  that is close to mechanical (migration drives urban population, which sits in the
  metric's denominator) but a *magnitude* that is behaviourally meaningful, the built
  stock adjusting to migrants at only about a quarter the rate of population. The
  contribution is thus primarily methodological, a stable metric and a demonstration
  that per-capita designs can manufacture the very drivers they appear to detect.
keywords:
  - SDG 11.3.1
  - Urban land-use efficiency
  - Built-up per capita rate (BpCR)
  - LCRPGR
  - Dynamic panel GMM
  - GHSL / Degree of Urbanisation
journal:
  name: Development Economics seminar working paper
format:
  elsevier-pdf:
    keep-tex: true
    template-partials:
      - before-body.tex
    include-in-header:
      text: |
        \usepackage{makecell}
        \usepackage{float}
        \geometry{top=1.45cm,bottom=1.45cm,left=1.8cm,right=1.8cm}
        \usepackage{placeins}
        \floatplacement{figure}{ht}
        \floatplacement{table}{!ht}
        \setlength{\floatsep}{6pt plus 2pt minus 2pt}
        \setlength{\textfloatsep}{7pt plus 2pt minus 2pt}
        \setlength{\intextsep}{6pt plus 2pt minus 2pt}
        \setlength{\bibsep}{0pt plus 0.5pt}
        \usepackage{multicol}
        \let\oldthebibliography\thebibliography
        \let\endoldthebibliography\endthebibliography
        \renewenvironment{thebibliography}[1]{\scriptsize\begin{multicols}{2}\raggedright\emergencystretch=2em\oldthebibliography{#1}}{\endoldthebibliography\end{multicols}}
    journal:
      formatting: preprint      # preprint | review | doubleblind
      model: 3p                 # 1p | 3p | 5p (3p = standard single-column-ish)
      cite-style: authoryear    # elsarticle author-year citations
  html:
    theme: [litera, theme.scss]
    embed-resources: true
    include-in-header:
      text: |
        <style>
        body { font-family: "Source Serif 4","Iowan Old Style",Palatino,Georgia,"Times New Roman",serif; line-height: 1.62; }
        h1,h2,h3,h4,h5,.title,.subtitle { font-family: "Source Sans 3","Source Sans Pro",-apple-system,"Segoe UI",Roboto,Helvetica,Arial,sans-serif; }
        p { text-align: justify; }
        nav#TOC > ul, nav[role="doc-toc"] > ul { display: block !important; }
        /* kableExtra writes the footnote into the first cell only (no colspan); lay
           each footnote row out as a full-width block so the note uses the full width */
        table tfoot tr { display: block !important; width: 100%; }
        table tfoot tr td { display: none !important; }
        table tfoot tr td:first-child { display: block !important; border: 0 !important; padding: .1rem 0 !important; font-size: .8rem; color: #4b5563; text-align: left !important; }
        </style>
    toc: true
    toc-depth: 2
    toc-title: Contents
    number-sections: true
    tbl-cap-location: top
    fig-cap-location: bottom
    smooth-scroll: true
    anchor-sections: true
    link-external-newwindow: true
    citations-hover: true
    code-tools: false
    format-links: false
    other-links:
      - text: Paper (PDF)
        href: paper.pdf
        icon: file-earmark-pdf
      - text: Supplementary (HTML)
        href: supplementary.html
        icon: file-earmark-text
      - text: Supplementary (PDF)
        href: supplementary.pdf
        icon: file-earmark-pdf
      - text: Presentation (slides)
        href: ../04_presentation/presentation.html
        icon: easel
      - text: Presentation (PDF)
        href: ../04_presentation/presentation.pdf
        icon: file-earmark-pdf
      - text: Replication Package
        href: https://github.com/ofurkancoban/UrbanLandUseEfficiency_DevEconResearchProj
        icon: file-earmark-zip
bibliography: references.bib
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
#| include: false
library(here)
library(dplyr)
library(readr)
library(fixest)
panel <- read_csv(here("03_datasets/processed/reg_panel_urban.csv"),
    show_col_types = FALSE
)
gmm <- readRDS(here("03_datasets/processed/dynamic_gmm.rds"))
# format-aware table output: LaTeX for the PDF, HTML for the web build
out_fmt <- if (knitr::is_latex_output()) "latex" else "html"
pc <- if (knitr::is_latex_output()) "\\%" else "%" # percent sign, escaped for LaTeX only
```

# Introduction

The world is urbanising rapidly: by 2050 about two in three people are projected to
live in urban areas [@undesa2018wup], and the accompanying land expansion is
striking, with global urban land cover forecast to grow by 1.2 million km$^2$
between 2000 and 2030, an area about the size of South Africa [@seto2012global].
This is not a developing-world phenomenon alone: the developed world is projected to
add a comparable amount of new urban land [@gao2020]. For development policy the
central question is not whether cities expand but how efficiently they do so, that
is, how much additional land each new urban resident consumes. When built-up area
grows in step with population cities stay compact; when it grows much faster they
sprawl, paving over farmland, entrenching car dependence, raising per-resident
infrastructure costs, and lifting per-capita emissions. Measuring land-use
efficiency well is therefore a precondition for monitoring sustainable urbanisation.

The United Nations monitors this through target 11.3 of its Sustainable Development
Goals, whose official indicator 11.3.1 is the ratio of the Land Consumption Rate to
the Population Growth Rate (LCRPGR) [@unhabitat2025metadata]. As the agreed global
yardstick, its statistical properties determine how every country's progress is
judged, yet it is fragile by construction: with the population-growth rate in the
denominator, the ratio diverges and flips sign whenever population is roughly
stable, so exactly where land-use efficiency matters most for policy it becomes
uninformative.

This paper makes three contributions. First, it documents, on a harmonised global
dataset, that LCRPGR is statistically unstable, with its instability concentrated
where population growth is near zero, and that the cost is not cosmetic: on
identical data and specification, the official ratio leaves every driver
undetectable at the 5% level while the log-based Built-up per Capita Rate (BpCR)
recovers precise ones. Second, using this stable metric on a dynamic panel of 193
United Nations member states (five-year intervals, 1985--2020; Arellano-Bond and
Blundell-Bond GMM against the Nickell bias a static reading ignores), it finds that
the behavioural drivers the design appears to identify are fragile once the metric's
arithmetic is taken seriously: income is a null and the one surviving association,
net in-migration, has a mechanical negative sign but a behaviourally informative
magnitude. Third, it shows that in
per-capita regressions of the BpCR family the negative density (compact-city)
coefficient is an artefact of the metric's own arithmetic, a spurious correlation of
the kind long known in statistics [@pearson1897; @kronmal1993] that arises because
density is built from the same population term the metric itself contains and so
inherits the sign of the date at which it is measured (@sec-strategy). This is a
caution for per-capita designs like ours, not a claim about the footprint-area
regressions on which much of the drivers literature rests.

Empirically, one further headline fact stands alongside these three contributions:
sprawl is the global norm rather than the exception, with built-up area per capita
rising in a clear majority of country-periods. Beyond that, the main lesson is
methodological, not substantive: per-capita land-use metrics can manufacture the
drivers they seem to detect, and the results above should be read as a caution to
that effect rather than as a catalogue of confirmed causes of sprawl.

# Measurement Framework {#sec-measurement}

This section sets out the conceptual core of the paper. We first define the
official SDG 11.3.1 indicator and show why it is statistically unstable, and we
then introduce the stable, log-based alternative used in the empirical analysis.

## The SDG 11.3.1 Indicator and Its Limitations

Following the UN-Habitat metadata [@unhabitat2025metadata], the official
indicator divides an arithmetic land consumption rate by a logarithmic
population growth rate:
$$
\text{LCRPGR}_i = \frac{\text{LCR}_i}{\text{PGR}_i},\qquad
\text{LCR}_i=\frac{V_{i,t}-V_{i,t-z}}{V_{i,t-z}}\cdot\frac{1}{z},\qquad
\text{PGR}_i=\frac{\ln(P_{i,t}/P_{i,t-z})}{z},
$$
where $V_{i,t}$ and $V_{i,t-z}$ are the urban built-up areas at the end and start
of the period, $P_{i,t}$ and $P_{i,t-z}$ the corresponding urban populations, $z$
the period length in years, and $\ln(\cdot)$ the natural logarithm. LCR is thus the
*arithmetic* annual growth of built-up area and PGR the *logarithmic* growth of
population; this arithmetic-LCR / logarithmic-PGR pairing is the metadata's own
operational specification, not our reading [@unhabitat2025metadata;
@corbane2017lue], and @nicolau2019lcrpgr show formally that mixing the two scales is
what destabilises the ratio. Two structural problems follow. **(1) Division by a
near-zero denominator:** when population is roughly stable ($\text{PGR}\approx 0$)
the ratio diverges towards $\pm\infty$ and flips sign, so ordinary demographic
differences produce wild, uninterpretable values. **(2) Arithmetic-versus-logarithmic
asymmetry:** the arithmetic numerator and logarithmic denominator make the ratio
non-symmetric, its tails dominated by an arithmetic artefact rather than real
land-use behaviour [@nicolau2019lcrpgr].

## A Stable Metric: The Built-up per Capita Rate (BpCR)

The UN-Habitat metadata itself concedes that the ratio is hard to interpret and
offers built-up area per capita ($V/P$) as a secondary indicator
[@unhabitat2025metadata], a level at one point in time. Building on the log-based
direction @Lu2026 advocate, we turn it into a trend by its annual growth in
logarithmic form:
$$
\text{BpCR}_{i,t}=\frac{1}{z}\ln\!\left(\frac{V_{i,t}/P_{i,t}}{V_{i,t-z}/P_{i,t-z}}\right)
= \text{LCR}_{i,\log}-\text{PGR}_{i,\log},
$$
where $\text{LCR}_{i,\log}=\frac{1}{z}\ln(V_{i,t}/V_{i,t-z})$ and
$\text{PGR}_{i,\log}=\frac{1}{z}\ln(P_{i,t}/P_{i,t-z})$ are the logarithmic growth
rates of built-up area and population. Here $\text{PGR}_{i,\log}$ is identical to
the PGR of the official indicator above, which is already logarithmic, whereas
$\text{LCR}_{i,\log}$ is the logarithmic counterpart of the arithmetic
$\text{LCR}_i$ in that indicator. Replacing the arithmetic LCR with its
logarithmic form is precisely what removes the arithmetic-versus-logarithmic
asymmetry that destabilises LCRPGR.

As the **difference of two logarithmic growth rates**, BpCR is finite wherever urban
built-up and population are positive, symmetric, additively decomposable, and
directly interpretable:
$\text{BpCR}>0$ is sprawl (more land per resident), $\text{BpCR}<0$ densification,
$\text{BpCR}=0$ land growing in step with population. It is a *descriptive* quantity;
the "efficiency" label it and the SDG target carry is normative, and lower BpCR is
not unambiguously better, since extreme densification has welfare costs of its own.
Where LCRPGR is well-behaved BpCR reproduces its policy signal, but the two are not
interchangeable: BpCR stays informative in the near-zero-growth region where the
ratio breaks down and, as @sec-results shows, recovers within-country relationships
the ratio's instability hides. In the empirical sections we benchmark it
against the *logarithmic* form of the ratio,
$\text{LCRPGR}_{\log}=\text{LCR}_{i,\log}/\text{PGR}_{i,\log}$ (*LCRPGR (log)*), a
deliberately conservative choice: already free of the arithmetic-versus-logarithmic
asymmetry, it differs from BpCR only in being a ratio, so any instability it shows
comes purely from the near-zero denominator; the official arithmetic ratio would be
more unstable still.

# Literature Review {#sec-literature}

The literature on urban land use splits into two strands. One asks what *drives*
urban expansion; the other scrutinises how land-use efficiency is *measured*. This
paper sits at their intersection.

## Drivers of Urban Land Expansion

@seto2011meta synthesise 326 remote-sensing studies (1970--2000) and document that
urban land has grown faster than urban population almost everywhere, driven mainly
by income and, in India and Africa, by population growth. @oueslati2015sprawl
estimate the determinants of sprawl on 282 European cities and find that rising
income and population both enlarge the footprint, with cities spreading even where
population is flat. Together these establish income and population as the leading
drivers, but they work across cities or countries at a single cross-section,
leaving open whether the same relationships hold *within* a country over time.

## Critiques of the Indicator: Formula and Data

@nicolau2019lcrpgr show formally that the SDG 11.3.1 formula is structurally
unstable and that a logarithmic, efficiency-based alternative is more robust.
@Lu2026 show that efficiency estimates swing widely with the choice among six
built-up and four population datasets and argue for a log-based framework, and
@zhong2025 similarly stress that data and methodological choices shape the
resulting picture. This strand establishes that the choice of metric, not only of
data, can change the conclusions drawn about sustainable urbanisation.

## Research Gap and Contribution

Our contribution sits at the intersection of the two strands. On measurement, we
adopt the specific log-based BpCR formulation @Lu2026 propose, in the spirit of the
log-based alternative @nicolau2019lcrpgr more generally recommend. On method, we make
it dynamic: a two-way fixed-effects panel with Arellano-Bond and Blundell-Bond GMM,
estimated on 193 United Nations member states over 1985--2020, that examines the
persistence of per-capita land use and probes its purported economic drivers, most of
which, as the Results section shows, turn out not to survive scrutiny.

# Data {#sec-data}

## Data Sources

The empirical analysis draws on three official, global providers, the same sources
used in international SDG monitoring. Built-up area, population, and the urban mask
come from the **Global Human Settlement Layer (GHSL)** of the European Commission's
Joint Research Centre [@pesaresi2024ghsl; @schiavina2023smod]: GHS-BUILT-S for
built-up surface, GHS-POP for population disaggregated onto it, and GHS-SMOD for the
Degree of Urbanisation (DEGURBA) classification that flags urban cells. We use the
five-year layers from 1975 to 2020, discarding the modelled 2025 and 2030
projections, with 1975 serving only as the base for the 1980 growth rate (so BpCR is
observed over 1980--2020 and, after the lagged dependent variable, the estimation
sample spans 1985--2020). Administrative units come from the **Global Administrative
Unit Layers (GAUL)** of the UN FAO [@gaul2024], giving UN-validated, politically
neutral boundaries. Socio-economic controls come from
the **United Nations**: GDP per capita in constant 2020 US dollars from UN SNAAMA
and international net migration as a share of population from UN DESA World Population
Prospects 2024 [@unsd2025snaama; @undesa2024wpp]. For the heterogeneity analysis we
classify countries into UN development groups (developed, developing, and
least-developed) using the UN M49 regions and the UN Committee for Development Policy
LDC list [@unsd2024m49]. @tbl-data summarises each dataset and the coverage used.

```{r}
#| label: tbl-data
#| tbl-cap: "Dataset specifications. GHSL grids use the World Mollweide projection (ESRI:54009); built-up and population at 100 m, the SMOD urban mask at 1 km. The Years column reports the coverage used in this study, not each source's native span: the GHSL products span 1975-2030 natively, of which we take 1975-2020 (the 1975 layers serving only as the base for the 1980 growth rate) and discard the modelled 2025 and 2030 projections; the UN series (SNAAMA 1970-2024, WPP 1950-2023) are read only over the epochs the panel needs (@sec-data)."
data.frame(
    Dataset = c(
        "GHS-BUILT-S R2023A", "GHS-POP R2023A", "GHS-SMOD R2023A",
        "GAUL 2024/2025", "UN SNAAMA", "UN DESA WPP 2024", "UN M49 / CDP"
    ),
    Provider = c("EC JRC", "EC JRC", "EC JRC", "FAO", "UN Stats", "UN DESA", "UN"),
    Type = c("Raster", "Raster", "Raster", "Vector", "Tabular", "Tabular", "Classification"),
    Resolution = c("100 m", "100 m", "1 km", "Levels 0-2", "Country-year", "Country-year", "Country"),
    CRS = c(
        "Mollweide (54009)", "Mollweide (54009)", "Mollweide (54009)",
        "WGS84 (EPSG:4326)", "--", "--", "--"
    ),
    Years = c(
        "1975-2020, 5-yr", "1975-2020, 5-yr", "1975-2020, 5-yr",
        "2024/2025", "1985-2020, 5-yr", "1980-2020, annual", "2024"
    ),
    Variable = c(
        "Built-up area (m2/cell)", "Population (persons/cell)",
        "Degree of Urbanisation (codes 10-30)", "Admin boundaries",
        "GDP p.c. (const. 2020 US$)", "Net migration (% of pop.)", "Development group"
    ),
    check.names = FALSE
) |>
    knitr::kable(
        format = out_fmt, booktabs = TRUE, escape = TRUE, linesep = "",
        align = c("l", "l", "l", "l", "l", "l", "l")
    ) |>
    kableExtra::kable_styling(latex_options = c("scale_down"))
```

## Constructing the Panel Following the UN Metadata

A central design choice is that we build the panel exactly as the SDG 11.3.1
metadata prescribes [@unhabitat2025metadata], so that our BpCR and the official
LCRPGR are computed on identical inputs and the comparison is like-for-like. The
construction follows the metadata stages.

@fig-layers illustrates the construction for one example window, with each layer
drawn over a satellite basemap. Panel (a) is the basemap for
reference; panel (b) shows the GHS-BUILT-S surface as the built-up share of each cell; panel
(c) the GHS-SMOD Degree of Urbanisation, which classifies each cell from rural to
urban centre; and panel (d) the urban-masked built-up that actually enters the
analysis, retaining built-up only in cells classified as urban (SMOD at least 21). The same masking is applied to population, so that both components are
measured on a consistent urban footprint.

*Analysis area.* SDG 11.3.1 is an urban (city-scale) indicator, measured inside
cities rather than over whole national territories. We delineate the urban area
with DEGURBA on the GHS-SMOD grid (a cell is urban when its settlement code is at
least 21, that is, urban centres and dense or semi-dense towns and suburbs) and
measure all built-up and population quantities only over these urban cells. The
built-up and population layers are native 100 m grids and the SMOD urban mask is a
1 km grid; we resolve the resolution mismatch by aggregating the 100 m layers to
the 1 km SMOD grid before masking, so that all three layers share a common cell.

![Building the analysis input following the SDG 11.3.1 metadata (Istanbul /
Bosphorus window), each layer drawn over an Esri satellite basemap. (a) Basemap.
(b) GHS-BUILT-S built-up surface (native 100 m, shown here as the built-up share
aggregated to 1 km cells for display, to match the 1 km SMOD grid). (c) GHS-SMOD
Degree of Urbanisation (1 km); cells coded 21 or higher count as urban.
(d) Urban-masked built-up, the input actually used: panel (b) kept only where
panel (c) is urban. The gold line is the GAUL country boundary.
](../04_outputs/figures/metadata_layers.png){#fig-layers width=74%}

*Components and aggregation.* The unit of analysis is the country, so for each
country-period we sum urban built-up area $V$ and urban population $P$ across cells
*before* forming any rate (the metadata's aggregate-then-rate rule): averaging
city-level ratios would let a small town with a near-zero denominator dominate the
national figure. Boundaries are held fixed at their current GAUL definition for
every epoch, so each country is a constant spatial footprint and the within-country
changes reflect land-use dynamics rather than shifting borders.

*Rates and controls.* From these national totals we form LCRPGR and BpCR as
defined in @sec-measurement. The two urban controls come from the same GHSL/SMOD
layers: **urban density** is urban population per urban-cell area, in logs (a
settlement density, discussed in @sec-strategy), and **urban population share** is
urban population divided by national total population. The remaining controls are international: **GDP per
capita** (constant 2020 US dollars, logs) from UN SNAAMA and **net migration**
(international, share of population) from UN DESA WPP 2024. All variables align to
the same country-period grid.

The resulting panel is unbalanced: it nominally spans 193 United Nations member
states, but not fully evenly. Sudan and South Sudan contribute only three of eight
epochs, part of the 37 GDP-missing country-periods discussed below (@sec-limitations),
and Tuvalu only one, DEGURBA flagging no urban centre there until recent decades; that
single-period case is the one the within estimator drops for lack of variation.
@tbl-flow traces how the record narrows
from the observed BpCR series to each estimator's sample. Two
points carry through the rest of the paper: the fixed-effects columns drop Tuvalu as
a single-period within-singleton (1,499 observations, 192 countries), while the GMM
columns keep all 193 countries and report genuinely non-zero rows (1,307 and 1,500),
net of the structural zero-padding `plm::pgmm` assigns to data-sparse countries
(@sec-strategy).

```{r}
#| label: tbl-flow
#| tbl-cap: "Sample flow, from the observed BpCR series to each estimator's working sample. Arellano-Bond and Blundell-Bond figures are genuinely non-zero rows, net of structural zero-padding for data-sparse countries (@sec-strategy)."
data.frame(
  Stage = c("Dependent variable eligible (1985-2020)", "Complete-case working sample",
            "Fixed-effects estimation", "Arellano-Bond (differenced equation)",
            "Blundell-Bond (level equation)", "Clean-timing common sample"),
  `Country-periods` = c("1,538", "1,500", "1,499", "1,307", "1,500", "1,336"),
  Countries = c("193", "193", "192", "193", "193", "--"),
  Note = c("1980 epoch is lag base only", "lag DV and four controls present",
           "Tuvalu singleton dropped", "of $193\\times7$ nominal",
           "of $193\\times8$ nominal", "t, t-1, t-2 all observed"),
  check.names = FALSE
) |>
  knitr::kable(format = out_fmt, booktabs = TRUE, escape = FALSE, linesep = "",
    align = c("l", "r", "r", "l")) |>
  kableExtra::kable_styling(font_size = 9)
```

# Descriptive Statistics {#sec-descriptive}

```{r desc-stats}
#| include: false
d <- subset(panel, !is.na(bpcr) & year >= 1985 & year <= 2020)
ds <- list(
    n      = nrow(d),
    lcr_sd = sd(d$lcrpgr_log, na.rm = TRUE),
    lcr_lo = min(d$lcrpgr_log, na.rm = TRUE),
    lcr_hi = max(d$lcrpgr_log, na.rm = TRUE),
    bp_sd  = sd(d$bpcr, na.rm = TRUE),
    ratio  = round(sd(d$lcrpgr_log, na.rm = TRUE) / sd(d$bpcr, na.rm = TRUE), -1),
    shr_cp = 100 * mean(d$bpcr > 0, na.rm = TRUE),
    med    = median(d$bpcr, na.rm = TRUE),
    shr_l1 = 100 * mean(d$lcrpgr > 1, na.rm = TRUE)
)
cm <- aggregate(bpcr ~ iso3, d, mean)
ds$ctry_pos <- sum(cm$bpcr > 0)
ds$ctry_tot <- nrow(cm)
ds$ctry_pct <- round(100 * ds$ctry_pos / ds$ctry_tot)
```

```{r}
#| label: tbl-desc
#| tbl-cap: "Summary statistics, 1985-2020 (country-periods with an observed BpCR). LCRPGR (log) is the log-form ratio benchmark; BpCR is the log growth of built-up area per capita."
dd <- subset(panel, !is.na(bpcr) & year >= 1985 & year <= 2020)
vars <- c(
    bpcr = "BpCR", lcrpgr_log = "LCRPGR (log)", ln_urban_density = "ln(Urban Density)",
    urban_pop_share = "Urban Pop. Share", ln_gdp_pc = "ln(GDP per capita)",
    net_migr_pct = paste0("Net Migration (", pc, " Pop)")
)
rows <- lapply(names(vars), function(v) {
    x <- dd[[v]]
    data.frame(
        Variable = vars[[v]], N = sum(!is.na(x)),
        Mean = mean(x, na.rm = TRUE), SD = sd(x, na.rm = TRUE),
        Min = min(x, na.rm = TRUE), Median = median(x, na.rm = TRUE), Max = max(x, na.rm = TRUE),
        check.names = FALSE
    )
})
tab <- do.call(rbind, rows)
num <- sapply(tab[, 3:7], function(c) ifelse(abs(c) >= 100, sprintf("%.0f", c), sprintf("%.3f", c)))
tab[, 3:7] <- num
knitr::kable(tab,
    format = out_fmt, booktabs = TRUE, escape = FALSE, linesep = "",
    align = c("l", "r", "r", "r", "r", "r", "r"), row.names = FALSE
) |>
    kableExtra::kable_styling(font_size = 9)
```

@tbl-desc reports summary statistics, and even before any regression the raw data
reveal why the metric matters. Over the `r format(ds$n, big.mark=",")`
country-periods, LCRPGR (log) has a standard deviation of about `r round(ds$lcr_sd)`
and ranges from roughly `r round(ds$lcr_lo)` to `r round(ds$lcr_hi)`, its near-zero
denominator producing enormous sign-flipping outliers, whereas BpCR stays in a
tight band (standard deviation about `r sprintf("%.4f", ds$bp_sd)`, roughly
`r format(ds$ratio, big.mark=",")` times narrower) on the same data. Sprawl is the
global norm in *direction*: BpCR is positive in `r sprintf("%.0f", ds$shr_cp)`% of
country-periods and `r ds$ctry_pos` of `r ds$ctry_tot` countries (`r ds$ctry_pct`%)
have a positive average, and the official arithmetic ratio exceeds its sprawl
threshold of one in `r sprintf("%.0f", ds$shr_l1)`% of cases. The typical *magnitude*
is nonetheless modest: the median country-period BpCR is
`r sprintf("%+.4f", ds$med)`, about `r sprintf("%.1f", 100*ds$med)`% more land per
resident per year, so sprawl is pervasive but slow. The qualitative message is the
same under either metric; BpCR simply delivers it without the instability.


```{r}
#| label: fig-trends
#| out-width: 56%
#| fig-cap: "Country trajectories for seven regional representatives, 1985--2020. (a) LCRPGR (log) swings erratically across its zero and unity reference lines; (b) BpCR is smooth on the same data, with the fastest-urbanising economies showing the strongest early sprawl."
knitr::include_graphics(
    if (knitr::is_latex_output()) {
        "../04_outputs/figures/country_trends.pdf"
    } else {
        "../04_outputs/figures/country_trends.png"
    }
)
```
Country trajectories make the contrast concrete (@fig-trends): LCRPGR (log) swings
erratically, crossing zero and spiking to large positive and negative values hard
to read as land-use behaviour, whereas the BpCR series for the same countries are
smooth and interpretable, most drifting gradually towards balance while the
fastest-urbanising economies show the strongest early sprawl.

# Empirical Strategy {#sec-strategy}

**Research question.** We ask two linked questions. The substantive one is what
drives *within-country* changes in per-capita urban land use over time, and how
persistent they are. The measurement one is whether the official SDG 11.3.1
indicator captures this reliably, or whether the choice of metric changes the
conclusions. The measurement question is settled by **H1**, a metric-property
comparison rather than a behavioural hypothesis: BpCR captures land-use efficiency
more reliably (a better statistical fit) than LCRPGR. The substantive question is
formalised in four testable behavioural hypotheses on the drivers of BpCR, each with
a sign expected a priori: **H2**, higher urban density lowers land per capita ($-$);
**H3**, higher income lowers it ($-$); **H4**, international net in-migration lowers
it ($-$); and **H5**, per-capita land use is path-dependent ($+$). We keep these
fixed labels throughout.

## Specification

We proceed in two stages, each with its own specification. In the **static** stage
we compare the two metrics on an equal footing. For each metric
$\text{Metric} \in \{\text{LCRPGR}_{\log}, \text{BpCR}\}$ (we use the conservative
log form of the ratio, @sec-measurement) we estimate a two-way fixed-effects model
without dynamics:
$$
\small \text{Metric}_{i,t} = \beta_1\ln(\text{UrbDens})_{i,t-1} + \beta_2\,\text{UrbShare}_{i,t-1} + \beta_3\ln(\text{GDPpc})_{i,t} + \beta_4\,\text{NetMigr}_{i,t} + \mu_i + \delta_t + \varepsilon_{i,t}.
$$ {#eq-static}
Because the data, controls, and fixed effects are held identical, any difference
between the LCRPGR (log) and BpCR estimates of @eq-static is attributable to the
metric alone (H1).

In the **dynamic** stage, having established BpCR as the usable metric, we add its
own lag to measure path-dependence, so the dependent variable is now specifically
BpCR:
$$
\small \text{BpCR}_{i,t} = \rho\,\text{BpCR}_{i,t-1} + \beta_1\ln(\text{UrbDens})_{i,t-1} + \beta_2\,\text{UrbShare}_{i,t-1} + \beta_3\ln(\text{GDPpc})_{i,t} + \beta_4\,\text{NetMigr}_{i,t} + \mu_i + \delta_t + \varepsilon_{i,t}.
$$ {#eq-dynamic}
The coefficient $\rho$ measures how strongly a country's per-capita land use
persists from one period to the next (H5).

The country fixed effects $\mu_i$ absorb all time-invariant national
characteristics, so identification comes from changes *within* a country over time,
not comparisons *between* countries; the period fixed effects $\delta_t$ absorb
global shocks common to all countries in an epoch. Standard errors are clustered by
country. Cluster-robust inference is reliable in the pooled models but less so in
the development-group splits (@sec-results), where the number of clusters is smaller
(44 to 101 countries) and only eight periods are observed.

Each regressor has a clear theoretical link to land consumed per resident.
**Urban density** captures the compact-city mechanism (H2), though ours is a
*settlement* density (population per urban-cell area, about ten times the built-up
surface), not the conventional built-up density, chosen to keep the dependent
variable's built-up term $V$ out of the control, a choice weighed in @sec-results.
**Urban population share** captures the stage of urbanisation. Testing either
requires care, because both are built from the urban population count that also sits
inside the dependent variable. Writing BpCR out in full,
$$
\text{BpCR}_{i,t}=\tfrac{1}{z}\big[\ln V_{i,t}-\ln P^{u}_{i,t}-\ln V_{i,t-1}+\ln P^{u}_{i,t-1}\big],
$$ {#eq-decomp}
urban population enters negatively at $t$ and positively at $t-1$, so either control
inherits the sign of the date at which it is measured: contemporaneous timing pushes
it toward a spurious *negative* (compact-city) sign, lagged timing toward a same-signed
spurious *positive*, and only $t-2$ shares no term with the *contemporaneous*
$\text{BpCR}_{i,t}$ (it does share one with the lagged dependent variable
$\text{BpCR}_{i,t-1}$ in the dynamic model, a collinearity that affects precision, not
the identification logic, since $\text{BpCR}_{i,t-1}$ is a regressor, not the
estimation target). We adopt $t-1$ for
the main specifications (keeping the sample intact and permitting the dynamic GMM) and
check every urban-form conclusion against the clean $t-2$ timing (@sec-results).
**GDP per capita** is ambiguous a priori: cross-country evidence links income to
sprawl, while within a country income can fund densification; our design tests
which force dominates (H3), a contrast developed in @sec-discussion. It shares only
*total* population with BpCR, through its per-capita denominator, a weak one-signed
entanglement that biases its coefficient toward positive and leaves any negative
estimate conservative. **Net migration**, measured nationally, captures
*international* migration and is expected to lower per-capita land if the built stock
adjusts more slowly than population to arriving residents (H4). A caveat symmetric to
the density one applies: migration *drives* population growth, which enters BpCR
negatively through the PGR term (@eq-decomp), so a mechanical channel runs from more
migration to lower BpCR. It is not the literal identity that afflicts density and
urban share, since the dependent variable uses *urban* population while the control
is *national* migration and built-up area does respond (the decomposition below);
but the channel is real, so we read the coefficient as a partly-mechanical
conditional association. Internal rural-to-urban migration is not observed
(@sec-discussion). Finally, the **lagged dependent variable** captures
path-dependence (H5). H1 concerns the metric and is addressed by comparing the
LCRPGR and BpCR runs.

## Dynamic-Panel Estimation and the Nickell Bias

Adding a lagged dependent variable to a fixed-effects model biases $\rho$
downward: the within transformation that removes $\mu_i$ correlates the transformed
lagged regressor with the transformed error, a bias of order $1/T$ that is
substantial with only eight periods. We correct it with GMM estimators for dynamic
panels. The **Arellano-Bond** [@arellano1991] (difference) estimator first-differences
away the fixed effects and instruments the differenced lagged term with deeper level
lags (lags two to five of BpCR); the **Blundell-Bond** [@blundell1998] (system)
estimator adds a levels equation instrumented by past differences, gaining efficiency
when the series is persistent. Only the lagged dependent variable is instrumented; the four controls
are treated as strictly exogenous, a simplification (income in particular is
plausibly jointly determined with land use), so the control coefficients are
conditional associations rather than clean exogenous effects (@sec-discussion). We
take **Arellano-Bond as preferred** and report Blundell-Bond alongside, both in two
steps with Windmeijer-corrected standard errors [@windmeijer2005] and a collapsed
instrument matrix (one column per lag distance).

The two estimators' nominal equation sizes, $N_{AB}=193\times7=1{,}351$ and
$N_{\text{level}}=193\times8=1{,}544$, are balanced-grid figures, not complete-case
totals: `plm::pgmm` zero-substitutes only unavailable deep-lag *instruments*
[@holtzeakin1988; @roodman2009], never a control, so any country-period missing one
is dropped before estimation, not zero-filled. That substitution, though, forces
every country into an identical nominal $T-k$ row count regardless of how much real
data it contributed, padding concentrated entirely in the 30 data-sparse countries
(chiefly the GDP-missing ones, plus the single-period Tuvalu). @tbl-flow and
@tbl-gmm report the genuinely non-zero counts directly instead: 1,307
(Arellano-Bond) and 1,500 (Blundell-Bond level), 97% of each nominal figure.
Blundell-Bond's is one row above the 192-country fixed-effects sample (1,499):
its level equation needs no within-country variation to form a valid row, so it
alone recovers the single-period Tuvalu case that the fixed-effects singleton
correction must drop, while every other country contributes identically to both.
Arellano-Bond's is lower, the extra period first-differencing consumes on top of
the lag.

Because the models are estimated in two steps, over-identifying restrictions are
assessed with the robust **Hansen** $J$ test, for which a $p$-value above 0.10 does
not reject instrument validity. Serial correlation is checked with the
Arellano-Bond **AR(1)** and **AR(2)** tests of the differenced residuals: by
construction AR(1) should be significant and AR(2) should not, the latter being the
condition under which the deeper lags are valid instruments. The collapsed
instrument set is also kept well below the number of countries (193) to avoid the
instrument-proliferation bias.

# Results {#sec-results}

## The Metric Matters: LCRPGR vs. BpCR

We first compare the two metrics in the static (lag-free) two-way fixed-effects
model of @eq-static, holding the data, controls, and fixed effects identical so any
difference is due to the dependent variable alone. To keep specifications
comparable, all models in this section (static and dynamic) use the common sample
with the lagged dependent variable and all four controls available, so every column
reports the same 1,499 observations and differences reflect the specification, not
the sample. The contrast between the metrics is stark (@tbl-static).

```{r static-fit}
#| include: false
suppressMessages({
    library(fixest)
    library(dplyr)
})
# same estimation sample as the dynamic model: complete cases with the lagged DV
sp <- panel |>
    filter(year <= 2020, !is.na(bpcr), !is.na(lcrpgr_log)) |>
    arrange(iso3, year) |>
    group_by(iso3) |>
    mutate(
        bpcr_L = lag(bpcr), ln_urban_density_L = lag(ln_urban_density),
        urban_pop_share_L = lag(urban_pop_share)
    ) |>
    ungroup() |>
    filter(
        !is.na(ln_urban_density), !is.na(urban_pop_share),
        !is.na(ln_gdp_pc), !is.na(net_migr_pct), !is.na(bpcr_L),
        !is.na(ln_urban_density_L), !is.na(urban_pop_share_L)
    )
ctrl <- "ln_urban_density_L + urban_pop_share_L + ln_gdp_pc + net_migr_pct"
mL <- feols(as.formula(paste("lcrpgr_log ~", ctrl, "| iso3 + year")), sp, vcov = ~iso3)
mB <- feols(as.formula(paste("bpcr ~", ctrl, "| iso3 + year")), sp, vcov = ~iso3)
r2L <- round(r2(mL)["wr2"], 4)
r2B <- round(r2(mB)["wr2"], 4)
# Detectability, not explained variance, is the comparable quantity across two
# different dependent variables: t-statistics are scale-free, R^2 shares are not.
tL <- coeftable(mL)
tB <- coeftable(mB)
maxtL <- max(abs(tL[, 3]))
tmigL <- tL["net_migr_pct", 3]
tmigB <- tB["net_migr_pct", 3]
```

With the official LCRPGR as the dependent variable, the model detects almost nothing
at conventional levels: no control is distinguishable from zero at the 5% level, the
largest $t$-statistic is `r sprintf("%.2f", maxtL)` (GDP, significant only at 10%),
and the point estimates are implausibly large and imprecise, because the metric's
tail outliers swamp any systematic variation. Replacing the dependent variable with
BpCR and changing nothing else inverts the picture: the *same* net-migration
variable, noise under the ratio ($t = `r sprintf("%.2f", tmigL)`$), becomes one of
the most precisely estimated coefficients in the paper ($t = `r sprintf("%.2f", tmigB)`$,
computed from the unrounded estimate and so not exactly reproducible from @tbl-static's
four-decimal entries). @tbl-static also reports the within $R^2$
(`r sprintf("%.3f", r2L)` against `r sprintf("%.3f", r2B)`), but we read the contrast
as one of **detectability, not explained variance**, the two being variances of
different dependent variables: holding data, controls, and fixed effects identical,
the ratio buries the economic signal in outlier noise while the log difference
recovers it. That is the evidence for H1.

```{r}
#| label: tbl-static
#| results: asis
#| tbl-cap: "Static two-way fixed-effects model on the estimation sample (1985-2020), same sample and controls for both metrics."
star <- function(p) ifelse(p < .01, "***", ifelse(p < .05, "**", ifelse(p < .1, "*", "")))
est <- function(m, v) {
    ct <- coeftable(m)
    sprintf("%+.4f%s", ct[v, 1], star(ct[v, 4]))
}
sev <- function(m, v) {
    ct <- coeftable(m)
    sprintf("(%.4f)", ct[v, 2])
}
vars <- c(
    ln_urban_density_L = "ln(Urban Density)[t−1]", net_migr_pct = paste0("Net Migration (", pc, " Pop)"),
    ln_gdp_pc = "ln(GDP per capita)", urban_pop_share_L = "Urban Pop. Share[t−1]"
)
# two physical rows per variable (estimate, then SE) so zebra shading fills cleanly
body <- do.call(rbind, lapply(names(vars), function(v) {
    data.frame(
        Variables = c(vars[[v]], ""),
        `LCRPGR (log)` = c(est(mL, v), sev(mL, v)),
        `BpCR` = c(est(mB, v), sev(mB, v)),
        check.names = FALSE, row.names = NULL
    )
}))
gof <- data.frame(
    Variables = c("Observations", "Within R²", "Country FE", "Period FE"),
    `LCRPGR (log)` = c(format(nobs(mL), big.mark = ","), sprintf("%.4f", r2(mL)["wr2"]), "Yes", "Yes"),
    `BpCR` = c(format(nobs(mB), big.mark = ","), sprintf("%.4f", r2(mB)["wr2"]), "Yes", "Yes"),
    check.names = FALSE
)
nb <- nrow(body)
nv <- length(vars)
coef_grey <- as.vector(sapply(which(seq_len(nv) %% 2 == 1), function(k) c(2 * k - 1, 2 * k)))
gof_grey <- nb + which(seq_len(nrow(gof)) %% 2 == 1)
notes_s <- c(
    "*** p<0.01, ** p<0.05, * p<0.1.",
    "Standard errors clustered by country, in parentheses."
)
k <- knitr::kable(rbind(body, gof),
    format = out_fmt, booktabs = TRUE, escape = FALSE, linesep = "",
    align = c("l", "c", "c")
) |>
    kableExtra::kable_styling(font_size = 9) |>
    kableExtra::row_spec(c(coef_grey, gof_grey), background = "#efefef") |>
    kableExtra::row_spec(nb, hline_after = TRUE)
if (out_fmt == "latex") {
    k <- k |> kableExtra::footnote(
        general = notes_s, general_title = "",
        footnote_as_chunk = FALSE, escape = FALSE, threeparttable = TRUE
    )
    cat(as.character(k))
} else {
    cat("```{=html}\n", as.character(k),
        "\n<p style=\"font-size:.8rem;color:#4b5563;margin:.3rem 0 1rem;text-align:left\">",
        gsub("<", "&lt;", paste(notes_s, collapse = " ")),
        "</p>\n```\n",
        sep = ""
    )
}
```

## The Final Dynamic Model

```{r}
#| label: tbl-gmm
#| results: asis
#| tbl-cap: "Final dynamic-panel model: FE stepwise build-up, Arellano-Bond (difference) and Blundell-Bond (system) GMM."
if (!requireNamespace("kableExtra", quietly = TRUE)) install.packages("kableExtra")
mc <- as.data.frame(gmm$main_combo, stringsAsFactors = FALSE)
names(mc)[1] <- "Variables"
ncoef <- 5L # first 5 rows are coefficients
# split each coefficient row "coef<br>(se)" into two physical rows (est, se)
split_two <- function(row) {
    est <- se <- row
    for (j in 2:length(row)) {
        x <- row[[j]]
        if (grepl("<br>", x, fixed = TRUE)) {
            est[[j]] <- sub("<br>.*$", "", x)
            se[[j]] <- sub("^.*<br>", "", x)
        } else {
            est[[j]] <- x
            se[[j]] <- ""
        }
    }
    est[[1]] <- row[[1]]
    se[[1]] <- ""
    rbind(est, se)
}
coef_long <- do.call(rbind, lapply(seq_len(ncoef), function(i) split_two(mc[i, ])))
gof <- mc[(ncoef + 1):nrow(mc), ]
tab <- rbind(coef_long, gof)
# escape % for LaTeX only; HTML needs the literal character
esc <- if (knitr::is_latex_output()) {
    function(x) gsub("%", "\\%", x, fixed = TRUE)
} else {
    identity
}
names(tab) <- esc(names(tab))
tab[[1]] <- esc(tab[[1]])
nb <- nrow(coef_long)
coef_grey <- as.vector(sapply(which(seq_len(ncoef) %% 2 == 1), function(k) c(2 * k - 1, 2 * k)))
gof_grey <- nb + which(seq_len(nrow(gof)) %% 2 == 1)
notes_gmm <- c(
    "*** p<0.01, ** p<0.05, * p<0.1.",
    "Standard errors in parentheses: clustered by country in the fixed-effects columns; two-step GMM with the Windmeijer correction (Windmeijer, 2005) in the GMM columns, where the lagged dependent variable is instrumented by lags 2 to 5 of BpCR with a collapsed instrument matrix."
)
kbl <- knitr::kable(tab,
    format = out_fmt, booktabs = TRUE, escape = FALSE, linesep = "",
    align = c("l", rep("c", ncol(tab) - 1)), row.names = FALSE
) |>
    kableExtra::kable_styling(latex_options = c("scale_down"), font_size = 8) |>
    kableExtra::row_spec(c(coef_grey, gof_grey), background = "#efefef") |>
    kableExtra::row_spec(nb, hline_after = TRUE) |>
    kableExtra::column_spec(6, border_right = TRUE) |>
    kableExtra::add_header_above(c(" " = 1, "Fixed Effects" = 5, "GMM (Bias-Corrected)" = 2)) |>
    kableExtra::add_header_above(c(" " = 1, "Dependent variable: BpCR" = 7))
if (knitr::is_latex_output()) {
    kbl <- kbl |> kableExtra::footnote(
        general = notes_gmm, general_title = "",
        footnote_as_chunk = FALSE, escape = FALSE, threeparttable = TRUE
    )
    x <- as.character(kbl)
    x <- sub("(Period FE[^\n]*\\\\\\\\)", "\\1\n\\\\midrule", x) # LaTeX-only mid-rule
    cat(x)
} else {
    x <- as.character(kbl)
    cat("```{=html}\n", x,
        "\n<p style=\"font-size:.8rem;color:#4b5563;margin:.3rem 0 1rem;text-align:left\">",
        gsub("<", "&lt;", paste(notes_gmm, collapse = " ")),
        "</p>\n```\n",
        sep = ""
    )
}
```

*Path-dependence (H5).* The lagged term is positive and significant in every
estimator, so per-capita land use is unambiguously path-dependent; its precise
magnitude, though, we do not pin down, and report as a range rather than a point.
The @bond2002 bracket bounds a consistent $\rho$ between the downward-biased within
estimate ($`r sprintf("%.2f", gmm$rho_check$fe)`$) and the upward-biased pooled OLS
($`r sprintf("%.2f", gmm$rho_check$ols)`$). The estimates within it disagree
informatively: the full-model difference GMM ($`r sprintf("%.2f", gmm$rho_check$ab)`$,
@tbl-gmm) even falls below the lower bound, but its Hansen test fails, so we do not
read it. Removing the density and urban-share controls, which are arithmetically
entangled with BpCR (@eq-decomp) and contaminate $\rho$ in the differenced equation,
restores moment validity and lifts the estimate: @tbl-rholadder puts it between
$0.34$ and $0.46$ across the specifications whose Hansen tests clear the threshold,
close to the value the @nickell1981 correction implies
($`r sprintf("%.2f", gmm$rho_check$implied_T8)`$); the first-stage $F$ of
`r sprintf("%.1f", gmm$rho_check$first_stage_F)` does not indicate a weak-instrument
problem. (The system estimator lands nearby, at $0.45$, but since its own Hansen test
fails we treat that as coincidence, not corroboration.) We conclude that persistence
is substantial, plausibly a third to a half of one period's growth, but decline to
report a single coefficient, since the full-model moment failure is exactly why the
data do not pin one down.

```{r}
#| label: tbl-rholadder
#| tbl-cap: "Persistence and its diagnostics as controls enter (difference GMM). Each row adds controls to the same AR(1) dynamic model on the same sample; Arellano-Bond standard errors in parentheses. The Bond (2002) bracket is [FE within, pooled OLS], and the last column is the Arellano-Bond Hansen over-identification p-value. The clean specifications (through income) clear the test and lie inside the bracket; only the arithmetically entangled density and urban-share controls (bottom row) push the estimate below the bracket, and neither this row's Hansen test nor Blundell-Bond's clears the conventional threshold. The full row is the model whose control coefficients @tbl-gmm reports."
rl <- as.data.frame(gmm$rho_ladder, stringsAsFactors = FALSE, check.names = FALSE)
rl <- rl[, c("Specification", "Pooled OLS", "FE (within)", "Arellano-Bond", "AB in bracket", "AB Hansen (p)")]
knitr::kable(rl,
    format = out_fmt, booktabs = TRUE, escape = TRUE, linesep = "",
    align = c("l", "r", "r", "r", "c", "r"), row.names = FALSE
) |>
    kableExtra::kable_styling(latex_options = c("scale_down"), font_size = 9) |>
    kableExtra::row_spec(4, background = "#efefef")
```

*Net migration (H4).* Of the four drivers, net in-migration is the only one whose
association is not overturned as an artefact. We read its coefficient primarily from
**fixed effects** ($-0.0064$), immune to the Hansen concerns above since it needs no
instruments; Arellano-Bond corroborates closely ($-0.0062$) despite its own marginal
Hansen test, and Blundell-Bond, whose Hansen test fails outright, lands nearby
($-0.0052$, @tbl-gmm) but carries no independent weight here. The coefficient is
negative, highly significant, and stable across development groups (@tbl-hetero) and
timings (@tbl-timingfe). Its interpretation
needs care in two directions. The negative *sign* is close to mechanical: net
migration is itself a strong driver of urban population, which enters BpCR negatively
(@eq-decomp), so once construction lags population over the five-year window a
negative coefficient follows almost by construction. But the result is not *merely*
mechanical, and we do not dismiss it as such. The decomposition^[The three figures are
the net-migration coefficient from one and the same two-way fixed-effects
specification (identical controls, fixed effects, and sample) estimated on each
additive component of BpCR in turn, the population growth rate PGR(log), the land
growth rate LCR(log), and BpCR itself; since
$\text{BpCR}=\text{LCR}_{\log}-\text{PGR}_{\log}$ by construction, the identity carries
to the coefficients ($0.24-0.88=-0.64$, in percentage points a year, i.e. 100 times
the raw $-0.0064$ of @tbl-gmm).] shows a one-point rise in net in-migration going with
urban population growth higher by about
`r sprintf("%.2f", 100*gmm$mig_decomp$pgr[[1]])` percentage points a year *and*
built-up growth higher by about `r sprintf("%.2f", 100*gmm$mig_decomp$lcr[[1]])`, so
per-capita land falls by the difference
($`r sprintf("%.2f", 100*gmm$mig_decomp$bpcr[[1]])`$). That built-up response is a
genuine behavioural elasticity, not part of any identity, and its size relative to
population, cities building for migrants at only about a quarter the rate over five
years, is the informative, behaviourally meaningful content of the result, a
structural fact about how slowly the built stock adjusts. What the negative
coefficient alone does *not* license is reading it as evidence that migration
triggers deliberate densification, a claim H4 never made and this design cannot test.
Migrants also plausibly select into fast-building economies, so treating
migration as strictly exogenous is a genuine simplification, but two checks bound its
cost. It passes a Wooldridge lead test (future migration does not predict current
BpCR, $p=`r sprintf("%.2f", gmm$mig_exog$lead[["Pr(>|t|)"]])`$), and relaxing the
assumption barely moves the estimate: treated as *predetermined* and instrumented by
its own lags it returns
$`r sprintf("%.4f", gmm$mig_exog$predeterm$coef[["Estimate"]])`$ against the baseline
$-0.0062$, and only the fully *endogenous* treatment, resting on deep lags of a short
panel, attenuates it (to
$`r sprintf("%.4f", gmm$mig_exog$endo$coef[["Estimate"]])`$, insignificant). Strict
exogeneity is thus a simplification, not a material source of bias here; because
selection still cannot be ruled out, we report the coefficient as a conditional
association, not a causal effect. It is, moreover, only the *international* channel:
the internal rural-to-urban migration that dominates urban growth in developing and
least-developed countries is unobserved here (@sec-discussion), so even this
association captures the smaller part of the migration story.

*Urban density (H2).* Entered at $t-1$, density is *positive* under fixed effects
($+0.014$, significant) and Arellano-Bond ($+0.037$, significant) but essentially zero
under Blundell-Bond ($-0.001$, insignificant), so the estimators disagree and density
is not robustly identified in the GMM; we do not rely on its magnitude. (Its $t-1$
value is $+0.009$, insignificant, on the smaller common-timing sample of
@tbl-timingfe, against $+0.014$ on the full 1,499; both small and positive.) The
*negative* compact-city coefficient appears only under contemporaneous timing, a
spurious sign from @eq-decomp; the arithmetically clean $t-2$ coefficient is $+0.011$
($p<0.05$), which is a weak but genuine result, past *settlement* density goes with
slightly *more* land per capita, the opposite of the compact-city direction. The $t-1$
and $t-2$ estimates are themselves statistically indistinguishable from one another, so
we do not read $t-1$ as showing the theory's predicted contamination *in excess of* a
clean baseline; @eq-decomp predicts only that $t-1$'s bias runs the same, non-negative
direction as the clean estimate, not a specific magnitude above it, and the sharply
identified part of the evidence is the large sign reversal between $t$ and the lagged
timings, not a $t-1$-versus-$t-2$ comparison. That
control is settlement density (population per urban-cell area), though, not the
*built-up* density the compact-city mechanism turns on, which shares *both* $V$ and
the population term of BpCR, is even more entangled with it (@eq-decomp), and so is
not cleanly identifiable in a per-capita design. We therefore neither confirm nor
refute the built-up compact-city mechanism; we show that the negative coefficient
routinely reported for it is a timing artefact, and that the one clean density
estimate available points, weakly, against it. @tbl-timingfe also shows net migration
immovable across all three timings ($-0.006$ throughout), the mark of a control that
shares no growth term with the dependent variable.

```{r}
#| label: tbl-timingfe
#| tbl-cap: "Clean-timing check (two-way fixed effects, one common sample): the two population-sharing controls entered contemporaneously (t), lagged (t-1), and double-lagged (t-2). The t-2 double lag costs each country a further epoch, so the common sample is 1,336 country-periods, not the main model's 1,499. Only the t-2 column shares no term with the dependent variable (@eq-decomp) and is therefore the arithmetically clean reading. Urban density flips from a spurious negative at t to a small positive at the clean t-2, while net migration is immovable across all three timings. Estimate with standard error in parentheses."
tf <- as.data.frame(gmm$timing_fe, stringsAsFactors = FALSE, check.names = FALSE)
tf[-1] <- lapply(tf[-1], function(x) gsub("<br>", " ", x, fixed = TRUE))
if (knitr::is_latex_output()) tf[] <- lapply(tf, function(x) gsub("%", "\\%", x, fixed = TRUE))
knitr::kable(tf,
    format = out_fmt, booktabs = TRUE, escape = FALSE, linesep = "",
    align = c("l", "c", "c", "c"),
    col.names = c("", "Both $t$", "Both $t\\text{-}1$", "Both $t\\text{-}2$ (clean)"),
    row.names = FALSE
) |>
    kableExtra::kable_styling(latex_options = "scale_down", font_size = 9) |>
    kableExtra::row_spec(2, background = "#efefef") |>
    kableExtra::add_header_above(c(" " = 1, "Timing of density and urban share" = 3))
```

*Income (H3).* Log GDP per capita carries its expected negative sign in every
specification but is **not** significant in the preferred column ($-0.006$,
$p = 0.17$) or elsewhere; the sign is stable, the precision is not. We therefore do
**not** claim H3 is supported and read the sign as suggestive only, noting that part
of its apparent significance in earlier single-year-migration versions was migration
variation, absorbed once the flow is window-matched (@sec-data).

*Urban population share.* Once lagged, the stage of urbanisation is insignificant, so
what matters is the urban *form* a country builds, not how far along the transition
it already is.

A word on the over-identification tests is in order, because in the *full* model
they are not comfortable: with only 15 and 21 instruments (so this is not instrument
proliferation) neither clears the conventional threshold, the Arellano-Bond Hansen
$J$ at $p=0.05$ and Blundell-Bond at $p=0.01$. We do not paper over it: a failing
Hansen
test leaves the over-identifying restrictions unvalidated, so the *full-model* GMM
estimates, $\rho$ in particular, cannot be read as consistent, and we do not
interpret them. Nothing load-bearing rests on those columns. The four control
coefficients are read from the **fixed-effects columns**, which need no instrument
validity; Arellano-Bond agrees with them in sign throughout, and in significance for
the two coefficients that are significant at all, density and migration. Only
migration, though, also matches closely in magnitude ($-0.0064$ vs $-0.0062$,
@tbl-gmm); density's magnitude diverges sharply between the two estimators (as
already noted above), and income's and urban share's estimates, being insignificant
in both columns, are too imprecise for their (also divergent) magnitudes to carry
much weight either way. GMM enters only to bound the Nickell bias on
$\rho$, which we report as a range (@tbl-rholadder), never from the full model but
from the moment-valid specifications whose own Hansen tests clear the threshold; even
there we decline a point estimate, for the reasons given above. AR(2) holds
throughout ($p=0.76$).

## Heterogeneity by Development Group


```{r}
#| label: tbl-hetero
#| results: asis
#| tbl-cap: "Heterogeneity: static two-way fixed-effects BpCR model by United Nations development group (same estimation sample as the static metric comparison above). The density and urban-share coefficients are entered at t-1 and carry the timing artefact of @eq-decomp; they are shown for completeness but not interpreted (the large LDC urban-share value collapses to insignificance at the clean t-2 timing). Only the net-migration row is read from this table."
suppressMessages({
    library(fixest)
    library(dplyr)
})
sph <- panel |>
    filter(year <= 2020, !is.na(bpcr), !is.na(lcrpgr_log)) |>
    arrange(iso3, year) |>
    group_by(iso3) |>
    mutate(
        bpcr_L = lag(bpcr), ln_urban_density_L = lag(ln_urban_density),
        urban_pop_share_L = lag(urban_pop_share)
    ) |>
    ungroup() |>
    filter(
        !is.na(ln_urban_density), !is.na(urban_pop_share), !is.na(ln_gdp_pc),
        !is.na(net_migr_pct), !is.na(bpcr_L), !is.na(ln_urban_density_L),
        !is.na(urban_pop_share_L), !is.na(income_group)
    )
ctrl <- "ln_urban_density_L + urban_pop_share_L + ln_gdp_pc + net_migr_pct"
fit <- function(g) {
    feols(as.formula(paste("bpcr ~", ctrl, "| iso3 + year")),
        filter(sph, income_group == g),
        vcov = ~iso3
    )
}
m1 <- fit("Developed")
m2 <- fit("Developing")
m3 <- fit("LDC")
star <- function(p) ifelse(p < .01, "***", ifelse(p < .05, "**", ifelse(p < .1, "*", "")))
est <- function(m, v) {
    ct <- coeftable(m)
    sprintf("%+.4f%s", ct[v, 1], star(ct[v, 4]))
}
sev <- function(m, v) {
    ct <- coeftable(m)
    sprintf("(%.4f)", ct[v, 2])
}
nC <- function(g) as.character(n_distinct(filter(sph, income_group == g)$iso3))
vars <- c(
    ln_urban_density_L = "ln(Urban Density)[t−1]", net_migr_pct = paste0("Net Migration (", pc, " Pop)"),
    ln_gdp_pc = "ln(GDP per capita)", urban_pop_share_L = "Urban Pop. Share[t−1]"
)
body <- do.call(rbind, lapply(names(vars), function(v) {
    data.frame(
        Variables = c(vars[[v]], ""),
        `Developed` = c(est(m1, v), sev(m1, v)),
        `Developing` = c(est(m2, v), sev(m2, v)),
        `LDC` = c(est(m3, v), sev(m3, v)),
        check.names = FALSE, row.names = NULL
    )
}))
gof <- data.frame(
    Variables = c("Observations", "Within R²", "Countries", "Country FE", "Period FE"),
    `Developed` = c(format(nobs(m1), big.mark = ","), sprintf("%.4f", r2(m1)["wr2"]), nC("Developed"), "Yes", "Yes"),
    `Developing` = c(format(nobs(m2), big.mark = ","), sprintf("%.4f", r2(m2)["wr2"]), nC("Developing"), "Yes", "Yes"),
    `LDC` = c(format(nobs(m3), big.mark = ","), sprintf("%.4f", r2(m3)["wr2"]), nC("LDC"), "Yes", "Yes"),
    check.names = FALSE
)
nb <- nrow(body)
nv <- length(vars)
coef_grey <- as.vector(sapply(which(seq_len(nv) %% 2 == 1), function(k) c(2 * k - 1, 2 * k)))
gof_grey <- nb + which(seq_len(nrow(gof)) %% 2 == 1)
notes_h <- c(
    "*** p<0.01, ** p<0.05, * p<0.1.",
    "Standard errors are clustered at the country level, in parentheses, to robustify against potential serial correlation within subgroups."
)
k <- knitr::kable(rbind(body, gof),
    format = out_fmt, booktabs = TRUE, escape = FALSE, linesep = "",
    align = c("l", "c", "c", "c")
) |>
    kableExtra::kable_styling(font_size = 9) |>
    kableExtra::row_spec(c(coef_grey, gof_grey), background = "#efefef") |>
    kableExtra::row_spec(nb, hline_after = TRUE) |>
    kableExtra::add_header_above(c(" " = 1, "Dependent variable: BpCR" = 3))
if (out_fmt == "latex") {
    k <- k |> kableExtra::footnote(
        general = notes_h, general_title = "",
        footnote_as_chunk = FALSE, escape = FALSE, threeparttable = TRUE
    )
    cat(as.character(k))
} else {
    # HTML: note as a full-width paragraph below the table (not a table cell)
    cat("```{=html}\n", as.character(k),
        "\n<p style=\"font-size:.8rem;color:#4b5563;margin:.3rem 0 1rem;text-align:left\">",
        gsub("<", "&lt;", paste(notes_h, collapse = " ")),
        "</p>\n```\n",
        sep = ""
    )
}
```

The pooled model imposes a common slope on a heterogeneous set of countries.
@tbl-hetero relaxes this by estimating the static model of @eq-static separately for
the three UN development groups on the same sample as @tbl-static, using the static
fixed-effects estimator because the subgroups (44 to 101 countries) are too small
for the deep-lag GMM instruments to be reliable.

Three patterns stand out. First, and most important, the net-migration association
is negative, highly significant, and near-identical across all three groups (about
$-0.0065$ in each, $p<0.01$), so the headline result is not confined to any
development stage. Second, income carries its expected negative sign in every group
but is not significant in any of them, consistent with the null in the pooled
model. Third, the urban-form coefficients (density and urban share) in this table are
entered at $t-1$ and so carry the same timing artefact as the main model (@eq-decomp,
@tbl-timingfe): we show them for completeness but do not interpret them, and read
only migration from the table. The starkest illustration is the large, significant
urban-share coefficient among the least-developed countries ($+0.112$); re-estimated
at the clean $t-2$ timing on that sub-sample it falls to $+0.045$ and loses
significance ($p=0.16$), exactly the collapse the artefact predicts, so it is a
measurement artefact, not a genuine least-developed-country effect, and none of the
urban-form terms reproduces the compact-city negative sign.

## Hypotheses

The metric comparison supports **H1** (BpCR is the more reliable, better-fitting
metric), and the dynamic model supports **H5** in direction (per-capita land use is
path-dependent), though the magnitude of persistence is only bracketed, not
point-identified (@sec-results).
**H4** (international in-migration is associated with less land per capita) is
supported empirically, but only partially as originally intended: it is the one
association that survives every estimator, group, and timing, yet its negative
*sign* is close to mechanical (migration drives urban population, which enters BpCR
negatively), so it is not evidence of deliberate densification the way H4 was framed;
its *magnitude*, the slower five-year adjustment of the built stock relative to
population, is nonetheless a genuine behavioural quantity (@sec-results). The other
two hypotheses are not supported. **H2** (density reduces
land per capita): the negative compact-city coefficient conventionally reported is a
measurement artefact (@eq-decomp), turning positive and significant at the clean $t-2$
timing, so the one clean estimate available points weakly *against* the compact-city
direction rather than merely being silent on it; because the proper compact-city
variable (built-up density) is even more entangled, though, we read this as tentative
rather than as refuting compact cities outright.
**H3** (income reduces land per capita) carries a
consistently negative sign but cannot be distinguished from zero, so we read it as
suggestive only.

# Discussion {#sec-discussion}

## Interpretation and Relation to the Literature

Our results are, above all, a caution about reading behavioural drivers off
per-capita metrics. @seto2011meta and @oueslati2015sprawl identify income and
population growth as the leading drivers of urban expansion; on BpCR, once its
arithmetic is respected, none survives as a clean *sign*-identified behavioural
driver. Migration, the only association left standing, has a negative sign that is
close to mechanical, because migration drives urban population, which enters BpCR
negatively, so a negative coefficient follows almost by construction once the built
stock lags population. Its magnitude, however, is a genuine behavioural quantity: the
decomposition shows built-up area grows with migration at only about a quarter the
rate of population, a real measure of how slowly land tracks people (@sec-results). Income is a null, negative everywhere but
indistinguishable from zero, in part because measuring migration over the dependent
variable's five-year window absorbs variation that earlier single-year specifications
left for income to pick up (@sec-data).

The urban-form finding is a methodological caution (@sec-results): any per-capita
regression controlling for density at the dependent variable's own date recovers a
negative compact-city coefficient whether or not the mechanism exists, because
population enters BpCR with opposite signs at $t$ and $t-1$ (@eq-decomp); the
mechanism proper, resting on the still more entangled built-up density, is not
identifiable this way. On measurement, where @nicolau2019lcrpgr argue formally that
LCRPGR is unstable and @Lu2026 show its estimates are data-sensitive, our
single-dataset comparison (@tbl-static) adds that the choice of metric, not only of
data, can determine the conclusions drawn about sustainable urbanisation.

## Limitations and Future Research {#sec-limitations}

Several caveats apply. First, the estimates are associations, not causal effects:
the fixed effects absorb a large class of confounders and the GMM corrects the
lagged-DV bias, but the controls are treated as strictly exogenous. For migration
this simplification is bounded (@sec-results): it passes a lead test and is
invariant to a predetermined treatment, yet selection into fast-building economies
cannot be fully ruled out, so we read it as a conditional association. For density
the within/first-difference divergence is consistent with a violation of strict
exogeneity (though measurement error could also contribute), which is why we read
its sign from the clean $t-2$ timing rather than interpret its GMM magnitude. Income
we do not interpret in any case, so its exogeneity is moot. Second, the five-year
epochs capture
medium-run trajectories rather than short-run adjustment. Third, and more seriously,
the urban area uses a fixed threshold (SMOD at least 21) on the *multitemporal* mask,
so a cell's urban membership is recomputed each epoch: when a low-density fringe cell
first crosses the threshold, its land and residents enter the urban set together,
mechanically raising measured per-capita land at the margin. This composition effect
can inflate the descriptive "sprawl is the norm" finding; it bears less on the
within-country coefficients, identified net of country and period fixed effects,
though it is not fully absorbed by them. A fixed final-epoch mask and alternative
thresholds are the natural robustness reanalyses and the main measurement extension
of this work.
Fourth, and most consequentially, our net-migration series is national and captures
*international* migration only, whereas the dominant engine of urban growth and
sprawl, especially in developing and least-developed countries, is *internal*
rural-to-urban migration, which is unobserved here. The surviving migration
association thus captures only the smaller international slice; the principal
migration driver of urban land use lies outside the design entirely, a further
reason we do not present this as identifying the main drivers of sprawl. This scope
limitation is also, more precisely, an omitted-variable concern: if internal
migration correlates with international net migration within a country over time
and independently affects BpCR, the international coefficient is not fully
insulated from it. We cannot rule this out without ADM2-level data, but the likely
direction is informative: since internal migration plausibly lowers BpCR through the
same absorption channel as international migration, a positive within-country
correlation between the two would inflate our coefficient's magnitude rather than
reverse its sign, so the qualitative finding is more robust to this omission than its
precise size. Fifth, the panel applies present-day boundaries
retroactively, so a GDP control is unavailable wherever a country did not yet exist
as a separate reporting unit, leaving 37 country-periods, mostly 1985 successor
states, that complete-case estimation drops.
These point to natural extensions: a sub-national analysis exploiting the ADM2
detail in the data, which would also measure internal migration
directly;^[A first such case study, for Germany at the *Kreis* level, is provided as supplementary material at <https://ofurkancoban.github.io/UrbanLandUseEfficiency_DevEconResearchProj/05_paper/supplementary.pdf>.]
designs aimed at causal identification; and an assessment of BpCR as a
practical, stable complement to LCRPGR in official SDG 11.3.1 monitoring.

# Conclusion {#sec-conclusion}

Urban sprawl is real and measurable, if modest in pace: across 1985--2020, built-up
area per capita rose in 63% of country-periods (67% of countries on average), though
by a slow median of about 0.4% a year. The difficulty is not the phenomenon but the
yardstick. The official LCRPGR is unstable by construction, producing
sign-flipping outliers exactly where land-use efficiency matters most; our log-based
Built-up per Capita Rate is finite and interpretable for every country-period with an
observed urban settlement and
recovers, on identical data, an economic signal the official ratio leaves
statistically undetectable at the 5% level. That demonstration, rather than the
metric itself, is our contribution to the measurement debate.

Using this stable metric on a dynamic panel of 193 United Nations member states, we
find per-capita urban land use clearly path-dependent, though its persistence is only
bracketed once the dynamic model's failing moment conditions are respected. The
search for behavioural drivers is, tellingly, mostly negative: the negative density
coefficient widely cited as compact-city evidence is an artefact of the metric's own
arithmetic (and the mechanism proper, resting on the even more entangled built-up
density, is not cleanly testable in a per-capita design); income cannot be
distinguished from zero; and the one association that survives, net in-migration, has
a mechanical negative sign but a behaviourally meaningful magnitude, a descriptive
regularity robust across estimator, timing, and the checks of @sec-results, though
still a single reduced-form association to be weighted accordingly. The substantive
contribution is thus modest and the methodological one central: for monitoring
sustainable urbanisation, the choice of metric is not a technicality but a
precondition for drawing the right conclusions, and per-capita metrics can manufacture
the very drivers they appear to detect.

# Data and Code Availability {.unnumbered}

All data are drawn from open, official sources (GHSL, GAUL, and the UN series cited
in @sec-data), and the complete pipeline that builds the panel and produces every
result is openly available in the replication repository
(<https://github.com/ofurkancoban/UrbanLandUseEfficiency_DevEconResearchProj>).
For space, several further supporting analyses referenced in the text are reported
there rather than in the paper, and leave the conclusions unchanged: the four-way
Arellano-Bond timing decomposition (the fixed-effects clean-timing check itself is
@tbl-timingfe here); the strict-exogeneity diagnostics for density (within versus
first-difference, plain-OLS and predetermined/endogenous instrumenting) and for
migration (the Wooldridge lead test and predetermined/endogenous treatments); the
instrument-depth robustness for $\rho$; the net-migration outlier robustness
(single-outlier and crisis-country exclusion, and endpoint-year timing) and its
small-cluster wild cluster bootstrap; the income-timing check; and the panel
unit-root and incremental-Hansen tests underlying the choice between the difference
and system GMM estimators.

# References {.unnumbered}

::: {#refs}
:::
