---
title: "Urban Land-Use Efficiency"
subtitle: "A Stable Metric for SDG 11.3.1, and Why Per-Capita Designs Cannot Cleanly Identify Its Drivers: 400 German Districts x 2000-2020 Panel Data Analysis"
case-study: "Sub-national Case Study: Germany (Kreis Level)"
doc-type: "Supplementary Material"
course-info: "Development Economics · SoSe 2026"
lecturer: "Lecturer: Prof. Dr. Jürgen Bitzer"
author:
  - name: Ömer Furkan Çoban
    email: oemer.furkan.coban@uni-oldenburg.de
    affiliations:
      - name: Carl von Ossietzky University of Oldenburg
        department: Applied Economics and Data Science
        city: Oldenburg
        country: Germany
    note: |
      The author used the Claude Opus 4.8 and Gemini 3.5 Flash large language models for
      improving the writing and for coding assistance in the preparation of this supplement.
      All analysis, results, and conclusions are the author's own, and the author
      takes full responsibility for the content.
date: "June 26, 2026"
abstract: |
  This supplement replicates the paper's driver analysis for the Built-up per Capita
  Rate (BpCR) at the sub-national level, using Germany (district / *Kreis* level) as
  a case study, and brings in internal (within-country) migration, which the national
  analysis cannot observe. It asks whether the within-country income, densification,
  and path-dependence results survive at a finer spatial scale.
keywords:
  - SDG 11.3.1
  - Built-up per capita rate (BpCR)
  - Sub-national analysis
  - Germany / Kreise
  - Internal migration
journal:
  name: Development Economics seminar working paper
bibliography: references.bib
format:
  elsevier-pdf:
    keep-tex: true
    template-partials:
      - before-body.tex
    include-in-header:
      text: |
        \usepackage{makecell}
        \usepackage{float}
        \geometry{top=2.5cm,bottom=2.5cm}
        \floatplacement{figure}{H}
        \floatplacement{table}{H}
        \setlength{\floatsep}{6pt plus 2pt minus 2pt}
        \setlength{\textfloatsep}{7pt plus 2pt minus 2pt}
        \setlength{\intextsep}{6pt plus 2pt minus 2pt}
    journal:
      formatting: preprint
      model: 3p
      cite-style: authoryear
  html:
    subtitle: "Supplementary Material · Sub-national Case Study: Germany (Kreis Level), Why Per-Capita Designs Cannot Cleanly Identify Its Drivers : 400 German Districts x 2000-2020 Panel Data Analysis"
    theme: [litera, theme.scss]
    embed-resources: true
    include-in-header:
      text: |
        <style>
        body { font-family: "Source Serif 4","Iowan Old Style",Palatino,Georgia,"Times New Roman",serif; line-height: 1.62; }
        h1,h2,h3,h4,h5,.title,.subtitle { font-family: "Source Sans 3","Source Sans Pro",-apple-system,"Segoe UI",Roboto,Helvetica,Arial,sans-serif; }
        p { text-align: justify; }
        nav#TOC > ul, nav[role="doc-toc"] > ul { display: block !important; }
        table tfoot tr { display: block !important; width: 100%; }
        table tfoot tr td { display: none !important; }
        table tfoot tr td:first-child { display: block !important; border: 0 !important; padding: .1rem 0 !important; font-size: .8rem; color: #4b5563; text-align: left !important; }
        </style>
    toc: true
    toc-depth: 2
    toc-title: Contents
    number-sections: true
    tbl-cap-location: top
    fig-cap-location: bottom
    smooth-scroll: true
    anchor-sections: true
    link-external-newwindow: true
    citations-hover: true
    format-links: false
    other-links:
      - text: Supplementary (PDF)
        href: supplementary.pdf
        icon: file-earmark-pdf
      - text: Main paper (HTML)
        href: paper.html
        icon: file-earmark-text
      - text: Main paper (PDF)
        href: paper.pdf
        icon: file-earmark-pdf
      - text: Presentation (slides)
        href: ../04_presentation/presentation.html
        icon: easel
      - text: Presentation (PDF)
        href: ../04_presentation/presentation.pdf
        icon: file-earmark-pdf
      - text: Replication Package
        href: https://github.com/ofurkancoban/UrbanLandUseEfficiency_DevEconResearchProj
        icon: file-earmark-zip
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
#| include: false
library(here); library(dplyr); library(readr); library(fixest)
# d_all keeps the 1990 base row (controls NA) that supplies the lagged DV for 1995
# and the deeper-lag history for the GMM; d is the 1995-2020 estimation window.
d_all <- read_csv(here("03_datasets/processed/deu_kreis_casestudy.csv"), show_col_types = FALSE) |>
    # INKAR reports the Binnenwanderungssaldo per 1,000 inhabitants; rescale to
    # percent of population so the coefficient scale matches the national control
    dplyr::mutate(int_migr = int_migr / 10) |>
    # As in the main paper: BpCR contains urban population negatively at t and
    # positively at t-1, so density and urban share (both built from that count)
    # inherit the sign of the date they are measured at. Main models use t-1;
    # t-2 shares no term with BpCR and is the mechanically clean check.
    dplyr::arrange(key, year) |>
    dplyr::group_by(key) |>
    dplyr::mutate(
        ln_density_L = dplyr::lag(ln_density),
        urban_pop_share_L = dplyr::lag(urban_pop_share),
        ln_density_L2 = dplyr::lag(ln_density, 2),
        urban_pop_share_L2 = dplyr::lag(urban_pop_share, 2)
    ) |>
    dplyr::ungroup()
d <- dplyr::filter(d_all, year >= 1995)
# format-aware table output: LaTeX for the PDF, HTML for the web build
out_fmt <- if (knitr::is_latex_output()) "latex" else "html"
```

# Data and Method

The dependent variable is built exactly as in the main paper, but for German
districts (*Kreise*) rather than countries. Urban built-up area $V$ and urban
population $P$ come from the GHSL grids [@pesaresi2024ghsl; @schiavina2023smod]
(urban-masked at GHS-SMOD $\geq 21$),
aggregated to the GAUL level-2 units for Germany at the five-year epochs
1985--2020; BpCR and urban density are then computed from
these totals just as nationally. Three series are taken from the German regional
database INKAR [@bbsr2024inkar], published by the Federal Institute for Research on
Building, Urban Affairs and Spatial Development (BBSR) and accessed through the
author's `inkaR` R package [@coban2026inkar]: GDP per capita
(workplace-based), the **internal
net-migration rate** (*Binnenwanderungssaldo*, reported per 1,000 inhabitants and
rescaled here to a percentage of population for comparability with the national
control), the latter being the within-country flow that the national UN measure
omits, and total district population, which serves as the denominator of the
urban population share (GHSL urban population over INKAR total district
population). GAUL
districts are matched to INKAR *Kreise* on a cleaned base name, disambiguating
same-named *kreisfreie Stadt* / *Landkreis* twins by geographic area, so that
`r length(unique(d$key))` of Germany's 400 districts enter the regressions. The
single exception is Freyung-Grafenau, a rural Bavarian-Forest district whose largest
settlement reaches only the rural-cluster class (GHS-SMOD 13): no cell attains the
urban threshold (SMOD $\geq 21$), so its urban built-up area and population are zero
and BpCR is undefined. This is the same urban definition used throughout the paper,
under which a unit with no urban cells is simply not an urban observation, not a
matching failure. The panel spans the epochs `r min(d$year)`--`r max(d$year)` for
which the INKAR controls are available (`r nrow(d)` district-periods). All specifications carry district and
period fixed effects with standard errors clustered by district; the dynamic
models add the lagged dependent variable and use Arellano-Bond and Blundell-Bond
GMM, as in the main paper.

@fig-layers-deu shows the same SDG 11.3.1 construction at the German district scale,
for Berlin and the surrounding Brandenburg countryside. As in the national figure,
built-up surface (panel b) is retained only where the GHS-SMOD grid classifies a cell
as urban (SMOD at least 21, panel c), giving the urban-masked built-up that enters the
analysis as $V$ (panel d); urban population $P$ is masked the same way. The cyan line
is the Berlin state boundary: the dense urban core falls inside it, while the rural
matrix that the mask removes makes up most of the surrounding districts.

![Building the analysis input following the SDG 11.3.1 metadata at the German *Kreis*
scale (Berlin / Brandenburg window), each layer drawn over an Esri satellite basemap.
(a) Basemap with the Berlin state boundary (cyan). (b) GHS-BUILT-S built-up surface,
shown as the built-up share of each 1 km cell. (c) GHS-SMOD Degree of Urbanisation
(1 km); cells coded 21 or higher count as urban. (d) Urban-masked built-up, the input
actually used: panel (b) kept only where panel (c) is urban.
](../04_outputs/figures/metadata_layers_deu.png){#fig-layers-deu width=88%}

# Drivers of Per-Capita Land Use

We estimate the drivers of BpCR at the district level exactly as in the main paper:
a two-way fixed-effects model built up control-by-control, then Arellano-Bond
[@arellano1991] (difference) and Blundell-Bond [@blundell1998] (system) GMM that
add the lagged dependent variable to capture path-dependence and correct its
Nickell bias. The key addition relative
to the national models is **internal (within-country) net migration**, the INKAR
*Binnenwanderungssaldo*, which the national panel (international migration only)
cannot observe.

```{r}
#| label: tbl-gmm
#| results: asis
#| tbl-cap: "Dynamic-panel model at the German Kreis level: FE stepwise build-up, Arellano-Bond (difference) and Blundell-Bond (system) GMM."
suppressMessages({library(fixest); library(plm); library(dplyr)})
stars <- function(p) ifelse(is.na(p),"",ifelse(p<.01,"***",ifelse(p<.05,"**",ifelse(p<.1,"*",""))))
fc    <- function(est,se,p) ifelse(is.na(est),"",sprintf("%+.4f%s<br>(%.4f)",est,stars(p),se))
fmt_w <- function(x) ifelse(is.na(x),"–",sprintf("%.3f",x))
fmt_p <- function(x) ifelse(is.na(x),"–",sprintf("%.2f",x))
fe_terms <- c("bpcr_L","ln_density_L","int_migr","ln_gdp","urban_pop_share_L")
term_lab <- c("BpCR[t−1]","ln(Urban Density)[t−1]","Internal Net Migration (% Pop)","ln(GDP per capita)","Urban Pop. Share[t−1]")
gmm_keep <- c("lag(bpcr, 1)","ln_density_L","int_migr","ln_gdp","urban_pop_share_L")
ctl <- "ln_density_L + int_migr + ln_gdp + urban_pop_share_L"
d2 <- filter(d, !is.na(bpcr_L), !is.na(ln_density_L), !is.na(urban_pop_share_L))  # lagged controls exist from 2000
fe_col <- function(m){ ct<-coeftable(m)
  vapply(fe_terms, function(v) if(v %in% rownames(ct)) fc(ct[v,1],ct[v,2],ct[v,4]) else "", character(1)) }
ff <- function(rhs) feols(as.formula(paste("bpcr ~",rhs,"| key + factor(year)")), d2, vcov=~key)
b1<-ff("bpcr_L"); b2<-ff("bpcr_L + ln_density_L"); b3<-ff("bpcr_L + ln_density_L + int_migr")
b4<-ff("bpcr_L + ln_density_L + int_migr + ln_gdp"); b5<-ff("bpcr_L + ln_density_L + int_migr + ln_gdp + urban_pop_share_L")
# GMM panel starts at 1995: with the controls dated t-1, the 1990 base row has no
# lagged controls at all and makes the instrument matrix singular. Even from 1995
# the differenced equation retains few epochs, and the diagnostics below reject;
# the district conclusions are read from the FE columns (see Limitations).
pd <- pdata.frame(as.data.frame(filter(d_all, year >= 1995) |> arrange(key,year) |>
                                mutate(t=as.integer(factor(year)))), index=c("key","t"))
gmm_col <- function(tr){
  m <- pgmm(as.formula(paste("bpcr ~ lag(bpcr,1) +",ctl,"| lag(bpcr,2:3)")),
            data=pd, effect="twoways", model="twosteps", transformation=tr, collapse=TRUE)
  s <- summary(m, robust=TRUE); cf <- s$coefficients
  list(coef=vapply(gmm_keep, function(v) fc(cf[v,1],cf[v,2],cf[v,4]), character(1)),
       nobs=sum(lengths(m$residuals)), ninst=ncol(m$W[[1]]),
       hansen=s$sargan$p.value, ar1=s$m1$p.value, ar2=s$m2$p.value)
}
ab <- gmm_col("d"); bb <- gmm_col("ld")
body <- data.frame(Variables=term_lab, Base=fe_col(b1), `+ Density`=fe_col(b2), `+ Net Migr.`=fe_col(b3),
  `+ GDP`=fe_col(b4), `+ Urban %`=fe_col(b5), `Arellano-Bond`=ab$coef, `Blundell-Bond`=bb$coef, check.names=FALSE)
ab_obs<-ab$nobs; bb_obs<-bb$nobs-ab$nobs
fed <- function(m) c(format(nobs(m),big.mark=","), fmt_w(unname(r2(m)["wr2"])),"Yes","Yes","–","–","–","–")
diag <- data.frame(Variables=c("Observations","Within R$^2$","District FE","Period FE","Instruments","Hansen (p)","AR(1) (p)","AR(2) (p)"),
  Base=fed(b1), `+ Density`=fed(b2), `+ Net Migr.`=fed(b3), `+ GDP`=fed(b4), `+ Urban %`=fed(b5),
  `Arellano-Bond`=c(format(ab_obs,big.mark=","),"–","Yes","Yes",as.character(ab$ninst),fmt_p(ab$hansen),fmt_p(ab$ar1),fmt_p(ab$ar2)),
  `Blundell-Bond`=c(format(bb_obs,big.mark=","),"–","Yes","Yes",as.character(bb$ninst),fmt_p(bb$hansen),fmt_p(bb$ar1),fmt_p(bb$ar2)),
  check.names=FALSE)
mc <- rbind(body, diag); names(mc)[1] <- "Variables"
ncoef <- 5L
split_two <- function(row){ est<-se<-row
  for(j in 2:length(row)){ x<-row[[j]]
    if(grepl("<br>",x,fixed=TRUE)){ est[[j]]<-sub("<br>.*$","",x); se[[j]]<-sub("^.*<br>","",x) } else { est[[j]]<-x; se[[j]]<-"" } }
  est[[1]]<-row[[1]]; se[[1]]<-""; rbind(est,se) }
coef_long <- do.call(rbind, lapply(seq_len(ncoef), function(i) split_two(mc[i,])))
gof <- mc[(ncoef+1):nrow(mc),]
tab <- rbind(coef_long, gof)
# escape % for LaTeX only; HTML needs the literal character
esc <- if (knitr::is_latex_output()) function(x) gsub("%","\\%",x,fixed=TRUE) else identity
names(tab)<-esc(names(tab)); tab[[1]]<-esc(tab[[1]])
nb <- nrow(coef_long)
coef_grey <- as.vector(sapply(which(seq_len(ncoef)%%2==1), function(k) c(2*k-1,2*k)))
gof_grey  <- nb + which(seq_len(nrow(gof))%%2==1)
kbl <- knitr::kable(tab, format=out_fmt, booktabs=TRUE, escape=FALSE, linesep="",
             align=c("l", rep("c", ncol(tab)-1)), row.names=FALSE) |>
  kableExtra::kable_styling(latex_options=c("HOLD_position","scale_down")) |>
  kableExtra::row_spec(c(coef_grey, gof_grey), background="#efefef") |>
  kableExtra::row_spec(nb, hline_after=TRUE) |>
  kableExtra::column_spec(6, border_right=TRUE) |>
  kableExtra::add_header_above(c(" "=1, "Fixed Effects"=5, "GMM (Bias-Corrected)"=2)) |>
  kableExtra::add_header_above(c(" "=1, "Dependent variable: BpCR"=7))
x <- as.character(kbl)
if (knitr::is_latex_output()) {
  x <- sub("(Period FE[^\n]*\\\\\\\\)", "\\1\n\\\\midrule", x)
  cat(x)
} else {
  cat("```{=html}\n", x, "\n```\n", sep = "")
}
```

\vspace{-1.0em}

\footnotesize
*Note:* \*\*\* p<0.01, \*\* p<0.05, \* p<0.1. Standard errors in parentheses: clustered by district in the fixed-effects columns; two-step GMM with the Windmeijer finite-sample correction [@windmeijer2005] in the Arellano-Bond and Blundell-Bond columns. The lagged dependent variable is instrumented by lags 2 to 3 of BpCR, with the instrument matrix collapsed. In the GMM columns the Observations row reports post-transformation equation observations (the differenced-equation set for Arellano-Bond; the larger level-equation set for Blundell-Bond). Sample: 399 of Germany's 400 districts; Freyung-Grafenau is excluded because no cell reaches the urban threshold (SMOD of at least 21), leaving its urban built-up and population at zero and BpCR undefined.
\normalsize

Three points emerge. First, per-capita land use is **path-dependent** at the
district level too: the lagged dependent variable is positive and significant in
every fixed-effects specification ($\rho \approx `r sprintf("%.2f", coef(b5)["bpcr_L"])`$
in the full model), echoing the persistence found across countries, so that
result is not an artefact of country-level aggregation; we read persistence from
these columns rather than from the dynamic-GMM point estimate, whose own
specification tests fail below, the same standard the main paper applies to its
own Hansen-failing full-model estimate. Second, the **income** coefficient is
*positive* and significant under fixed effects
(`r sprintf("%+.4f", coef(b4)["ln_gdp"])` in the +GDP column), the opposite sign
from the persistently negative, though insignificant, national estimate; it turns
insignificant in Arellano-Bond and negative and significant in Blundell-Bond, so
no stable income effect emerges from this shorter panel either, but it is not the
same null found nationally. Third, **internal net migration**, the flow the
national panel cannot observe, enters with an essentially zero coefficient: six
five-year epochs give little power to identify a within-district migration
channel, and unlike the national international-migration result it is not robust
across specifications.

*The GMM does not survive the correct timing.* With the controls dated $t-1$, the
district Arellano-Bond estimator fails both of its specification tests (Hansen and
AR(2) reject), where the same estimator with contemporaneous controls passes them.
We report the columns for completeness but draw no conclusion from them: at this
scale, with the controls correctly timed, the differenced equation retains too few
epochs for the deep-lag moment conditions to hold. The fixed-effects columns, which
need no instrument validity, carry the district evidence.

# Timing and the Compact-City Artefact {#sec-timing-deu}

The main paper's central methodological claim is that urban density and urban
population share are arithmetically entangled with BpCR, which contains urban
population negatively at $t$ and positively at $t-1$, so each control inherits the
sign of the date at which it is measured. The German districts provide an
independent test of that claim on different data, at a different scale, with a
different migration control. @tbl-timing-deu re-estimates the fixed-effects model
under the three timings on one common sample.

```{r}
#| label: tbl-timing-deu
#| results: asis
#| tbl-cap: "Timing of the two population-sharing controls at the German Kreis level: two-way fixed effects on the common sample where t, t-1 and t-2 are all observed. Only the t-2 column shares no term with the dependent variable."
cs <- dplyr::filter(d_all, !is.na(bpcr_L), !is.na(ln_gdp), !is.na(int_migr),
                    !is.na(ln_density), !is.na(urban_pop_share),
                    !is.na(ln_density_L), !is.na(urban_pop_share_L),
                    !is.na(ln_density_L2), !is.na(urban_pop_share_L2))
one <- function(dv, sv) {
  m <- feols(as.formula(paste("bpcr ~ bpcr_L +", dv, "+ int_migr + ln_gdp +", sv,
                              "| key + factor(year)")), cs, vcov = ~key)
  ct <- coeftable(m)
  c(fc(ct["bpcr_L",1], ct["bpcr_L",2], ct["bpcr_L",4]),
    fc(ct[dv,1], ct[dv,2], ct[dv,4]),
    fc(ct["int_migr",1], ct["int_migr",2], ct["int_migr",4]),
    fc(ct["ln_gdp",1], ct["ln_gdp",2], ct["ln_gdp",4]),
    fc(ct[sv,1], ct[sv,2], ct[sv,4]),
    format(nobs(m), big.mark = ","))
}
tlab <- c("BpCR[t−1]","ln(Urban Density)","Internal Net Migration (% Pop)",
          "ln(GDP per capita)","Urban Pop. Share","Observations")
tt <- data.frame(Term = tlab,
  `Both t` = one("ln_density","urban_pop_share"),
  `Both t-1` = one("ln_density_L","urban_pop_share_L"),
  `Both t-2 (clean)` = one("ln_density_L2","urban_pop_share_L2"),
  check.names = FALSE, stringsAsFactors = FALSE)
ncf <- 5L
tl <- do.call(rbind, lapply(seq_len(ncf), function(i) split_two(tt[i,])))
ta <- rbind(tl, tt[(ncf+1):nrow(tt),])
names(ta) <- esc(names(ta)); ta[[1]] <- esc(ta[[1]])
nbf <- nrow(tl)
cg <- as.vector(sapply(which(seq_len(ncf)%%2==1), function(k) c(2*k-1,2*k)))
kk <- knitr::kable(ta, format=out_fmt, booktabs=TRUE, escape=FALSE, linesep="",
                   align=c("l", rep("c", ncol(ta)-1)), row.names=FALSE) |>
  kableExtra::kable_styling(latex_options="HOLD_position", font_size=9) |>
  kableExtra::row_spec(c(cg, nbf+1), background="#efefef") |>
  kableExtra::row_spec(nbf, hline_after=TRUE) |>
  kableExtra::add_header_above(c(" "=1, "Timing of density / urban share"=3)) |>
  kableExtra::add_header_above(c(" "=1, "Dependent variable: BpCR (fixed effects)"=3))
if (knitr::is_latex_output()) cat(as.character(kk)) else cat("```{=html}\n", as.character(kk), "\n```\n", sep="")
```

Urban density reproduces the national pattern closely. It is
**$-0.029$** ($p<0.01$) when measured contemporaneously, **$+0.044$** at
$t-1$, and **$+0.031$** ($p<0.01$) at the mechanically clean $t-2$ timing, the
same sign sequence as the national $-0.024$, $+0.009$, and $+0.011$ ($p<0.05$).
Urban share shows the same sign flip at $t$ and $t-1$ ($-0.044$, then $+0.023$)
but diverges at the clean timing: **$+0.043$** ($p<0.01$) here against an
insignificant $-0.012$ nationally, so unlike density its clean-timing reading
does not replicate. The persistence term itself loses significance here
($+0.057$, against a stable $0.20$--$0.25$ nationally), likely the extra sample
loss from a second lag; we do not read $\rho$ from this column. The
district-level accounting check gives the same
signature as the national one: regressing BpCR on urban population at $t$ and
$t-1$ within districts returns $-0.011$ and $+0.013$, equal and opposite. Since
these are different data, a different country, a different spatial scale, and a
different migration control, the density reproduction is strong evidence that
the compact-city sign in per-capita land-use regressions is arithmetic rather
than economic, exactly as the main
paper argues. **H2 finds no support here either.**

# Limitations

Several limitations qualify this case study and mark it as a corroborating local
illustration rather than a stand-alone causal analysis.

*Coverage and matching.* GAUL districts are matched to INKAR by name (`r length(unique(d$key))`
of Germany's 400 districts); the only exclusion, Freyung-Grafenau, has no
urban-classified cells and hence no defined BpCR. The match relies on a cleaned
base name with an area-based rule for same-named city/rural twins rather than an
official code, so it is accurate here but would need a code crosswalk to be
fully automatic in other settings.

*Short, recent panel.* The INKAR controls are available only from 1995, so BpCR
is estimable over six five-year epochs (1995--2020), shorter than the national
panel. A further epoch is lost to lagging: 1995's urban population share is
observed, but its 1990 lag is not, since the INKAR series itself starts only in
1995, so every column in @tbl-gmm runs on 2000--2020 (five epochs,
`r nrow(d2)` district-periods, not the full 2,394), and the models speak to
medium-run trajectories rather than short-run adjustment. Differencing costs
Arellano-Bond one epoch more, retaining 2005--2020 only.

*GMM diagnostics.* Once the population-sharing controls are correctly dated at
$t-1$, neither dynamic estimator survives its specification tests at this scale:
the Arellano-Bond Hansen and AR(2) tests both reject, as do Blundell-Bond's.
Arellano-Bond passes both with contemporaneous controls, an artefact per
@sec-timing-deu and so no comfort; Blundell-Bond fails regardless of timing, a
feature of the system estimator here, not a timing artefact. The reason is
mechanical: lagging the controls costs the district panel an epoch, leaving the
differenced equation too few periods for deep-lag moments to hold.
We therefore draw the district conclusions from the fixed-effects columns, which
require no instrument validity, and report the GMM columns only for completeness.
This is a limitation of the district panel's length, not a challenge to the
national GMM, which is estimated on eight epochs and passes its AR(2) test
comfortably.

*Internal migration.* The internal net-migration control (INKAR
*Binnenwanderungssaldo*) is the within-country flow the national analysis cannot
observe, and it was the main reason for building this case study. It does not
deliver: within districts its coefficient is essentially zero, and it is not
robust to the timing of the other controls (@tbl-timing-deu). This is a
substantive disappointment rather than a technicality, since the national paper's
strongest result is the *international* migration channel and the natural
conjecture is that internal flows work the same way. A plausible reading is that
any clearer migration-densification association is cross-sectional (denser
districts attract internal migrants), a pattern this within-district design does
not test formally; four to six epochs in any case give little power. Identifying
the internal channel convincingly would need a longer or finer panel, and remains
the most valuable extension of the national analysis.

*Measurement and scope.* As in the main paper, the built-up and population layers
are satellite-based estimates, the urban area is defined at a fixed Degree-of-Urbanisation
threshold (GHS-SMOD of at least 21), and present-day GAUL boundaries are applied
throughout; the income and migration controls are administrative aggregates. The
estimates are associations, not causal effects, and they describe a single country,
so they illustrate rather than establish external validity.

# Conclusion

This supplement takes the paper's driver analysis to the sub-national (German
district) scale, across `r nrow(d)` district-periods, and delivers one clear
replication, one reversal, and one disappointment.

The replication is the paper's central methodological finding, and it is the most
valuable thing here. Urban density is significantly *negative* when measured at
the same date as the dependent variable and significantly *positive* at every
timing that is not, including the arithmetically clean $t-2$ specification; the
district accounting check reproduces the national one almost exactly
(@sec-timing-deu). Different data, different country, different scale, different
migration measure, same signature. That the compact-city coefficient in
per-capita land-use regressions is an artefact of the metric's own arithmetic is
therefore not a quirk of the global panel.

The reversal is income: positive and significant under fixed effects, unlike
the persistently negative, insignificant national estimate, though it flips
again under Blundell-Bond and drops to insignificance under Arellano-Bond, so
no single stable effect emerges, but it is not simply the national null
repeating. Path-dependence replicates
($\rho \approx `r sprintf("%.2f", coef(b5)["bpcr_L"])`$ under fixed effects, in
the range of the national estimates), though only there at this panel length,
since the dynamic-GMM estimator fails its own specification tests. The
disappointment is internal migration: the flow this
case study was built to observe is essentially zero within districts and not
robust across specifications, so the national international-migration result finds
no sub-national echo. Whether that reflects a different channel or simply
insufficient power is a question a longer district panel would answer.

# References {.unnumbered}

::: {#refs}
:::
