---
title: "Supplementary Material"
subtitle: "Sub-national Case Study: Germany (Kreis level)"
author: "Ömer Furkan Çoban"
date: today
format:
  pdf:
    number-sections: true
    geometry: [top=2.5cm, bottom=2.5cm, left=2.5cm, right=2.5cm]
    include-in-header:
      text: |
        \usepackage{makecell}
        \usepackage{booktabs}
        \usepackage{float}
        \floatplacement{table}{H}
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
#| include: false
library(here); library(dplyr); library(readr); library(fixest)
d <- read_csv(here("03_datasets/processed/deu_kreis_casestudy.csv"), show_col_types = FALSE)
sd_ratio <- round(sd(d$lcrpgr_log, na.rm = TRUE) / sd(d$bpcr, na.rm = TRUE))
star <- function(p) ifelse(p<.01,"***",ifelse(p<.05,"**",ifelse(p<.1,"*","")))
```

This supplement replicates the paper's metric comparison and driver analysis at the
**sub-national** level, using Germany as a case study. It serves two purposes: to
show that the instability of the official LCRPGR indicator, and the usability of
the Built-up per Capita Rate (BpCR), persist at a finer spatial scale; and to bring
in **internal (within-country) migration**, which the national analysis cannot
observe.

# Data and Method

The dependent variable is built exactly as in the main paper, but for German
districts (*Kreise*) rather than countries. Urban built-up area $V$ and urban
population $P$ come from the GHSL grids (urban-masked at GHS-SMOD $\geq 21$),
aggregated to the GAUL level-2 units for Germany at the five-year epochs
1985--2020; BpCR, the log-LCRPGR ratio, and urban density are then computed from
these totals just as nationally. Two controls are taken from the German regional
database INKAR (Federal Institute for Research on Building, Urban Affairs and
Spatial Development, BBSR): GDP per capita (workplace-based) and the **internal
net-migration rate** (*Binnenwanderungssaldo*, per 1,000 inhabitants), the
latter being the within-country flow that the national UN measure omits. Urban population share is the GHSL urban population over the INKAR total district
population. GAUL
districts are matched to INKAR *Kreise* by name and type (city vs. rural district);
`r length(unique(d$key))` districts match cleanly and enter the regressions, over
the epochs `r min(d$year)`--`r max(d$year)` for which the INKAR controls are
available (`r nrow(d)` district-periods). All specifications carry district and
period fixed effects with standard errors clustered by district; the dynamic
models add the lagged dependent variable and use Arellano-Bond and Blundell-Bond
GMM, as in the main paper.

# The Metric Breaks Down Locally Too

Even at the district level the contrast is stark. Across the `r nrow(d)`
district-periods the log-LCRPGR ratio has a standard deviation roughly
**`r format(sd_ratio, big.mark=",")` times** larger than that of BpCR: at this fine
scale many districts have near-zero population growth, so the ratio diverges even
more violently than nationally, whereas BpCR stays in a tight, interpretable band.

```{r tbl-fe}
#| label: tbl-fe
#| tbl-cap: "Static two-way fixed-effects models at the German Kreis level (1995-2020); log-LCRPGR ratio vs. BpCR, same sample and controls."
mL <- feols(lcrpgr_log ~ ln_density + urban_pop_share + ln_gdp + int_migr | key + year, d, vcov = ~key)
mB <- feols(bpcr       ~ ln_density + urban_pop_share + ln_gdp + int_migr | key + year, d, vcov = ~key)
est <- function(m,v){ ct<-coeftable(m); sprintf("%+.4f%s", ct[v,1], star(ct[v,4])) }
sev <- function(m,v){ ct<-coeftable(m); sprintf("(%.4f)", ct[v,2]) }
vars <- c(ln_density="ln(Urban Density)", urban_pop_share="Urban Pop. Share",
          ln_gdp="ln(GDP per capita)", int_migr="Internal net migration (per 1{,}000)")
body <- do.call(rbind, lapply(names(vars), function(v) data.frame(
  Variables=c(vars[[v]], ""),
  `LCRPGR (log)`=c(est(mL,v), sev(mL,v)),
  `BpCR`        =c(est(mB,v), sev(mB,v)), check.names=FALSE, row.names=NULL)))
gof <- data.frame(Variables=c("Observations","Within R$^2$","District FE","Period FE"),
  `LCRPGR (log)`=c(format(nobs(mL),big.mark=","), sprintf("%.4f",r2(mL)["wr2"]),"Yes","Yes"),
  `BpCR`        =c(format(nobs(mB),big.mark=","), sprintf("%.4f",r2(mB)["wr2"]),"Yes","Yes"),
  check.names=FALSE)
knitr::kable(rbind(body,gof), format="latex", booktabs=TRUE, escape=FALSE, linesep="",
             align=c("l","c","c")) |>
  kableExtra::kable_styling(latex_options="HOLD_position", font_size=9) |>
  kableExtra::row_spec(nrow(body), hline_after=TRUE) |>
  kableExtra::footnote(general=c("*** p<0.01, ** p<0.05, * p<0.1.",
                                 "Standard errors clustered by district, in parentheses."),
                       general_title="", footnote_as_chunk=FALSE, escape=FALSE, threeparttable=TRUE)
```

As nationally, BpCR yields interpretable, significant coefficients while LCRPGR does
not. Income enters BpCR negative and significant: within German districts, getting
richer is associated with *less* built-up land per resident, the same densification
result found across countries. The LCRPGR column returns enormous, mostly
insignificant coefficients dominated by its unstable tails.

# Dynamic Models and Internal Migration

```{r tbl-gmm}
#| label: tbl-gmm
#| tbl-cap: "Dynamic-panel GMM at the German Kreis level (dependent variable: BpCR); two-step with Windmeijer-corrected errors, lagged BpCR instrumented by lags 2-3, collapsed."
suppressMessages(library(plm))
pd <- pdata.frame(as.data.frame(d |> mutate(t = as.integer(factor(year)))), index = c("key","t"))
gmm_col <- function(tr){
  m <- pgmm(bpcr ~ lag(bpcr,1) + ln_density + urban_pop_share + ln_gdp + int_migr | lag(bpcr,2:3),
            data=pd, effect="twoways", model="twosteps", transformation=tr, collapse=TRUE)
  s <- summary(m, robust=TRUE); cf <- s$coefficients
  v <- c("lag(bpcr, 1)","ln_density","urban_pop_share","ln_gdp","int_migr")
  c(rbind(sapply(v,function(k) sprintf("%+.4f%s", cf[k,1], star(cf[k,4]))),
          sapply(v,function(k) sprintf("(%.4f)", cf[k,2]))),
    sprintf("%.2f", s$sargan$p.value), sprintf("%.2f", s$m2$p.value))
}
ab <- gmm_col("d"); bb <- gmm_col("ld")
rn <- c("BpCR$_{t-1}$","","ln(Urban Density)","","Urban Pop. Share","","ln(GDP per capita)","","Internal net migration","",
        "Sargan (p)","AR(2) (p)")
tab <- data.frame(Variables=rn, `Arellano-Bond`=ab, `Blundell-Bond`=bb, check.names=FALSE)
knitr::kable(tab, format="latex", booktabs=TRUE, escape=FALSE, linesep="",
             align=c("l","c","c")) |>
  kableExtra::kable_styling(latex_options="HOLD_position", font_size=9) |>
  kableExtra::row_spec(10, hline_after=TRUE) |>
  kableExtra::footnote(general=c("*** p<0.01, ** p<0.05, * p<0.1.",
                                 "Windmeijer-corrected robust standard errors in parentheses."),
                       general_title="", footnote_as_chunk=FALSE, escape=FALSE, threeparttable=TRUE)
```

Three points emerge. First, per-capita land use is **strongly path-dependent** at
the district level too: the lagged dependent variable is large and significant
($\rho \approx 0.82$), echoing the persistence found across countries. Second, the
**income-densification** result holds: the difference-GMM (Arellano-Bond) estimate
keeps the negative, significant sign of the static model, in line with the national
finding. Third, **internal net migration** is not significant within districts in
this short panel; its clearer association with densification is cross-sectional,
with the districts that attract internal migrants being the denser ones.

# Caveats

The case study matches present-day GAUL districts to INKAR by name, so a minority
of districts are dropped where the match is ambiguous; the controls limit the
window to 1995--2020; and the GMM diagnostics, acceptable for the Arellano-Bond
difference estimator, are weaker for the system estimator at this fine, short
panel. The results should therefore be read as a corroborating local illustration
rather than a stand-alone causal analysis. With those caveats, the case study
reproduces the paper's two central messages sub-nationally: LCRPGR is unusable
where population growth is near zero while BpCR recovers an interpretable
income-densification relationship; and it shows that the densifying role of
in-migration also holds for the internal flows the national analysis cannot see.
