---
title: "Urban Land-Use Efficiency"
subtitle: "A Stable Metric for SDG 11.3.1 and Its Economic and Structural Drivers"
case-study: "Sub-national Case Study: Germany (Kreis Level)"
doc-type: "Supplementary Material"
course-info: "Development Economics · SoSe 2026"
lecturer: "Lecturer: Prof. Dr. Jürgen Bitzer"
author:
  - name: Ömer Furkan Çoban
    email: oemer.furkan.coban@uni-oldenburg.de
    affiliations:
      - name: Carl von Ossietzky University of Oldenburg
        department: Applied Economics and Data Science
        city: Oldenburg
        country: Germany
    note: |
      The author used the Claude Opus 4.8 and Gemini 3.5 large language models for
      improving the writing and for coding assistance in the preparation of this supplement.
      All analysis, results, and conclusions are the author's own, and the author
      takes full responsibility for the content.
date: today
abstract: |
  This supplement replicates the paper's driver analysis for the Built-up per Capita
  Rate (BpCR) at the sub-national level, using Germany (district / *Kreis* level) as
  a case study, and brings in internal (within-country) migration, which the national
  analysis cannot observe. It asks whether the within-country income-densification
  and path-dependence results survive at a finer spatial scale.
keywords:
  - SDG 11.3.1
  - Built-up per capita rate (BpCR)
  - Sub-national analysis
  - Germany / Kreise
  - Internal migration
journal:
  name: Development Economics seminar working paper
bibliography: references.bib
format:
  elsevier-pdf:
    keep-tex: true
    template-partials:
      - before-body.tex
    include-in-header:
      text: |
        \usepackage{makecell}
        \usepackage{float}
        \geometry{top=3cm,bottom=3cm}
        \floatplacement{figure}{H}
        \floatplacement{table}{H}
        \setlength{\floatsep}{6pt plus 2pt minus 2pt}
        \setlength{\textfloatsep}{7pt plus 2pt minus 2pt}
        \setlength{\intextsep}{6pt plus 2pt minus 2pt}
    journal:
      formatting: preprint
      model: 3p
      cite-style: authoryear
  html:
    subtitle: "Supplementary Material · Sub-national Case Study: Germany (Kreis Level)"
    theme: [litera, theme.scss]
    embed-resources: true
    include-in-header:
      text: |
        <style>
        body { font-family: "Source Serif 4","Iowan Old Style",Palatino,Georgia,"Times New Roman",serif; line-height: 1.62; }
        h1,h2,h3,h4,h5,.title,.subtitle { font-family: "Source Sans 3","Source Sans Pro",-apple-system,"Segoe UI",Roboto,Helvetica,Arial,sans-serif; }
        p { text-align: justify; }
        nav#TOC > ul, nav[role="doc-toc"] > ul { display: block !important; }
        table tfoot tr { display: block !important; width: 100%; }
        table tfoot tr td { display: none !important; }
        table tfoot tr td:first-child { display: block !important; border: 0 !important; padding: .1rem 0 !important; font-size: .8rem; color: #4b5563; text-align: left !important; }
        </style>
    toc: true
    toc-depth: 2
    toc-title: Contents
    number-sections: true
    tbl-cap-location: top
    fig-cap-location: bottom
    smooth-scroll: true
    anchor-sections: true
    link-external-newwindow: true
    citations-hover: true
    format-links: false
    other-links:
      - text: Supplementary (PDF)
        href: supplementary.pdf
        icon: file-earmark-pdf
      - text: Main paper (HTML)
        href: paper.html
        icon: file-earmark-text
      - text: Presentation (slides)
        href: ../04_presentation/presentation.html
        icon: easel
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
#| include: false
library(here); library(dplyr); library(readr); library(fixest)
# d_all keeps the 1990 base row (controls NA) that supplies the lagged DV for 1995
# and the deeper-lag history for the GMM; d is the 1995-2020 estimation window.
d_all <- read_csv(here("03_datasets/processed/deu_kreis_casestudy.csv"), show_col_types = FALSE)
d <- dplyr::filter(d_all, year >= 1995)
# format-aware table output: LaTeX for the PDF, HTML for the web build
out_fmt <- if (knitr::is_latex_output()) "latex" else "html"
```

# Data and Method

The dependent variable is built exactly as in the main paper, but for German
districts (*Kreise*) rather than countries. Urban built-up area $V$ and urban
population $P$ come from the GHSL grids [@pesaresi2024ghsl; @schiavina2023smod]
(urban-masked at GHS-SMOD $\geq 21$),
aggregated to the GAUL level-2 units for Germany at the five-year epochs
1985--2020; BpCR and urban density are then computed from
these totals just as nationally. Two controls are taken from the German regional
database INKAR [@bbsr2024inkar], published by the Federal Institute for Research on
Building, Urban Affairs and Spatial Development (BBSR): GDP per capita
(workplace-based) and the **internal
net-migration rate** (*Binnenwanderungssaldo*, per 1,000 inhabitants), the
latter being the within-country flow that the national UN measure omits. Urban population share is the GHSL urban population over the INKAR total district
population. GAUL
districts are matched to INKAR *Kreise* on a cleaned base name, disambiguating
same-named *kreisfreie Stadt* / *Landkreis* twins by geographic area, so that
`r length(unique(d$key))` of Germany's 400 districts enter the regressions. The
single exception is Freyung-Grafenau, a rural Bavarian-Forest district whose largest
settlement reaches only the rural-cluster class (GHS-SMOD 13): no cell attains the
urban threshold (SMOD $\geq 21$), so its urban built-up area and population are zero
and BpCR is undefined. This is the same urban definition used throughout the paper,
under which a unit with no urban cells is simply not an urban observation, not a
matching failure. The panel spans the epochs `r min(d$year)`--`r max(d$year)` for
which the INKAR controls are available (`r nrow(d)` district-periods). All specifications carry district and
period fixed effects with standard errors clustered by district; the dynamic
models add the lagged dependent variable and use Arellano-Bond and Blundell-Bond
GMM, as in the main paper.

# Drivers of Per-Capita Land Use

We estimate the drivers of BpCR at the district level exactly as in the main paper:
a two-way fixed-effects model built up control-by-control, then Arellano-Bond
(difference) and Blundell-Bond (system) GMM that add the lagged dependent variable
to capture path-dependence and correct its Nickell bias. The key addition relative
to the national models is **internal (within-country) net migration**, the INKAR
*Binnenwanderungssaldo*, which the national panel (international migration only)
cannot observe.

```{r}
#| label: tbl-gmm
#| results: asis
#| tbl-cap: "Dynamic-panel model at the German Kreis level: FE stepwise build-up, Arellano-Bond (difference) and Blundell-Bond (system) GMM."
suppressMessages({library(fixest); library(plm); library(dplyr)})
stars <- function(p) ifelse(is.na(p),"",ifelse(p<.01,"***",ifelse(p<.05,"**",ifelse(p<.1,"*",""))))
fc    <- function(est,se,p) ifelse(is.na(est),"",sprintf("%+.4f%s<br>(%.4f)",est,stars(p),se))
fmt_w <- function(x) ifelse(is.na(x),"–",sprintf("%.3f",x))
fmt_p <- function(x) ifelse(is.na(x),"–",sprintf("%.2f",x))
fe_terms <- c("bpcr_L","ln_density","int_migr","ln_gdp","urban_pop_share")
term_lab <- c("BpCR[t−1]","ln(Urban Density)","Internal Net Migration","ln(GDP per capita)","Urban Pop. Share")
gmm_keep <- c("lag(bpcr, 1)","ln_density","int_migr","ln_gdp","urban_pop_share")
ctl <- "ln_density + int_migr + ln_gdp + urban_pop_share"
d2 <- filter(d, !is.na(bpcr_L))   # bpcr_L comes from the full series, so 1995 is kept (lag = BpCR 1990)
fe_col <- function(m){ ct<-coeftable(m)
  vapply(fe_terms, function(v) if(v %in% rownames(ct)) fc(ct[v,1],ct[v,2],ct[v,4]) else "", character(1)) }
ff <- function(rhs) feols(as.formula(paste("bpcr ~",rhs,"| key + factor(year)")), d2, vcov=~key)
b1<-ff("bpcr_L"); b2<-ff("bpcr_L + ln_density"); b3<-ff("bpcr_L + ln_density + int_migr")
b4<-ff("bpcr_L + ln_density + int_migr + ln_gdp"); b5<-ff("bpcr_L + ln_density + int_migr + ln_gdp + urban_pop_share")
# GMM panel includes the 1990 base row (NA controls) so the lag and deeper-lag instruments reach back
pd <- pdata.frame(as.data.frame(d_all |> arrange(key,year) |> mutate(t=as.integer(factor(year)))), index=c("key","t"))
gmm_col <- function(tr){
  m <- pgmm(as.formula(paste("bpcr ~ lag(bpcr,1) +",ctl,"| lag(bpcr,2:3)")),
            data=pd, effect="twoways", model="twosteps", transformation=tr, collapse=TRUE)
  s <- summary(m, robust=TRUE); cf <- s$coefficients
  if (tr=="d") rho_ab <<- cf["lag(bpcr, 1)", 1]
  list(coef=vapply(gmm_keep, function(v) fc(cf[v,1],cf[v,2],cf[v,4]), character(1)),
       nobs=sum(lengths(m$residuals)), ninst=ncol(m$W[[1]]),
       hansen=s$sargan$p.value, ar1=s$m1$p.value, ar2=s$m2$p.value)
}
ab <- gmm_col("d"); bb <- gmm_col("ld")
body <- data.frame(Variables=term_lab, Base=fe_col(b1), `+ Density`=fe_col(b2), `+ Net Migr.`=fe_col(b3),
  `+ GDP`=fe_col(b4), `+ Urban %`=fe_col(b5), `Arellano-Bond`=ab$coef, `Blundell-Bond`=bb$coef, check.names=FALSE)
ab_obs<-ab$nobs; bb_obs<-bb$nobs-ab$nobs
fed <- function(m) c(format(nobs(m),big.mark=","), fmt_w(unname(r2(m)["wr2"])),"Yes","Yes","–","–","–","–")
diag <- data.frame(Variables=c("Observations","Within R$^2$","District FE","Period FE","Instruments","Hansen (p)","AR(1) (p)","AR(2) (p)"),
  Base=fed(b1), `+ Density`=fed(b2), `+ Net Migr.`=fed(b3), `+ GDP`=fed(b4), `+ Urban %`=fed(b5),
  `Arellano-Bond`=c(format(ab_obs,big.mark=","),"–","Yes","Yes",as.character(ab$ninst),fmt_p(ab$hansen),fmt_p(ab$ar1),fmt_p(ab$ar2)),
  `Blundell-Bond`=c(format(bb_obs,big.mark=","),"–","Yes","Yes",as.character(bb$ninst),fmt_p(bb$hansen),fmt_p(bb$ar1),fmt_p(bb$ar2)),
  check.names=FALSE)
mc <- rbind(body, diag); names(mc)[1] <- "Variables"
ncoef <- 5L
split_two <- function(row){ est<-se<-row
  for(j in 2:length(row)){ x<-row[[j]]
    if(grepl("<br>",x,fixed=TRUE)){ est[[j]]<-sub("<br>.*$","",x); se[[j]]<-sub("^.*<br>","",x) } else { est[[j]]<-x; se[[j]]<-"" } }
  est[[1]]<-row[[1]]; se[[1]]<-""; rbind(est,se) }
coef_long <- do.call(rbind, lapply(seq_len(ncoef), function(i) split_two(mc[i,])))
gof <- mc[(ncoef+1):nrow(mc),]
tab <- rbind(coef_long, gof)
esc <- function(x) gsub("%","\\%",x,fixed=TRUE); names(tab)<-esc(names(tab)); tab[[1]]<-esc(tab[[1]])
nb <- nrow(coef_long)
coef_grey <- as.vector(sapply(which(seq_len(ncoef)%%2==1), function(k) c(2*k-1,2*k)))
gof_grey  <- nb + which(seq_len(nrow(gof))%%2==1)
kbl <- knitr::kable(tab, format=out_fmt, booktabs=TRUE, escape=FALSE, linesep="",
             align=c("l", rep("c", ncol(tab)-1)), row.names=FALSE) |>
  kableExtra::kable_styling(latex_options=c("HOLD_position","scale_down")) |>
  kableExtra::row_spec(c(coef_grey, gof_grey), background="#efefef") |>
  kableExtra::row_spec(nb, hline_after=TRUE) |>
  kableExtra::column_spec(6, border_right=TRUE) |>
  kableExtra::add_header_above(c(" "=1, "Fixed Effects"=5, "GMM (Bias-Corrected)"=2)) |>
  kableExtra::add_header_above(c(" "=1, "Dependent variable: BpCR"=7))
x <- as.character(kbl)
if (knitr::is_latex_output()) {
  x <- sub("(Period FE[^\n]*\\\\\\\\)", "\\1\n\\\\midrule", x)
  cat(x)
} else {
  cat("```{=html}\n", x, "\n```\n", sep = "")
}
```

\vspace{-1.0em}

\footnotesize
*Note:* \*\*\* p<0.01, \*\* p<0.05, \* p<0.1. Standard errors in parentheses: clustered by district in the fixed-effects columns; two-step GMM with the Windmeijer finite-sample correction [@windmeijer2005] in the Arellano-Bond and Blundell-Bond columns. The lagged dependent variable is instrumented by lags 2 to 3 of BpCR, with the instrument matrix collapsed. In the GMM columns the Observations row reports post-transformation equation observations (differenced for Arellano-Bond; the larger stacked level-equation set for Blundell-Bond). Sample: 399 of Germany's 400 districts; Freyung-Grafenau is excluded because no cell reaches the urban threshold (SMOD of at least 21), leaving its urban built-up and population at zero and BpCR undefined.
\normalsize

Three points emerge. First, per-capita land use is **strongly path-dependent** at
the district level too: the lagged dependent variable is large and significant
($\rho \approx `r sprintf("%.2f", rho_ab)`$), echoing the persistence found across countries. Second, the
**income-densification** result holds: ln(GDP per capita) is negative and
significant across the fixed-effects build-up and in the difference-GMM
(Arellano-Bond) estimate, in line with the national finding. Third, **internal net migration** is not significant within districts in
this short panel; its clearer association with densification is cross-sectional,
with the districts that attract internal migrants being the denser ones.

# Limitations

Several limitations qualify this case study and mark it as a corroborating local
illustration rather than a stand-alone causal analysis.

*Coverage and matching.* GAUL districts are matched to INKAR by name (`r length(unique(d$key))`
of Germany's 400 districts); the only exclusion, Freyung-Grafenau, has no
urban-classified cells and hence no defined BpCR. The match relies on a cleaned
base name with an area-based rule for same-named city/rural twins rather than an
official code, so it is accurate here but would need a code crosswalk to be
fully automatic in other settings.

*Short, recent panel.* The INKAR controls are available only from 1995, so the
estimation window is six five-year epochs (1995--2020). This is shorter than the
national panel and, together with the five-year spacing, means the models speak to
medium-run trajectories rather than short-run adjustment. The brevity matters most
for the dynamic estimators: because the GHSL series used here begins in 1985 (BpCR
from 1990), the deeper lags that instrument the Arellano-Bond difference equation
are available only from 2000 onward, so the GMM is estimated on fewer periods than
the within (fixed-effects) columns even though the latter retain all of 1995--2020.

*GMM diagnostics.* The Arellano-Bond difference estimator is well behaved: the
Hansen test does not reject the instruments, first-order serial correlation is
present and second-order is absent, as required. The Blundell-Bond system
estimator, by contrast, is rejected at this fine and short scale (its Hansen and
AR(2) tests fail), so we treat Arellano-Bond as the preferred dynamic estimator and
read Blundell-Bond only as a sign-and-significance cross-check.

*Internal migration.* The internal net-migration control (INKAR
*Binnenwanderungssaldo*) is the within-country flow the national analysis cannot
observe, but it is not statistically significant within districts in this short
panel; its association with densification appears to be cross-sectional (the
districts that attract internal migrants are the already-denser ones) rather than a
within-district dynamic that six epochs can identify.

*Measurement and scope.* As in the main paper, the built-up and population layers
are satellite-based estimates, the urban area is defined at a fixed Degree-of-Urbanisation
threshold (GHS-SMOD of at least 21), and present-day GAUL boundaries are applied
throughout; the income and migration controls are administrative aggregates. The
estimates are associations, not causal effects, and they describe a single country,
so they illustrate rather than establish external validity.

# Conclusion

This supplement reproduces the paper's substantive findings at the sub-national
(German district) scale, across `r nrow(d)` district-periods. Per-capita urban land
use is strongly path-dependent within districts (Arellano-Bond
$\rho \approx `r sprintf("%.2f", rho_ab)`$), echoing the persistence found across
countries, and it declines with income: getting richer is associated with less
built-up land per resident, the same compact-city, income-densification relationship
the national panel shows. The result is therefore not an artefact of country-level
aggregation; it holds at a much finer spatial resolution and within a single
national institutional setting.

The case study also adds the internal migration that the national analysis cannot
see. The within-country net-migration flow is not statistically significant in this
short, six-epoch panel; its sign and the broader cross-sectional pattern are
consistent with in-migrants being absorbed into the existing urban stock rather than
driving outward expansion, but identifying that channel would require a longer panel.
Taken together, the German evidence corroborates, at a scale the national panel
cannot reach, that economic growth within a place is accompanied by densification
rather than sprawl.

# References {.unnumbered}

::: {#refs}
:::
